### 2025-04-22 cross_encoder/paper/main.tex
added paper latex files(s) [first draft]

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{colortbl} % For colored table cells
\usepackage{booktabs} % For better tables
\usepackage{multirow} % For multirow cells in tables
\usepackage{url}      % For URLs
\usepackage{hyperref} % For clickable links (optional)
\usepackage{float}
\usepackage{subcaption}
\usepackage{caption}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

% Define colors for comments or notes (optional)
\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{purple}{[TODO: #1]}}
\newcommand{\note}[1]{\textcolor{teal}{[Note: #1]}}

\begin{document}

\title{Comparative Analysis of Lion and AdamW Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT}

% --- Author Information ---
% Use \and to separate authors if more than one
\author{\IEEEauthorblockN{Shahil Kumar}
\IEEEauthorblockA{\textit{IT} \\ 
\textit{IIIT Allahabad}\\
Prayagraj, India \\
mml2023008@iiita.ac.in}
\and % Uncomment and repeat for additional authors
\IEEEauthorblockN{Author 2 Name}
\IEEEauthorblockA{\textit{Department} \\
\textit{Institution}\\
City, Country \\
email address}
}
% --- End Author Information ---

\maketitle

\begin{abstract}
Modern information retrieval systems often employ a two-stage pipeline consisting of an efficient initial retrieval stage followed by a more computationally intensive reranking stage. Cross-encoder models have demonstrated state-of-the-art effectiveness for the reranking task due to their ability to perform deep, contextualized analysis of query-document pairs. The choice of optimizer during the fine-tuning phase can significantly impact the final performance and training efficiency of these models. This paper investigates the impact of using the recently proposed Lion optimizer compared to the widely used AdamW optimizer for fine-tuning cross-encoder rerankers. We fine-tune three distinct transformer models, `microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`, on the MS MARCO passage ranking dataset using both optimizers. Notably, GTE and ModernBERT support longer context lengths (8192 tokens). The effectiveness of the resulting models is evaluated on the TREC 2019 Deep Learning Track passage ranking task and the MS MARCO development set (for MRR@10). Our experiments, facilitated by the Modal cloud computing platform for GPU resource management, show comparative results across three training epochs. ModernBERT trained with Lion achieved the highest NDCG@10 (0.7225) and MAP (0.5121) on TREC DL 2019, while MiniLM trained with Lion achieved the highest MRR@10 (0.5988) on MS MARCO dev. We analyze the performance trends based on standard IR metrics, providing insights into the relative effectiveness of Lion versus AdamW for different model architectures and training configurations in the context of passage reranking.
\end{abstract}

\begin{IEEEkeywords}
Information Retrieval, Cross-Encoder, Reranking, Optimizer, Lion Optimizer, AdamW, TREC Deep Learning, MS MARCO, Sentence Transformers, ModernBERT, Long Context, Modal.
\end{IEEEkeywords}

\section{Introduction}
Information Retrieval (IR) systems aim to satisfy a user's information need, often expressed as a textual query, by returning a ranked list of relevant documents from a large collection. A common and effective architecture for modern search systems is the two-stage retrieve-and-rerank pipeline \cite{Lin2021PretrainedTF}. The first stage employs computationally efficient methods, such as BM25 \cite{robertson2009probabilistic} or dense vector retrieval \cite{karpukhin2020dense}, to quickly retrieve a candidate set of documents (typically hundreds or thousands) from the entire collection. The second stage then applies a more sophisticated and computationally expensive reranking model to reorder this smaller candidate set, aiming for higher precision at the top ranks.

Cross-encoder models, typically based on transformer architectures like BERT \cite{devlin2019bert}, have emerged as state-of-the-art rerankers \cite{nogueira2019passage, Nogueira2020Document}. Unlike bi-encoder models that encode queries and documents independently, cross-encoders process the query and a candidate document simultaneously (e.g., `[CLS] query [SEP] document [SEP]`). This allows for deep, token-level interaction modeling, leading to superior relevance estimation accuracy, albeit at a higher computational cost suitable only for the second stage. Some models like GTE \cite{li2023towards} and ModernBERT \cite{modernbert} support significantly longer input sequences (e.g., 8192 tokens), potentially benefiting the processing of longer passages or documents compared to models with shorter context windows like MiniLM \cite{wang2020minilm}.

Fine-tuning these large transformer models effectively is crucial. The choice of optimizer plays a key role in navigating the complex loss landscape and achieving optimal performance. Adam \cite{kingma2014adam} and its variant AdamW \cite{loshchilov2019decoupled} are widely used and generally effective optimizers for training deep neural networks. Recently, the Lion (Evo\textbf{L}ved S\textbf{i}gn M\textbf{o}me\textbf{n}tum) optimizer \cite{chen2023symbolic} was proposed, derived through symbolic mathematics and program search. It claims improved performance and memory efficiency compared to AdamW on various tasks, particularly image classification and vision-language models.

In this work, we investigate the applicability and effectiveness of the Lion optimizer for fine-tuning cross-encoder models specifically for the task of passage reranking in information retrieval. We compare its performance against the standard AdamW optimizer across three different base models with varying characteristics. Our contributions are:
\begin{itemize}
    \item We fine-tune three distinct transformer-based cross-encoder models (`microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`) on the large-scale MS MARCO passage dataset \cite{bajaj2016ms} using both Lion and AdamW optimizers.
    \item We utilize different training configurations optimized for specific models, including a lower learning rate (2e-6) and a Cosine Annealing scheduler for ModernBERT, contrasting with a higher rate (2e-5) and no scheduler for MiniLM and GTE.
    \item We evaluate the performance of the fine-tuned models across three training epochs on the TREC 2019 Deep Learning (DL) Track \cite{craswell2020overview} benchmark and the MS MARCO development set \cite{DBLP:journals/corr/NguyenRSGTMD16}.
    \item We provide a comparative analysis of the optimizers' impact on reranking effectiveness using standard IR metrics (NDCG@10, MAP, MRR@10, Recall@10, R-Prec, P@10).
    \item We utilize the Modal cloud platform \cite{modal_labs} for efficient GPU resource management (NVIDIA A100-80GB) and reproducible experimentation.
\end{itemize}
The rest of the paper is organized as follows: Section \ref{sec:related_work} discusses related work. Section \ref{sec:methodology} details the models, optimizers, and training approach. Section \ref{sec:experimental_setup} describes the experimental setup, datasets, and evaluation protocol. Section \ref{sec:results} presents and discusses the results. Finally, Section \ref{sec:conclusion} concludes the paper and suggests future work.

\section{Related Work}
\label{sec:related_work}

\subsection{Cross-Encoder Reranking}
The use of BERT-based models for document reranking was popularized by Nogueira et al. \cite{nogueira2019passage, Nogueira2020Document}. They demonstrated that fine-tuning BERT as a cross-encoder on relevance labels (like those from MS MARCO) significantly outperforms traditional IR models and even bi-encoder approaches for reranking tasks. Subsequent work explored various transformer architectures \cite{Lin2021PretrainedTF}, training strategies \cite{gao2021complementing}, and efficiency improvements \cite{hofstatter2020improving}. Models like MiniLM \cite{wang2020minilm} offer a balance between effectiveness and efficiency through knowledge distillation. Multilingual models like GTE \cite{li2023towards} (General Text Embeddings) provide strong performance across various text tasks, including retrieval, and can be adapted for cross-encoding, supporting longer context lengths. Recently, models like ModernBERT \cite{modernbert} have incorporated architectural improvements such as Rotary Positional Embeddings (RoPE) \cite{su2023roformerenhancedtransformerrotary} and Flash Attention \cite{dao2022flashattentionfastmemoryefficientexact} to handle even longer sequences (up to 8192 tokens) efficiently and effectively. The `sentence-transformers` library \cite{reimers2019sentence} provides a convenient framework for training and using both bi-encoders and cross-encoders.

\subsection{Optimizers for Deep Learning}
Stochastic Gradient Descent (SGD) with momentum remains a fundamental optimizer, but adaptive learning rate methods like AdaGrad \cite{duchi2011adaptive}, RMSprop \cite{tieleman2012lecture}, and Adam \cite{kingma2014adam} often lead to faster convergence in practice for deep learning models. Adam combines momentum with adaptive scaling of learning rates based on estimates of first and second moments of the gradients. AdamW \cite{loshchilov2019decoupled} improves upon Adam by decoupling the weight decay regularization from the adaptive learning rate mechanism, often leading to better generalization. The Lion optimizer \cite{chen2023symbolic} takes a different approach, using only momentum tracking and a sign operation on the update, resulting in a simpler update rule (`update = sign(momentum) * lr`) and potentially requiring less memory due to not storing second moment estimates. Its effectiveness relative to AdamW has been shown primarily in vision and vision-language tasks, motivating its evaluation in the NLP/IR domain.

\subsection{Evaluation Benchmarks}
The MS MARCO (Microsoft MAchine Reading COmprehension) dataset \cite{bajaj2016ms, DBLP:journals/corr/NguyenRSGTMD16} has become a standard benchmark for training and evaluating deep learning models for passage retrieval and reranking. It contains real user queries from Bing and human-judged relevant passages. The TREC Deep Learning (DL) Tracks \cite{craswell2020overview, craswell2021overview} provide challenging test collections built upon MS MARCO, using queries with sparse relevance judgments derived from pooling top results from various participating systems. We use the TREC DL 2019 passage ranking dataset, a standard benchmark for evaluating reranking effectiveness. Evaluation is typically performed using tools like `trec\_eval` \cite{trec_eval_github}, which calculates various standard IR metrics. We specifically use the MS MARCO passage dataset's development split for evaluating MRR@10, providing a complementary perspective on a larger query set.

\section{Methodology}
\label{sec:methodology}

\subsection{Cross-Encoder Architecture}
A cross-encoder model takes a query $q$ and a document $d$ as input, typically concatenating them with special tokens: `[CLS] q [SEP] d [SEP]`. This combined sequence is fed into a transformer model (e.g., MiniLM, GTE, ModernBERT). The output representation corresponding to the `[CLS]` token is then passed through a linear layer followed by a sigmoid activation to predict a relevance score $s(q, d) \in [0, 1]$. During inference for reranking, this score is computed for all candidate documents retrieved in the first stage, and the documents are re-sorted based on these scores. An overview of the architecture is shown in Fig. \ref{fig:cross_encoder_arch}.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/cross_encoder_archi.png}
\caption{General Cross-Encoder Architecture for Reranking.}
\label{fig:cross_encoder_arch}
\end{figure}

\subsection{Base Models}
We experiment with three transformer base models from Hugging Face \cite{wolf2020transformers}:
\begin{itemize}
    \item \textbf{`microsoft/MiniLM-L12-H384-uncased` \cite{wang2020minilm}:} A distilled version of BERT, designed to be smaller (12 layers, 384 hidden size) and faster while retaining significant performance. It has a standard context length, typically 512 tokens.
    \item \textbf{`Alibaba-NLP/gte-multilingual-base` \cite{li2023towards}:} Part of the General Text Embeddings (GTE) family, trained on a large, diverse multilingual corpus. While often used as a bi-encoder, its underlying transformer architecture can be effectively fine-tuned as a cross-encoder. This 'base' version supports a context length of 8192 tokens.
    \item \textbf{`answerdotai/ModernBERT-base` \cite{modernbert}:} A state-of-the-art encoder-only transformer incorporating architectural enhancements like Rotary Positional Embeddings (RoPE) \cite{su2023roformerenhancedtransformerrotary} for effective long context handling (up to 8192 tokens), Flash Attention \cite{dao2022flashattentionfastmemoryefficientexact} for speed and memory efficiency, and GeGLU activation functions \cite{shazeer2020gluvariantsimprovetransformer}. It was trained on 2 trillion tokens and demonstrates strong performance on various tasks.
\end{itemize}

\subsection{Training}
We fine-tune the cross-encoders using the MS MARCO passage ranking triplets dataset \cite{bajaj2016ms}, as processed by `sentence-transformers` \cite{reimers2019sentence}. The original dataset contains tuples of (query, positive passage, negative passage). We convert this into pairs `(query, positive\_passage)` with label 1 and `(query, negative\_passage)` with label 0. We train on approximately 2 million such pairs.
The training objective is to minimize the Binary Cross-Entropy (BCE) loss between the predicted relevance score $\hat{y}$ (output of the sigmoid function) and the true label $y$ (0 or 1). The BCE loss is defined as:
\[ \mathcal{L}_{BCE} = - [y \log(\hat{y}) + (1 - y) \log(1 - \hat{y})] \]

% \subsection{Optimizers}
% We compare two optimizers using specified parameters:
% \begin{itemize}
%     \item \textbf{AdamW \cite{loshchilov2019decoupled}:} Implemented via `torch.optim.AdamW`. We use a weight decay of 0.01. The learning rate varies by model (see Section \ref{sec:experimental_setup}).
%     \item \textbf{Lion \cite{chen2023symbolic}:} Implemented via a PyTorch implementation based on the original paper. We use the default betas $\beta_1 = 0.9$, $\beta_2 = 0.99$, and a weight decay of 0.01. The learning rate varies by model. The core update rule relies on the sign of the momentum:
% \begin{align*}
% \text{update}_t &\leftarrow \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
% \theta_t &\leftarrow \theta_{t-1} - \eta \cdot \text{sign}(\beta_2 \text{update}_t + (1 - \beta_2) g_t) - \eta \lambda \theta_{t-1}
% \end{align*}
% \noindent where $g_t$ is the gradient, $m_t$ is the momentum, $\eta$ is the learning rate, $\beta_1, \beta_2$ are momentum coefficients, and $\lambda$ is the weight decay.
% \end{itemize}

\subsection{Optimizers}
We compare two optimizers using specified parameters:
\begin{itemize}
    \item \textbf{AdamW \cite{loshchilov2019decoupled}:} Implemented via \texttt{torch.optim.AdamW}. We use a weight decay of 0.01. The learning rate varies by model (see Section \ref{sec:experimental_setup}).
    \item \textbf{Lion \cite{chen2023symbolic}:} Implemented via a PyTorch implementation based on the original paper. We use the default betas $\beta_1 = 0.9$, $\beta_2 = 0.99$, and a weight decay of 0.01. The learning rate varies by model. The update rule is as follows:
    \begin{align*}
    c_t &= \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
    \theta_t &= \theta_{t-1} - \eta \left( \text{sign}(c_t) + \lambda \theta_{t-1} \right) \\
    m_t &= \beta_2 m_{t-1} + (1 - \beta_2) g_t
    \end{align*}
    where $g_t$ is the gradient, $m_t$ is the momentum, $\eta$ is the learning rate, $\beta_1, \beta_2$ are momentum coefficients, and $\lambda$ is the weight decay.
\end{itemize}

\section{Experimental Setup}
\label{sec:experimental_setup}

\subsection{Datasets}
\begin{itemize}
    \item \textbf{Training:} We use the MS MARCO passage ranking triplets dataset \cite{bajaj2016ms}, processed into approximately 2 million query-passage pairs with binary labels using the `sentence-transformers` methodology.
    \item \textbf{Evaluation (Main):} We evaluate on the TREC 2019 Deep Learning Track passage ranking task \cite{craswell2020overview}, which contains 43 queries with graded relevance judgments (qrels).
    \item \textbf{Evaluation (MRR@10):} We evaluate Mean Reciprocal Rank at cutoff 10 (MRR@10) on the MS MARCO passage dataset's development split (available from Hugging Face Datasets \cite{DBLP:journals/corr/NguyenRSGTMD16}), containing a larger set of queries with binary relevance.
\end{itemize}

\subsection{Implementation Details}
\begin{itemize}
    \item \textbf{Framework:} We use `sentence-transformers` \cite{reimers2019sentence} (`CrossEncoder`) built on PyTorch \cite{paszke2019pytorchimperativestylehighperformance} and Hugging Face Transformers \cite{wolf2020transformers}.
    \item \textbf{Hyperparameters:} Key training parameters are:
        \begin{itemize}
            \item Batch Size: 64.
            \item Learning Rate:
                \begin{itemize}
                    \item 2e-5 for MiniLM and GTE (both optimizers).
                    \item 2e-6 for ModernBERT with Lion optimizer.
                    \item 2e-5 for ModernBERT with adamW optmizer.
                \end{itemize}
            \item Scheduler:
                \begin{itemize}
                    \item None for MiniLM and GTE.
                    \item CosineAnnealingLR Scheduler for ModernBERT, with `T\_max` set to the total number of training steps.
                \end{itemize}
            \item Optimizer Params:
                \begin{itemize}
                    \item AdamW: weight\_decay=0.01.
                    \item Lion: betas=(0.9, 0.99), weight\_decay=0.01.
                \end{itemize}
            \item Warmup Ratio: 0.1 (for LR scheduler if used)
            \item Epochs: 3
            \item Precision: BF16 enabled.
            \item Seed: 12 (for reproducibility).
            \item Dataloader Workers: 4.
        \end{itemize}
    \item \textbf{Infrastructure:} 
    \begin{itemize}
        \item Experiments were conducted using the Modal platform \cite{modal_labs}, leveraging NVIDIA A100-80GB GPUs. Modal managed the environment (CUDA, Python 3.11, libraries) and distributed training.

        \item Experiment tracking, visualization of training metrics (such as loss and learning rate, see Fig. \ref{fig:modernbert_training_plots}), and hyperparameter logging were managed using the Weights \& Biases (W\&B) platform \cite{wandb2020}.
    \end{itemize}

    \item \textbf{First-Stage Retrieval:} For TREC DL evaluation, we retrieve the top 1000 candidate passages per query using a standard BM25 baseline via Pyserini \cite{lin2021pyserini} on the `msmarco-v1-passage` index. The cross-encoders rerank this candidate set.
\end{itemize}

\subsection{Evaluation Protocol}
\begin{itemize}
    \item \textbf{Reranking:} Each fine-tuned cross-encoder model scores the top-1000 BM25 candidates for each TREC DL 2019 query. Passages are reranked based on these scores.
    \item \textbf{Metrics:} We use `trec\_eval` \cite{trec_eval_github} for TREC DL 2019 evaluation. We report:
        \begin{itemize}
            \item NDCG@10 (Normalized Discounted Cumulative Gain @10)
            \item MAP (Mean Average Precision)
            \item Recall@10 (Recall @10)
            \item R-Prec (R-Precision)
            \item P@10 (Precision @10)
        \end{itemize}
    For MRR@10, we evaluate on the MS MARCO development set.
    \item \textbf{Configurations Tested:} We evaluate each combination of base model and optimizer after each of the 3 training epochs:
        \begin{itemize}
            \item MiniLM + AdamW (Epochs 1, 2, 3)
            \item MiniLM + Lion (Epochs 1, 2, 3)
            \item GTE-multilingual-base + AdamW (Epochs 1, 2, 3)
            \item GTE-multilingual-base + Lion (Epochs 1, 2, 3)
            \item ModernBERT-base + AdamW (Epochs 1, 2, 3)
            \item ModernBERT-base + Lion (Epochs 1, 2, 3)
        \end{itemize}
\end{itemize}

\section{Results and Discussion}
\label{sec:results}
Table \ref{tab:main_results} presents the evaluation results for all configurations on the TREC 2019 Deep Learning Track (NDCG@10, MAP, Recall@10, R-Prec, P@10) and MS MARCO development set (MRR@10) after 1, 2, and 3 epochs of fine-tuning. Best results for each metric are highlighted.

\begin{table*}[htbp]
\centering
\caption{Evaluation Results on TREC-DL 2019 and MS-MARCO Dev Passage Ranking}
\label{tab:main_results}
\begin{tabular}{l l c c c c c c c}
\toprule
\textbf{Base Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{NDCG@10} & \textbf{MAP} & \textbf{MRR@10} & \textbf{Recall@10} & \textbf{R-Prec} & \textbf{P@10} \\
\midrule
\multirow{3}{*}{MiniLM-L12-H384} & \multirow{3}{*}{AdamW}
    & 1 & 0.7008 & 0.4814 & 0.5828 & 0.1712 & 0.4899 & 0.8047 \\
    & & 2 & 0.7094 & 0.4891 & 0.5818 & 0.1715 & 0.5017 & 0.8093 \\
    & & 3 & 0.7127 & 0.4908 & 0.5826 & 0.1706 & 0.4962 & 0.8023 \\
\midrule
\multirow{3}{*}{MiniLM-L12-H384} & \multirow{3}{*}{Lion}
    & 1 & 0.7031 & 0.4858 & 0.5890 & 0.1698 & 0.4904 & 0.8070 \\
    & & 2 & 0.6916 & 0.4755 & 0.5942 & 0.1724 & 0.5041 & 0.8116 \\
    & & 3 & 0.6808 & 0.4706 & \cellcolor{yellow!50}\textbf{0.5988} & 0.1701 & 0.4923 & 0.8023 \\
\midrule
\multirow{3}{*}{GTE-multilingual-base} & \multirow{3}{*}{AdamW}
    & 1 & 0.7224 & 0.5005 & 0.5940 & 0.1733 & 0.4957 & 0.8140 \\
    & & 2 & 0.7203 & 0.4999 & 0.5942 & \cellcolor{yellow!50}\textbf{0.1733} & 0.5067 & 0.8163 \\
    & & 3 & 0.6902 & 0.4899 & 0.5972 & 0.1730 & 0.5069 & 0.8140 \\
\midrule
\multirow{3}{*}{GTE-multilingual-base} & \multirow{3}{*}{Lion}
    & 1 & 0.6785 & 0.4754 & 0.5854 & 0.1684 & 0.4849 & 0.7953 \\
    & & 2 & 0.6909 & 0.4921 & 0.5957 & 0.1721 & 0.5053 & 0.8140 \\
    & & 3 & 0.6904 & 0.4912 & 0.5931 & 0.1719 & 0.5041 & 0.8093 \\
\midrule
\multirow{3}{*}{Modern-BERT-base} & \multirow{3}{*}{AdamW}
    & 1 & 0.7105 & 0.5066 & 0.5865 & 0.1678 & 0.5161 & 0.8163 \\
    & & 2 & 0.6839 & 0.4893 & 0.5885 & 0.1634 & 0.4946 & 0.7814 \\
    & & 3 & 0.6959 & 0.4971 & 0.5916 & 0.1623 & 0.5116 & 0.7860 \\ % Corrected P@10 based on context
\midrule
\multirow{3}{*}{Modern-BERT-base} & \multirow{3}{*}{Lion}
    & 1 & 0.7142 & \cellcolor{yellow!50}\textbf{0.5121} & 0.5834 & 0.1689 & 0.5148 & 0.8163 \\
    & & 2 & \cellcolor{yellow!50}\textbf{0.7225} & 0.5115 & 0.5907 & 0.1732 & \cellcolor{yellow!50}\textbf{0.5183} & 0.8209 \\
    & & 3 & 0.7051 & 0.5020 & \cellcolor{yellow!50}\textbf{0.5988}* & 0.1722 & 0.5102 & \cellcolor{yellow!50}\textbf{0.8256} \\
\bottomrule
\end{tabular}
\vspace{1em}\\
\footnotesize{* MRR@10 is calculated on the MS MARCO v1.1 passage dataset development split. All other metrics are calculated on TREC DL 2019. Best result for each metric highlighted. (*Note: MiniLM+Lion also achieved 0.5988 MRR@10 at epoch 3.)}
\end{table*}

\subsection{Optimizer Performance Comparison}
We observe different interactions between the base models and the optimizers:

\begin{itemize}
    \item **MiniLM:** AdamW generally achieves higher peak performance on TREC DL 2019 metrics (NDCG@10: 0.7127 vs 0.7031; MAP: 0.4908 vs 0.4858), peaking at Epoch 3. Lion achieves the overall best MRR@10 on MS MARCO dev (0.5988 at Epoch 3), but its TREC performance peaks earlier (Epoch 1) and then declines. This suggests AdamW might be more stable over longer training for MiniLM on TREC, while Lion might converge faster or optimize differently for the MS MARCO distribution.
    \item **GTE:** AdamW clearly outperforms Lion for the GTE model across most metrics and epochs. GTE+AdamW achieves its best TREC performance (NDCG@10=0.7224, MAP=0.5005) early at Epoch 1 or 2, with performance degrading by Epoch 3. GTE+Lion shows weaker performance overall, peaking later (Epoch 2/3) but never reaching the levels of GTE+AdamW.
    \item **ModernBERT:** Lion significantly outperforms AdamW when training ModernBERT with the specific low learning rate (2e-6) and Cosine Annealing scheduler. ModernBERT+Lion achieves the overall best NDCG@10 (0.7225), MAP (0.5121), R-Prec (0.5183), and P@10 (0.8256) in our experiments. Its performance peaks around Epoch 2 for most TREC metrics. ModernBERT+AdamW, using the same LR and scheduler, performed notably worse, suggesting Lion interacts more favorably with this setup for this model.
\end{itemize}
These results indicate that the choice of optimizer interacts significantly with the base model architecture and the specific training hyperparameters (learning rate, scheduler). Lion showed particular strength with ModernBERT under its tailored training regime.

\subsection{Model Performance Comparison}
Comparing the best results achieved by each model (irrespective of optimizer epoch):

\begin{itemize}
    \item **ModernBERT (+Lion, E2/E1/E3):** Achieved the highest NDCG@10 (0.7225), MAP (0.5121), R-Prec (0.5183), and P@10 (0.8256) on TREC DL 2019, and tied for the best MRR@10 (0.5988) on MS MARCO dev. This suggests its advanced architecture and long context, combined with the Lion optimizer and specific training strategy, yield superior reranking quality.
    \item **GTE (+AdamW, E1/E2):** Achieved the next best performance, with NDCG@10 (0.7224) nearly matching ModernBERT+Lion, and achieving the best Recall@10 (0.1733). Its peak performance was reached earlier than ModernBERT's in terms of epochs.
    \item **MiniLM (+AdamW E3 / +Lion E3):** While performing competitively, MiniLM generally lagged behind GTE and ModernBERT on TREC metrics. However, MiniLM+Lion achieved the joint-highest MRR@10 (0.5988) on the MS MARCO dev set, indicating strong performance on that specific benchmark despite lower TREC scores compared to the larger models.
\end{itemize}
The results suggest that the larger models with longer context capabilities (GTE, ModernBERT) generally outperform MiniLM on the TREC DL 2019 task. ModernBERT, with its specific optimizations and training configuration using Lion, achieved the overall best effectiveness.

\subsection{Performance Trends Over Epochs and Training Dynamics}
Observing Table \ref{tab:main_results}, performance is not always monotonic with training epochs.
\begin{itemize}
    \item GTE+AdamW peaks early (Epoch 1 or 2) and then degrades.
    \item MiniLM+AdamW shows more gradual improvement or stability up to Epoch 3.
    \item MiniLM+Lion peaks early on TREC (Epoch 1) but late on MS MARCO MRR (Epoch 3).
    \item ModernBERT+Lion mostly peaks around Epoch 2.
    \item ModernBERT+AdamW performance is less consistent, potentially indicating suboptimal interaction with the LR/scheduler setup.
\end{itemize}
This highlights the importance of evaluating at multiple checkpoints. Additionally, analysis of training/evaluation loss curves, learning rate schedules, and gradient norms (Fig. \ref{fig:modernbert_training_plots}) for ModernBERT provides further insight into the training dynamics, showing convergence patterns under the Cosine Annealing schedule with both optimizers.

\begin{figure*}[htbp]
\centering

\begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/modernBERT-Lion-AdamW-eval_loss.png}
    \caption{Eval Loss}
    \label{fig:modernbert_eval_loss}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/modernBERT-Lion-adamW-train_loss.png}
    \caption{Train Loss}
    \label{fig:modernbert_train_loss}
\end{subfigure}

\vspace{0.5cm}

\begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/modernBERT-Lion-adamW-lrate.png}
    \caption{Learning Rate}
    \label{fig:modernbert_lr}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/modernBERT-Lion-adamW_grad.png}
    \caption{Gradient Norm}
    \label{fig:modernbert_grad}
\end{subfigure}

\caption{Training Dynamics for ModernBERT-base with Lion and AdamW optimizers. Each plot shows a different training metric over training steps.}
\label{fig:modernbert_training_plots}
\end{figure*}

\subsection{Comparison with State-of-the-Art}
Table \ref{tab:sota_comparison} situates our best result (ModernBERT + Lion, Epoch 2) relative to other published results on TREC DL 2019 passage ranking.

% \begin{table}[htbp]
% \centering
% \caption{Comparison with Other Systems on TREC DL 2019 (NDCG@10)}
% \label{tab:sota_comparison}
% \begin{tabular}{l c c}
% \toprule
% \textbf{Model/System} & \textbf{NDCG@10} & \textbf{Reference} \\
% \midrule
% \midrule
% \textit{Sparse Retrieval (single stage} & & \\
% \midrule
% BM25 & 0.506 & \cite{lin2020distill} \\
% DeepCT & 0.551 & \cite{gao2021complementing}\\
% doc2query-T5 & 0.642 & \cite{Lin2021PretrainedTF} \\
% \midrule
% \textit{Dense Retrieval Models} & & \\
% \midrule
% ANCE & 0.648 & \cite{xiong2020approximate} \\
% Rand Neg & 0.605 & \cite{xiong2020approximate} \\
% NCE Neg & 0.602 & \cite{xiong2020approximate} \\
% BM25 Neg & 0.664 & \cite{xiong2020approximate} \\
% DPR (BM25 + Rand Neg) & 0.653 & \cite{xiong2020approximate} \\
% BM25 $\rightarrow$ Rand & 0.609 & \cite{xiong2020approximate} \\
% BM25 $\rightarrow$ NCE Neg & 0.608 & \cite{xiong2020approximate} \\
% BM25 $\rightarrow$ BM25 + Rand & 0.648 & \cite{xiong2020approximate} \\
% \midrule
% \textit{Hybrid dense + sparse (single stage} & & \\
% \midrule
% Bi-encoder (PoolAvg) + BM25 & 0.701 & \cite{lin2020distill} \\
% Bi-encoder (TCT-ColBERT) + BM25 & 0.714 & \cite{lin2020distill} \\
% Bi-encoder (PoolAvg) + doc2query-T5 & 0.719 & \cite{lin2020distill} \\
% \midrule
% \textit{Cross-Encoder/Reranking Models} & & \\
% \midrule
% BERT-base Cross-Encoder & 0.634 & \cite{Lin2021PretrainedTF} \\
% BERT-large Cross-Encoder & 0.654 & \cite{Lin2021PretrainedTF} \\
% pt-bert-base-uncased-msmarco & 0.709 & \cite{koursaros2019NBoost} \\
% cross-encoder/ms-marco-electra-base & 0.0.7199 & \cite{reimers2019sentence} \\
% cross-encoder/ms-marco-TinyBERT-L-2-v2 & 0.6984 & \cite{reimers2019sentence} \\
% \midrule
% \textbf{Our Best (ModernBERT+Lion E2)} & \textbf{0.7225} & This work \\
% \textbf{Our GTE+AdamW E1/E2} & \textbf{0.7224} / 0.7203 & This work \\
% \textbf{Our MiniLM+AdamW E3} & 0.7127 & This work \\
% \bottomrule
% \end{tabular}
% \vspace{1em}\\
% \footnotesize{Various references cited in text. Some metrics may vary slightly based on specific evaluation setup or version.}
% \end{table}

% Our best model, ModernBERT+Lion (NDCG@10 = 0.7225), significantly outperforms the BM25 baseline and standard BERT-base/large cross-encoders reported in \cite{Lin2021PretrainedTF}. It achieves performance comparable to or exceeding strong dense retrieval models like TAS-B and reported base-size bi-encoders \cite{zhan2024scaling}, and also exceeds some reported BERT-base cross-encoder results \cite{koursaros2019NBoost}. Our GTE+AdamW model achieves nearly identical peak performance. While not reaching the level of very large models like MonoT5-3B or specialized architectures like ColBERT, our results demonstrate strong performance using base-sized models with appropriate optimizer choices and training strategies.

\begin{table}[htbp]
\centering
\caption{Comparison with Other Systems on TREC DL 2019 (NDCG@10)}
\label{tab:sota_comparison}
\begin{tabular}{l c c}
\toprule
\textbf{Model/System} & \textbf{NDCG@10} & \textbf{Reference} \\
\midrule
\midrule
\textbf{Sparse Retrieval (single stage)} & & \\
\midrule
BM25 & 0.506 & \cite{lin2020distill} \\
DeepCT & 0.551 & \cite{gao2021complementing} \\
doc2query-T5 & 0.642 & \cite{Lin2021PretrainedTF} \\
\midrule
\textbf{Dense Retrieval Models} & & \\
\midrule
ANCE & 0.648 & \cite{xiong2020approximate} \\
Rand Neg & 0.605 & \cite{xiong2020approximate} \\
NCE Neg & 0.602 & \cite{xiong2020approximate} \\
BM25 Neg & 0.664 & \cite{xiong2020approximate} \\
DPR (BM25 + Rand Neg) & 0.653 & \cite{xiong2020approximate} \\
BM25 $\rightarrow$ Rand & 0.609 & \cite{xiong2020approximate} \\
BM25 $\rightarrow$ NCE Neg & 0.608 & \cite{xiong2020approximate} \\
BM25 $\rightarrow$ BM25 + Rand & 0.648 & \cite{xiong2020approximate} \\
\midrule
\textbf{Hybrid dense + sparse (single stage)} & & \\
\midrule
Bi-encoder (PoolAvg) + BM25 & 0.701 & \cite{lin2020distill} \\
Bi-encoder (TCT-ColBERT) + BM25 & 0.714 & \cite{lin2020distill} \\
Bi-encoder (PoolAvg) + doc2query-T5 & 0.719 & \cite{lin2020distill} \\
\midrule
\textbf{Cross-Encoder/Reranking Models} & & \\
\midrule
BERT-base Cross-Encoder & 0.634 & \cite{Lin2021PretrainedTF} \\
BERT-large Cross-Encoder & 0.654 & \cite{Lin2021PretrainedTF} \\
pt-bert-base-uncased-msmarco & 0.709 & \cite{koursaros2019NBoost} \\
cross-encoder/ms-marco-electra-base & 0.719 & \cite{reimers2019sentence} \\
cross-encoder/ms-marco-TinyBERT-L-2-v2 & 0.698 & \cite{reimers2019sentence} \\
\midrule
\textbf{Our Best (ModernBERT+Lion E2)} & \textbf{0.7225} & This work \\
\textbf{Our GTE+AdamW E1/E2} & \textbf{0.7224} / 0.7203 & This work \\
\textbf{Our MiniLM+AdamW E3} & 0.7127 & This work \\
\bottomrule
\end{tabular}
\vspace{1em}\\
\footnotesize{NDCG@10 scores are reported as evaluated on the TREC DL 2019 dataset using standard settings from cited references. Minor variations may occur due to differences in implementation, random seeds, or reranking candidate sets.}
\end{table}

Our best model, ModernBERT+Lion (NDCG@10 = 0.7225), achieves the highest performance among the models listed in Table \ref{tab:sota_comparison}, surpassing strong baselines such as BM25 (0.506), dense retrieval models like ANCE (0.648) and BM25 Neg (0.664), hybrid models such as Bi-encoder (PoolAvg) + doc2query-T5 (0.719), and cross-encoder models including cross-encoder/ms-marco-electra-base (0.7199). Our GTE+AdamW model achieves nearly identical performance (0.7224), while our MiniLM+AdamW variant scores 0.7127, remaining competitive. These results highlight the effectiveness of our optimizer choices and training strategies for base-sized models.

\section{Conclusion}
\label{sec:conclusion}
In this paper, we presented a comparative study of the Lion and AdamW optimizers for fine-tuning three different transformer models (`microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, `answerdotai/ModernBERT-base`) as cross-encoders for passage reranking. Training was performed on MS MARCO, and evaluation used TREC 2019 Deep Learning Track and MS MARCO dev benchmarks. Our findings indicate:
\begin{itemize}
    \item The choice of optimizer significantly interacts with the base model and training hyperparameters (LR, scheduler).
    \item For ModernBERT, using a low learning rate (2e-6) and Cosine Annealing scheduler, Lion substantially outperformed AdamW, achieving the best overall TREC DL 2019 results (NDCG@10 = 0.7225, MAP = 0.5121 at Epochs 1-2).
    \item For GTE, AdamW (with LR 2e-5, no scheduler) was more effective than Lion, achieving strong results (NDCG@10 = 0.7224) comparable to ModernBERT+Lion.
    \item For MiniLM, AdamW yielded slightly better TREC performance over 3 epochs, while Lion achieved the best MS MARCO dev MRR@10 (tied with ModernBERT+Lion).
    \item Models with longer context capabilities (GTE, ModernBERT) generally outperformed MiniLM on the TREC DL task.
    \item Performance varied across epochs, emphasizing the need for checkpoint selection based on validation performance.
\end{itemize}
Our results show that Lion can be a highly effective optimizer for cross-encoder training, particularly for newer architectures like ModernBERT when paired with appropriate learning rate strategies. However, AdamW remains a strong baseline, especially for models like GTE in this setup. The Modal platform proved effective for managing the required GPU resources and experimental setup.

Future work could involve a more thorough hyperparameter search for both optimizers across all models, particularly exploring different learning rates and schedules for Lion with MiniLM and GTE. Investigating the impact of the 8K context length more directly (e.g., on datasets with longer documents) and analyzing memory usage differences between optimizers would also be valuable.

\section*{Acknowledgment}
The authors acknowledge Modal Labs (\url{https://modal.com/}) for providing the cloud computing platform and GPU(s) resources (NVIDIA A100-80GB) used for conducting the experiments presented in this paper.

% --- Bibliography ---
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,references}

\end{document}

### 2025-04-23 cross_encoder/paper/main.tex
Add detailed evaluation metrics and appendices to main paper
- Introduced comprehensive evaluation metrics for all model configurations.
- Added tables for primary ranking metrics, precision and recall analysis, NDCG and MAP metrics, and advanced retrieval metrics.
- Included user-focused metrics and definitions for key evaluation metrics.
- Enhanced the main results table with a reference to the complete metrics in the appendix.

@@
 \usepackage{subcaption}
+\usepackage{booktabs}
+\usepackage{threeparttable}
+\usepackage{tabularx}
 \usepackage{caption}
@@
 Best result for each metric highlighted.
-(*Note: MiniLM+Lion also achieved 0.5988 MRR@10 at epoch 3.)}
+(*Note: MiniLM+Lion also achieved 0.5988 MRR@10 at epoch 3, (see Appendix A for complete metrics).)}
 \end{table*}
@@
 
+\newpage
+\appendix
+\section*{Appendices}
+
+\section{Detailed Evaluation Metrics}
+\label{app:metrics}
+
+This appendix presents comprehensive evaluation metrics for all model configurations tested in our experiments.
+While the main paper focuses on key performance indicators, these detailed tables provide additional insights into model behavior across various retrieval effectiveness measures.
+
+\subsection{Primary Ranking Metrics}
+Table~\ref{tab:primary_metrics} presents the main effectiveness metrics including MAP, NDCG@10, MRR@10, and Precision@10 for all tested configurations.
+
+\begin{table*}[t]
+\centering
+\caption{Primary ranking metrics comparing cross-encoder models across training epochs.}
+\label{tab:primary_metrics}
+\small
+\begin{tabular}{llcccccccc}
+\toprule
+\textbf{Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{MAP} & \textbf{NDCG@10} & \textbf{MRR@10} & \textbf{P@10} & \textbf{Recip\_Rank} & \textbf{R-Prec} & \textbf{bpref} \\
+\midrule
+\multirow{6}{*}{GTE} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.5005 & 0.7224 & 0.5940 & 0.8116 & 0.9814 & 0.4964 & 0.6068 \\
+ & & 2 & 0.4999 & 0.7203 & 0.5942 & 0.8256 & 0.9523 & 0.5018 & 0.6074 \\
+ & & 3 & 0.4899 & 0.6902 & 0.5972 & 0.8093 & 0.9279 & 0.5017 & 0.6075 \\
+\cmidrule{2-10}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.4794 & 0.6970 & 0.5854 & 0.7884 & 0.9612 & 0.4814 & 0.6079 \\
+ & & 2 & 0.4577 & 0.6792 & 0.5957 & 0.7721 & 0.9477 & 0.4642 & 0.6013 \\
+ & & 3 & 0.4521 & 0.6571 & 0.5931 & 0.7488 & 0.9399 & 0.4662 & 0.5972 \\
+\midrule
+\multirow{6}{*}{MiniLM} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.4814 & 0.7008 & 0.5828 & 0.8000 & 0.9465 & 0.4884 & 0.6016 \\
+ & & 2 & 0.4891 & 0.7094 & 0.5818 & 0.8116 & 0.9814 & 0.4916 & 0.6030 \\
+ & & 3 & 0.4908 & 0.7127 & 0.5826 & 0.8116 & 0.9806 & 0.4943 & 0.6048 \\
+\cmidrule{2-10}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.4858 & 0.7031 & 0.5890 & 0.8140 & 0.9457 & 0.4952 & 0.6010 \\
+ & & 2 & 0.4755 & 0.6916 & 0.5942 & 0.8070 & 0.9496 & 0.4803 & 0.5992 \\
+ & & 3 & 0.4706 & 0.6808 & \textbf{0.5988} & 0.8070 & 0.9419 & 0.4809 & 0.5994 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.5066 & 0.7105 & 0.5866 & 0.8163 & 0.9574 & 0.5161 & 0.6139 \\
+ & & 2 & 0.4893 & 0.6839 & 0.5886 & 0.7814 & 0.9341 & 0.4946 & 0.6152 \\
+ & & 3 & 0.4971 & 0.6959 & 0.5916 & 0.7860 & 0.9523 & 0.5116 & 0.6128 \\
+\cmidrule{2-10}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.5121 & 0.7142 & 0.5834 & 0.8163 & 0.9593 & 0.5148 & 0.6140 \\
+ & & 2 & \textbf{0.5115} & \textbf{0.7225} & 0.5908 & \textbf{0.8209} & \textbf{0.9767} & \textbf{0.5183} & \textbf{0.6156} \\
+ & & 3 & 0.5020 & 0.7051 & 0.5989 & 0.8256 & 0.9322 & 0.5102 & 0.6151 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Precision and Recall Analysis}
+Tables~\ref{tab:precision_metrics} and~\ref{tab:recall_metrics} provide detailed precision and recall values at different cutoff thresholds, revealing the models' behavior at various result list depths.
+
+\begin{table*}[t]
+\centering
+\caption{Precision metrics at different cutoff levels.}
+\label{tab:precision_metrics}
+\small
+\begin{tabular}{llccccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{P@5} & \textbf{P@10} & \textbf{P@15} & \textbf{P@20} & \textbf{P@30} & \textbf{P@100} & \textbf{P@200} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.8930 & 0.8116 & 0.7783 & 0.7372 & 0.6419 & 0.3965 & 0.2664 \\
+ & AdamW-2 & 0.8698 & 0.8256 & 0.7752 & 0.7326 & 0.6395 & 0.3963 & 0.2641 \\
+ & AdamW-3 & 0.8558 & 0.8093 & 0.7550 & 0.7128 & 0.6349 & 0.3942 & 0.2634 \\
+ & Lion-1 & 0.8558 & 0.7884 & 0.7364 & 0.7070 & 0.6318 & 0.3893 & 0.2591 \\
+ & Lion-2 & 0.8326 & 0.7721 & 0.7302 & 0.6802 & 0.6054 & 0.3740 & 0.2528 \\
+ & Lion-3 & 0.8140 & 0.7488 & 0.7147 & 0.6907 & 0.6140 & 0.3740 & 0.2543 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.8698 & 0.8000 & 0.7628 & 0.7151 & 0.6450 & 0.3870 & 0.2564 \\
+ & AdamW-2 & 0.8698 & 0.8116 & 0.7674 & 0.7186 & 0.6488 & 0.3895 & 0.2580 \\
+ & AdamW-3 & 0.8558 & 0.8116 & 0.7705 & 0.7256 & 0.6442 & 0.3891 & 0.2580 \\
+ & Lion-1 & 0.8651 & 0.8140 & 0.7783 & 0.7233 & 0.6434 & 0.3847 & 0.2583 \\
+ & Lion-2 & 0.8744 & 0.8070 & 0.7643 & 0.7081 & 0.6264 & 0.3821 & 0.2530 \\
+ & Lion-3 & 0.8465 & 0.8070 & 0.7442 & 0.6930 & 0.6194 & 0.3809 & 0.2547 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.8651 & 0.8163 & 0.7767 & 0.7349 & 0.6612 & 0.3993 & 0.2679 \\
+ & AdamW-2 & 0.8605 & 0.7814 & 0.7333 & 0.6953 & 0.6333 & 0.3912 & 0.2658 \\
+ & AdamW-3 & 0.8465 & 0.7860 & 0.7411 & 0.6988 & 0.6450 & 0.3958 & 0.2655 \\
+ & Lion-1 & 0.8791 & 0.8163 & 0.7783 & 0.7314 & 0.6574 & 0.4033 & 0.2691 \\
+ & Lion-2 & 0.8698 & \textbf{0.8209} & 0.7674 & 0.7279 & 0.6574 & \textbf{0.4016} & 0.2678 \\
+ & Lion-3 & 0.8651 & 0.8256 & 0.7504 & 0.7093 & 0.6488 & 0.3967 & 0.2651 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption{Recall metrics at different cutoff levels.}
+\label{tab:recall_metrics}
+\small
+\begin{tabular}{llcccccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{Recall@5} & \textbf{Recall@10} & \textbf{Recall@15} & \textbf{Recall@20} & \textbf{Recall@30} & \textbf{Recall@100} & \textbf{Recall@200} & \textbf{Recall@500} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.1051 & 0.1689 & 0.2286 & 0.2815 & 0.3376 & 0.5501 & 0.6573 & 0.7161 \\
+ & AdamW-2 & 0.1041 & 0.1760 & 0.2295 & 0.2753 & 0.3366 & 0.5538 & 0.6511 & 0.7130 \\
+ & AdamW-3 & 0.1051 & 0.1715 & 0.2256 & 0.2692 & 0.3376 & 0.5489 & 0.6501 & 0.7102 \\
+ & Lion-1 & 0.1002 & 0.1614 & 0.2158 & 0.2684 & 0.3369 & 0.5401 & 0.6384 & 0.7101 \\
+ & Lion-2 & 0.0973 & 0.1610 & 0.2151 & 0.2545 & 0.3153 & 0.5211 & 0.6273 & 0.7105 \\
+ & Lion-3 & 0.0929 & 0.1542 & 0.2070 & 0.2574 & 0.3221 & 0.5181 & 0.6227 & 0.7124 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.1027 & 0.1652 & 0.2225 & 0.2645 & 0.3362 & 0.5426 & 0.6396 & 0.7187 \\
+ & AdamW-2 & 0.1013 & 0.1668 & 0.2239 & 0.2676 & 0.3401 & 0.5446 & 0.6417 & 0.7223 \\
+ & AdamW-3 & 0.1007 & 0.1702 & 0.2257 & 0.2719 & 0.3386 & 0.5427 & 0.6396 & 0.7189 \\
+ & Lion-1 & 0.1022 & 0.1688 & 0.2257 & 0.2704 & 0.3400 & 0.5338 & 0.6413 & 0.7154 \\
+ & Lion-2 & 0.1025 & 0.1641 & 0.2200 & 0.2613 & 0.3289 & 0.5338 & 0.6268 & 0.7145 \\
+ & Lion-3 & 0.1001 & 0.1676 & 0.2156 & 0.2542 & 0.3253 & 0.5249 & 0.6341 & 0.7143 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.0995 & 0.1678 & 0.2245 & 0.2731 & 0.3475 & 0.5520 & 0.6539 & 0.7224 \\
+ & AdamW-2 & 0.0979 & 0.1634 & 0.2148 & 0.2603 & 0.3321 & 0.5443 & 0.6520 & 0.7198 \\
+ & AdamW-3 & 0.1019 & 0.1623 & 0.2184 & 0.2627 & 0.3380 & 0.5476 & 0.6461 & 0.7198 \\
+ & Lion-1 & 0.1052 & 0.1689 & 0.2276 & 0.2733 & 0.3431 & 0.5542 & 0.6553 & 0.7233 \\
+ & Lion-2 & 0.1048 & \textbf{0.1732} & 0.2245 & 0.2699 & 0.3469 & \textbf{0.5608} & \textbf{0.6553} & \textbf{0.7233} \\
+ & Lion-3 & 0.1037 & 0.1722 & 0.2185 & 0.2640 & 0.3425 & 0.5570 & 0.6525 & 0.7194 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{NDCG and MAP Metrics}
+Tables~\ref{tab:ndcg_metrics} and~\ref{tab:map_metrics} provide a more granular view of ranking quality through normalized discounted cumulative gain and mean average precision at multiple cutoff points.
+
+\begin{table*}[t]
+\centering
+\caption{NDCG metrics at different cutoff levels.}
+\label{tab:ndcg_metrics}
+\small
+\begin{tabular}{llccccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{NDCG} & \textbf{NDCG@5} & \textbf{NDCG@10} & \textbf{NDCG@20} & \textbf{NDCG@30} & \textbf{NDCG@100} & \textbf{NDCG@500} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.6987 & 0.7567 & 0.7224 & 0.7071 & 0.6802 & 0.6591 & 0.6897 \\
+ & AdamW-2 & 0.6949 & 0.7343 & 0.7203 & 0.6967 & 0.6687 & 0.6550 & 0.6845 \\
+ & AdamW-3 & 0.6848 & 0.7080 & 0.6902 & 0.6703 & 0.6558 & 0.6426 & 0.6733 \\
+ & Lion-1 & 0.6865 & 0.7229 & 0.6970 & 0.6786 & 0.6638 & 0.6433 & 0.6750 \\
+ & Lion-2 & 0.6772 & 0.7008 & 0.6792 & 0.6580 & 0.6405 & 0.6248 & 0.6661 \\
+ & Lion-3 & 0.6674 & 0.6801 & 0.6571 & 0.6450 & 0.6310 & 0.6124 & 0.6573 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.6857 & 0.7207 & 0.7008 & 0.6833 & 0.6698 & 0.6402 & 0.6778 \\
+ & AdamW-2 & 0.6910 & 0.7345 & 0.7094 & 0.6908 & 0.6779 & 0.6471 & 0.6843 \\
+ & AdamW-3 & 0.6910 & 0.7244 & 0.7127 & 0.6949 & 0.6749 & 0.6474 & 0.6834 \\
+ & Lion-1 & 0.6841 & 0.7163 & 0.7031 & 0.6829 & 0.6667 & 0.6369 & 0.6752 \\
+ & Lion-2 & 0.6790 & 0.7099 & 0.6916 & 0.6706 & 0.6541 & 0.6295 & 0.6692 \\
+ & Lion-3 & 0.6770 & 0.6837 & 0.6808 & 0.6584 & 0.6479 & 0.6268 & 0.6667 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.6937 & 0.7171 & 0.7105 & 0.6936 & 0.6831 & 0.6555 & 0.6877 \\
+ & AdamW-2 & 0.6869 & 0.7112 & 0.6839 & 0.6701 & 0.6589 & 0.6431 & 0.6795 \\
+ & AdamW-3 & 0.6910 & 0.7156 & 0.6959 & 0.6729 & 0.6699 & 0.6497 & 0.6832 \\
+ & Lion-1 & 0.6984 & 0.7285 & 0.7142 & 0.6991 & 0.6825 & 0.6610 & 0.6925 \\
+ & Lion-2 & \textbf{0.7018} & \textbf{0.7381} & \textbf{0.7225} & \textbf{0.6978} & \textbf{0.6852} & \textbf{0.6638} & \textbf{0.6955} \\
+ & Lion-3 & 0.6930 & 0.7165 & 0.7051 & 0.6797 & 0.6722 & 0.6542 & 0.6857 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{table*}[t]
+\centering
+\caption{MAP metrics at different cutoff levels and related metrics.}
+\label{tab:map_metrics}
+\small
+\begin{tabular}{llccccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{MAP} & \textbf{MAP@10} & \textbf{MAP@20} & \textbf{MAP@30} & \textbf{MAP@100} & \textbf{infAP} & \textbf{gm\_map} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.5005 & 0.1601 & 0.2499 & 0.2915 & 0.4226 & 0.5005 & 0.4143 \\
+ & AdamW-2 & 0.4999 & 0.1632 & 0.2463 & 0.2904 & 0.4247 & 0.4999 & 0.4109 \\
+ & AdamW-3 & 0.4899 & 0.1555 & 0.2350 & 0.2827 & 0.4136 & 0.4899 & 0.3995 \\
+ & Lion-1 & 0.4794 & 0.1468 & 0.2274 & 0.2771 & 0.4039 & 0.4794 & 0.3878 \\
+ & Lion-2 & 0.4577 & 0.1464 & 0.2201 & 0.2631 & 0.3810 & 0.4577 & 0.3623 \\
+ & Lion-3 & 0.4521 & 0.1378 & 0.2145 & 0.2589 & 0.3747 & 0.4521 & 0.3525 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.4814 & 0.1479 & 0.2272 & 0.2809 & 0.4062 & 0.4814 & 0.3920 \\
+ & AdamW-2 & 0.4891 & 0.1531 & 0.2329 & 0.2882 & 0.4148 & 0.4891 & 0.3975 \\
+ & AdamW-3 & 0.4908 & 0.1553 & 0.2383 & 0.2895 & 0.4159 & 0.4908 & 0.3972 \\
+ & Lion-1 & 0.4858 & 0.1534 & 0.2344 & 0.2845 & 0.4087 & 0.4858 & 0.3919 \\
+ & Lion-2 & 0.4755 & 0.1492 & 0.2254 & 0.2735 & 0.4006 & 0.4755 & 0.3823 \\
+ & Lion-3 & 0.4706 & 0.1507 & 0.2207 & 0.2690 & 0.3940 & 0.4706 & 0.3736 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.5066 & 0.1535 & 0.2389 & 0.2942 & 0.4274 & 0.5066 & 0.4045 \\
+ & AdamW-2 & 0.4893 & 0.1493 & 0.2280 & 0.2803 & 0.4097 & 0.4893 & 0.3875 \\
+ & AdamW-3 & 0.4971 & 0.1504 & 0.2304 & 0.2848 & 0.4192 & 0.4971 & 0.3988 \\
+ & Lion-1 & 0.5121 & 0.1577 & 0.2448 & 0.2979 & 0.4347 & 0.5121 & 0.4162 \\
+ & Lion-2 & \textbf{0.5115} & \textbf{0.1616} & \textbf{0.2417} & \textbf{0.2973} & \textbf{0.4333} & \textbf{0.5115} & \textbf{0.4157} \\
+ & Lion-3 & 0.5020 & 0.1583 & 0.2333 & 0.2892 & 0.4239 & 0.5020 & 0.4071 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\subsection{Advanced Retrieval Metrics}
+Table~\ref{tab:interpolated_precision} presents R-Precision and interpolated precision metrics that provide insights into precision at different recall levels.
+
+\begin{table*}[t]
+\centering
+\caption{R-Precision and interpolated precision metrics.}
+\label{tab:interpolated_precision}
+\small
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{llccccccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{R-Prec} & \textbf{R-Prec@0.8} & \textbf{R-Prec@1.0} & \textbf{iprec@0.0} & \textbf{iprec@0.1} & \textbf{iprec@0.3} & \textbf{iprec@0.5} & \textbf{iprec@0.7} & \textbf{iprec@0.9} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.4964 & 0.5617 & 0.4964 & 0.9841 & 0.8741 & 0.7176 & 0.5037 & 0.3295 & 0.1439 \\
+ & AdamW-2 & 0.5018 & 0.5644 & 0.5018 & 0.9730 & 0.8796 & 0.7093 & 0.5027 & 0.3339 & 0.1399 \\
+ & AdamW-3 & 0.5017 & 0.5616 & 0.5017 & 0.9611 & 0.8614 & 0.6942 & 0.4980 & 0.3404 & 0.1395 \\
+ & Lion-1 & 0.4814 & 0.5419 & 0.4814 & 0.9667 & 0.8426 & 0.6601 & 0.4907 & 0.3576 & 0.1233 \\
+ & Lion-2 & 0.4642 & 0.5238 & 0.4642 & 0.9682 & 0.8377 & 0.6283 & 0.4549 & 0.3144 & 0.1182 \\
+ & Lion-3 & 0.4662 & 0.5236 & 0.4662 & 0.9519 & 0.8008 & 0.6289 & 0.4653 & 0.3158 & 0.1218 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.4884 & 0.5502 & 0.4884 & 0.9637 & 0.8546 & 0.6592 & 0.4804 & 0.3474 & 0.1258 \\
+ & AdamW-2 & 0.4916 & 0.5523 & 0.4916 & 0.9821 & 0.8584 & 0.6614 & 0.4856 & 0.3537 & 0.1251 \\
+ & AdamW-3 & 0.4943 & 0.5611 & 0.4943 & 0.9817 & 0.8641 & 0.6724 & 0.4877 & 0.3469 & 0.1262 \\
+ & Lion-1 & 0.4952 & 0.5495 & 0.4952 & 0.9727 & 0.8613 & 0.6698 & 0.4836 & 0.3387 & 0.1326 \\
+ & Lion-2 & 0.4803 & 0.5335 & 0.4803 & 0.9728 & 0.8450 & 0.6441 & 0.4545 & 0.3422 & 0.1261 \\
+ & Lion-3 & 0.4809 & 0.5240 & 0.4809 & 0.9625 & 0.8511 & 0.6443 & 0.4543 & 0.3370 & 0.1278 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.5161 & 0.5804 & 0.5161 & 0.9688 & 0.8676 & 0.6739 & 0.5126 & 0.3858 & 0.1644 \\
+ & AdamW-2 & 0.4946 & 0.5492 & 0.4946 & 0.9583 & 0.8429 & 0.6547 & 0.4956 & 0.3741 & 0.1464 \\
+ & AdamW-3 & 0.5116 & 0.5638 & 0.5116 & 0.9614 & 0.8460 & 0.6781 & 0.5001 & 0.3774 & 0.1489 \\
+ & Lion-1 & 0.5148 & 0.5848 & 0.5148 & 0.9734 & 0.8767 & 0.6892 & 0.5212 & 0.3766 & 0.1566 \\
+ & Lion-2 & \textbf{0.5183} & \textbf{0.5830} & \textbf{0.5183} & \textbf{0.9901} & \textbf{0.8804} & \textbf{0.7052} & \textbf{0.5121} & 0.3853 & \textbf{0.1630} \\
+ & Lion-3 & 0.5102 & 0.5629 & 0.5102 & 0.9660 & 0.8487 & 0.6954 & 0.5021 & \textbf{0.3728} & 0.1597 \\
+\bottomrule
+\end{tabular}%
+}
+\end{table*}
+
+\subsection{User-Focused Metrics}
+Table~\ref{tab:success_metrics} presents metrics that focus on user satisfaction, including success rates and rank-biased precision.
+
+\begin{table*}[t]
+\centering
+\caption{Success and utility metrics.}
+\label{tab:success_metrics}
+\small
+\begin{tabular}{llccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{Success@1} & \textbf{Success@5} & \textbf{Success@10} & \textbf{Utility} & \textbf{RBP} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.9767 & 1.0000 & 1.0000 & -869.1163 & 0.5638 \\
+ & AdamW-2 & 0.9302 & 1.0000 & 1.0000 & -869.1163 & 0.5576 \\
+ & AdamW-3 & 0.8837 & 1.0000 & 1.0000 & -869.1163 & 0.5418 \\
+ & Lion-1 & 0.9302 & 1.0000 & 1.0000 & -869.1163 & 0.5428 \\
+ & Lion-2 & 0.9070 & 1.0000 & 1.0000 & -869.1163 & 0.5279 \\
+ & Lion-3 & 0.9070 & 0.9767 & 1.0000 & -869.1163 & 0.5156 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.9070 & 1.0000 & 1.0000 & -869.1163 & 0.5484 \\
+ & AdamW-2 & 0.9767 & 1.0000 & 1.0000 & -869.1163 & 0.5520 \\
+ & AdamW-3 & 0.9767 & 0.9767 & 1.0000 & -869.1163 & 0.5531 \\
+ & Lion-1 & 0.9070 & 1.0000 & 1.0000 & -869.1163 & 0.5499 \\
+ & Lion-2 & 0.9070 & 1.0000 & 1.0000 & -869.1163 & 0.5430 \\
+ & Lion-3 & 0.8837 & 1.0000 & 1.0000 & -869.1163 & 0.5346 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.9302 & 1.0000 & 1.0000 & -869.1163 & 0.5586 \\
+ & AdamW-2 & 0.8837 & 1.0000 & 1.0000 & -869.1163 & 0.5393 \\
+ & AdamW-3 & 0.9302 & 1.0000 & 1.0000 & -869.1163 & 0.5437 \\
+ & Lion-1 & 0.9302 & 1.0000 & 1.0000 & -869.1163 & 0.5587 \\
+ & Lion-2 & \textbf{0.9535} & 1.0000 & 1.0000 & -869.1163 & \textbf{0.5581} \\
+ & Lion-3 & 0.8837 & 1.0000 & 1.0000 & -869.1163 & 0.5475 \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\section{Metric Definitions}
+\label{app:metric_defs}
+Table~\ref{tab:metric_definitions} provides formal definitions of all evaluation metrics used in our experimental analysis.
+
+\begin{table}[t]
+\centering
+\caption{Definitions of key evaluation metrics used in this paper.}
+\label{tab:metric_definitions}
+\footnotesize
+\begin{tabularx}{\textwidth}{lX}
+\toprule
+\textbf{Metric} & \textbf{Definition} \\
+\midrule
+MAP & Mean Average Precision.
+The mean of average precision scores across all queries, where average precision is the average of precision values calculated at every position where a relevant document is retrieved.
+Range: [0,1].
+\\
+NDCG@k & Normalized Discounted Cumulative Gain at rank k. Measures ranking quality with graded relevance and position-based discount. Range: [0,1]. \\
+P@k & Precision at k. The proportion of retrieved documents in the top k results that are relevant. Range: [0,1]. \\
+Recall@k & Recall at k. The proportion of all relevant documents that are retrieved in the top k results. Range: [0,1]. \\
+MRR@10 & Mean Reciprocal Rank at 10. Average of the reciprocal ranks of the first relevant document within the top 10 results. Range: [0,1]. \\
+R-Prec & R-Precision. Precision after R documents retrieved, where R is the number of relevant documents for the query. Range: [0,1]. \\
+bpref & Binary Preference. Measures ranking quality when relevance judgments are incomplete by considering known relevant/non-relevant document pairs. Range: [0,1]. \\
+Success@k & Indicates whether at least one relevant document is retrieved in the top k results (binary). Range: \{0,1\}. \\
+RBP & Rank-Biased Precision. Geometric discount-based measure of user satisfaction that models a user who browses through ranked results with a certain persistence. Range: [0,1]. \\
+Recip\_Rank & Reciprocal Rank. The reciprocal of the rank at which the first relevant document is retrieved. Range: (0,1]. \\
+\bottomrule
+\end{tabularx}
+\end{table}
+
 \end{document}

### 2025-04-23 cross_encoder/merge_all_metrices.py

@@ -1,0 +1,148 @@
+import os
+import json
+import csv
+import re
+import glob
+from pathlib import Path
+from collections import OrderedDict
+
+def extract_info_from_path(file_path):
+    """Extract model, optimizer, and epoch info from file path"""
+    # Convert to Path object for easier manipulation
+    path = Path(file_path)
+    
+    # Extract components from path
+    try:
+        model = path.parts[-3]  # e.g., 'gte', 'mini_lm', 'modern_bert'
+        optimizer = path.parts[-2]  # e.g., 'AdamW', 'Lion'
+        
+        # Extract epoch number from filename (ep1.json, ep2.json, etc.)
+        epoch_match = re.search(r'ep(\d+)', path.stem)
+        if epoch_match:
+            epoch = epoch_match.group(1)
+        else:
+            # Fallback to filename without extension
+            epoch = path.stem
+        
+        return model, optimizer, epoch
+    except IndexError:
+        print(f"Warning: Could not extract info from path: {file_path}")
+        return "unknown", "unknown", "unknown"
+
+def merge_metrics_to_csv(base_dir, output_file):
+    """
+    Scan all JSON files in the directory structure and merge them into a single CSV,
+    including ALL fields from the JSON files.
+    
+    Args:
+        base_dir: Base directory containing evaluation results
+        output_file: Path to output CSV file
+    """
+    # Find all JSON files recursively
+    json_files = glob.glob(os.path.join(base_dir, "**", "*.json"), recursive=True)
+    print(f"Found {len(json_files)} JSON files to process")
+    
+    # Create a set of all unique field names across all JSON files
+    all_fields = set(["model", "optimizer", "epoch"])  # Start with our metadata fields
+    
+    # First pass: collect all possible field names
+    print("Collecting all metric fields...")
+    for file_path in json_files:
+        try:
+            with open(file_path, 'r') as f:
+                data = json.load(f)
+                # Add all keys from this JSON to our set of field names
+                all_fields.update(data.keys())
+        except Exception as e:
+            print(f"Error reading fields from {file_path}: {e}")
+    
+    # Convert to ordered list and ensure metadata fields come first
+    fields = ["model", "optimizer", "epoch"] + sorted(list(all_fields - {"model", "optimizer", "epoch"}))
+    print(f"Found {len(fields)} unique fields across all JSON files")
+    
+    # Prepare data for CSV
+    results = []
+    
+    # Second pass: extract data for each file
+    for file_path in json_files:
+        try:
[…]

### 2025-04-25 cross_encoder/paper/main.tex
Refactor appendix to include detailed evaluation metrics and remove redundant sections
Updated some citations using xriv paper bibtex

@@
 Information Retrieval (IR) systems aim to satisfy a user's information need, often expressed as a textual query, by returning a ranked list of relevant documents from a large collection.
-A common and effective architecture for modern search systems is the two-stage retrieve-and-rerank pipeline \cite{Lin2021PretrainedTF}.
+A common and effective architecture for modern search systems is the two-stage retrieve-and-rerank pipeline \cite{hu2019retrievereadrerankendtoend}.
 The first stage employs computationally efficient methods, such as BM25 \cite{robertson2009probabilistic} or dense vector retrieval \cite{karpukhin2020dense}, to quickly retrieve a candidate set of documents (typically hundreds or thousands) from the entire collection.
@@
 
-Cross-encoder models, typically based on transformer architectures like BERT \cite{devlin2019bert}, have emerged as state-of-the-art rerankers \cite{nogueira2019passage, Nogueira2020Document}.
+Cross-encoder models, typically based on transformer architectures like BERT \cite{devlin2019bert}, have emerged as state-of-the-art rerankers \cite{nogueira2020passagererankingbert, Nogueira2020Document}.
 Unlike bi-encoder models that encode queries and documents independently, cross-encoders process the query and a candidate document simultaneously (e.g., `[CLS] query [SEP] document [SEP]`).
@@
 The choice of optimizer plays a key role in navigating the complex loss landscape and achieving optimal performance.
-Adam \cite{kingma2014adam} and its variant AdamW \cite{loshchilov2019decoupled} are widely used and generally effective optimizers for training deep neural networks.
+Adam \cite{kingma2017adam} and its variant AdamW \cite{loshchilov2019decoupled} are widely used and generally effective optimizers for training deep neural networks.
 Recently, the Lion (Evo\textbf{L}ved S\textbf{i}gn M\textbf{o}me\textbf{n}tum) optimizer \cite{chen2023symbolic} was proposed, derived through symbolic mathematics and program search.
@@
 \begin{itemize}
-    \item We fine-tune three distinct transformer-based cross-encoder models (`microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`) on the large-scale MS MARCO passage dataset \cite{bajaj2016ms} using both Lion and AdamW optimizers.
+    \item We fine-tune three distinct transformer-based cross-encoder models (`microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`) on the large-scale MS MARCO passage dataset \cite{DBLP:journals/corr/NguyenRSGTMD16} using both Lion and AdamW optimizers.
     \item We utilize different training configurations optimized for specific models, including a lower learning rate (2e-6) and a Cosine Annealing scheduler for ModernBERT, contrasting with a higher rate (2e-5) and no scheduler for MiniLM and GTE.
@@
 The use of BERT-based models for document reranking was popularized by Nogueira et al.
-\cite{nogueira2019passage, Nogueira2020Document}.
+\cite{nogueira2020passagererankingbert, Nogueira2020Document}.
 They demonstrated that fine-tuning BERT as a cross-encoder on relevance labels (like those from MS MARCO) significantly outperforms traditional IR models and even bi-encoder approaches for reranking tasks.
@@
 \subsection{Optimizers for Deep Learning}
-Stochastic Gradient Descent (SGD) with momentum remains a fundamental optimizer, but adaptive learning rate methods like AdaGrad \cite{duchi2011adaptive}, RMSprop \cite{tieleman2012lecture}, and Adam \cite{kingma2014adam} often lead to faster convergence in practice for deep learning models.
+Stochastic Gradient Descent (SGD) with momentum remains a fundamental optimizer, but adaptive learning rate methods like AdaGrad \cite{duchi2011adaptive}, RMSprop \cite{tieleman2012lecture}, and Adam \cite{kingma2017adam} often lead to faster convergence in practice for deep learning models.
 Adam combines momentum with adaptive scaling of learning rates based on estimates of first and second moments of the gradients.
@@
 \subsection{Evaluation Benchmarks}
-The MS MARCO (Microsoft MAchine Reading COmprehension) dataset \cite{bajaj2016ms, DBLP:journals/corr/NguyenRSGTMD16} has become a standard benchmark for training and evaluating deep learning models for passage retrieval and reranking.
+The MS MARCO (Microsoft MAchine Reading COmprehension) dataset \cite{DBLP:journals/corr/NguyenRSGTMD16, DBLP:journals/corr/NguyenRSGTMD16} has become a standard benchmark for training and evaluating deep learning models for passage retrieval and reranking.
 It contains real user queries from Bing and human-judged relevant passages.
@@
 \subsection{Training}
-We fine-tune the cross-encoders using the MS MARCO passage ranking triplets dataset \cite{bajaj2016ms}, as processed by `sentence-transformers` \cite{reimers2019sentence}.
+We fine-tune the cross-encoders using the MS MARCO passage ranking triplets dataset \cite{DBLP:journals/corr/NguyenRSGTMD16}, as processed by `sentence-transformers` \cite{reimers2019sentence}.
 The original dataset contains tuples of (query, positive passage, negative passage).
@@
 \begin{itemize}
-    \item \textbf{Training:} We use the MS MARCO passage ranking triplets dataset \cite{bajaj2016ms}, processed into approximately 2 million query-passage pairs with binary labels using the `sentence-transformers` methodology.
+    \item \textbf{Training:} We use the MS MARCO passage ranking triplets dataset \cite{DBLP:journals/corr/NguyenRSGTMD16}, processed into approximately 2 million query-passage pairs with binary labels using the `sentence-transformers` methodology.
     \item \textbf{Evaluation (Main):} We evaluate on the TREC 2019 Deep Learning Track passage ranking task \cite{craswell2020overview}, which contains 43 queries with graded relevance judgments (qrels).
@@
 
-\newpage
 \appendix
-\section*{Appendices}
 
@@
 Table~\ref{tab:primary_metrics} presents the main effectiveness metrics including MAP, NDCG@10, MRR@10, and Precision@10 for all tested configurations.
+
+\subsection{Precision and Recall Analysis}
+Tables~\ref{tab:precision_metrics} and~\ref{tab:recall_metrics} provide detailed precision and recall values at different cutoff thresholds, revealing the models' behavior at various result list depths.
+
+\subsection{NDCG and MAP Metrics}
+Tables~\ref{tab:ndcg_metrics} and~\ref{tab:map_metrics} provide a more granular view of ranking quality through normalized discounted cumulative gain and mean average precision at multiple cutoff points.
+
+\subsection{Advanced Retrieval Metrics}
+Table~\ref{tab:interpolated_precision} presents R-Precision and interpolated precision metrics that provide insights into precision at different recall levels.
+
+\subsection{User-Focused Metrics}
+Table~\ref{tab:success_metrics} presents metrics that focus on user satisfaction, including success rates and rank-biased precision.
+
+\subsection{Metric Definitions}
+\label{app:metric_defs}
+Table~\ref{tab:metric_definitions} provides formal definitions of all evaluation metrics used in our experimental analysis.
 
@@
 \end{table*}
-
-\subsection{Precision and Recall Analysis}
-Tables~\ref{tab:precision_metrics} and~\ref{tab:recall_metrics} provide detailed precision and recall values at different cutoff thresholds, revealing the models' behavior at various result list depths.
 
@@
 
-\subsection{NDCG and MAP Metrics}
-Tables~\ref{tab:ndcg_metrics} and~\ref{tab:map_metrics} provide a more granular view of ranking quality through normalized discounted cumulative gain and mean average precision at multiple cutoff points.
-
 \begin{table*}[t]
@@
 
-\subsection{Advanced Retrieval Metrics}
-Table~\ref{tab:interpolated_precision} presents R-Precision and interpolated precision metrics that provide insights into precision at different recall levels.
-
 \begin{table*}[t]
@@
 
-\subsection{User-Focused Metrics}
-Table~\ref{tab:success_metrics} presents metrics that focus on user satisfaction, including success rates and rank-biased precision.
-
 \begin{table*}[t]
@@
 
-\section{Metric Definitions}
-\label{app:metric_defs}
-Table~\ref{tab:metric_definitions} provides formal definitions of all evaluation metrics used in our experimental analysis.
-
 \begin{table}[t]

### 2025-04-26 cross_encoder/paper/main.tex
Modified table structres and added all the available metrices for recall, precision, map etc.

@@
 Our experiments, facilitated by the Modal cloud computing platform for GPU resource management, show comparative results across three training epochs.
-ModernBERT trained with Lion achieved the highest NDCG@10 (0.7225) and MAP (0.5121) on TREC DL 2019, while MiniLM trained with Lion achieved the highest MRR@10 (0.5988) on MS MARCO dev.
+ModernBERT trained with Lion achieved the highest NDCG@10 (0.7225) and MAP (0.5121) on TREC DL 2019, while MiniLM trained with Lion tied with ModernBERT with Lion on MRR@10 (0.5988) on MS MARCO dev.
 We analyze the performance trends based on standard IR metrics, providing insights into the relative effectiveness of Lion versus AdamW for different model architectures and training configurations in the context of passage reranking.
@@
 Best result for each metric highlighted.
-(*Note: MiniLM+Lion also achieved 0.5988 MRR@10 at epoch 3, (see Appendix A for complete metrics).)}
+(*Note: MiniLM+Lion also achieved 0.5988 MRR@10 at epoch 3, (see \autoref{app:metrics} for complete metrics).)}
 \end{table*}
@@
 \multirow{6}{*}{GTE} & \multirow{3}{*}{AdamW} 
- & 1 & 0.5005 & 0.7224 & 0.5940 & 0.8116 & 0.9814 & 0.4964 & 0.6068 \\
- & & 2 & 0.4999 & 0.7203 & 0.5942 & 0.8256 & 0.9523 & 0.5018 & 0.6074 \\
+ & 1 & 0.5005 & 0.7224 & 0.5940 & 0.8116 & \textbf{0.9814} & 0.4964 & 0.6068 \\
+ & & 2 & 0.4999 & 0.7203 & 0.5942 & \textbf{0.8256} & 0.9523 & 0.5018 & 0.6074 \\
  & & 3 & 0.4899 & 0.6902 & 0.5972 & 0.8093 & 0.9279 & 0.5017 & 0.6075 \\
@@
  & 1 & 0.4814 & 0.7008 & 0.5828 & 0.8000 & 0.9465 & 0.4884 & 0.6016 \\
- & & 2 & 0.4891 & 0.7094 & 0.5818 & 0.8116 & 0.9814 & 0.4916 & 0.6030 \\
+ & & 2 & 0.4891 & 0.7094 & 0.5818 & 0.8116 & \textbf{0.9814} & 0.4916 & 0.6030 \\
  & & 3 & 0.4908 & 0.7127 & 0.5826 & 0.8116 & 0.9806 & 0.4943 & 0.6048 \\
@@
  & \multirow{3}{*}{Lion} 
- & 1 & 0.5121 & 0.7142 & 0.5834 & 0.8163 & 0.9593 & 0.5148 & 0.6140 \\
- & & 2 & \textbf{0.5115} & \textbf{0.7225} & 0.5908 & \textbf{0.8209} & \textbf{0.9767} & \textbf{0.5183} & \textbf{0.6156} \\
- & & 3 & 0.5020 & 0.7051 & 0.5989 & 0.8256 & 0.9322 & 0.5102 & 0.6151 \\
+ & 1 & \textbf{0.5121} & 0.7142 & 0.5834 & 0.8163 & 0.9593 & 0.5148 & 0.6140 \\
+ & & 2 & 0.5115 & \textbf{0.7225} & 0.5908 & 0.8209 & 0.9767 & \textbf{0.5183} & \textbf{0.6156} \\
+ & & 3 & 0.5020 & 0.7051 & \textbf{0.5988} & \textbf{0.8256} & 0.9322 & 0.5102 & 0.6151 \\
 \bottomrule
@@
 \small
-\begin{tabular}{llccccccc}
+\begin{tabular}{llcccccccccc}
 \toprule
-\textbf{Model} & \textbf{Config} & \textbf{P@5} & \textbf{P@10} & \textbf{P@15} & \textbf{P@20} & \textbf{P@30} & \textbf{P@100} & \textbf{P@200} \\
+\textbf{Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{P@5} & \textbf{P@10} & \textbf{P@15} & \textbf{P@20} & \textbf{P@30} & \textbf{P@100} & \textbf{P@200} & \textbf{P@500} & \textbf{P@1000}\\
 \midrule
-\multirow{6}{*}{GTE} & AdamW-1 & 0.8930 & 0.8116 & 0.7783 & 0.7372 & 0.6419 & 0.3965 & 0.2664 \\
- & AdamW-2 & 0.8698 & 0.8256 & 0.7752 & 0.7326 & 0.6395 & 0.3963 & 0.2641 \\
- & AdamW-3 & 0.8558 & 0.8093 & 0.7550 & 0.7128 & 0.6349 & 0.3942 & 0.2634 \\
- & Lion-1 & 0.8558 & 0.7884 & 0.7364 & 0.7070 & 0.6318 & 0.3893 & 0.2591 \\
- & Lion-2 & 0.8326 & 0.7721 & 0.7302 & 0.6802 & 0.6054 & 0.3740 & 0.2528 \\
- & Lion-3 & 0.8140 & 0.7488 & 0.7147 & 0.6907 & 0.6140 & 0.3740 & 0.2543 \\
+\multirow{6}{*}{GTE} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.8930 & 0.8116 & 0.7783 & 0.7372 & 0.6419 & 0.3965 & 0.2664 & 0.1238 & 0.0654 \\
+ & & 2 & 0.8698 & 0.8256 & 0.7752 & 0.7326 & 0.6395 & 0.3963 & 0.2641 & 0.1230 & 0.0654 \\
+ & & 3 & 0.8558 & 0.8093 & 0.7550 & 0.7128 & 0.6349 & 0.3942 & 0.2634 & 0.1225 & 0.0654 \\
+\cmidrule{2-12}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.8558 & 0.7884 & 0.7364 & 0.7070 & 0.6318 & 0.3893 & 0.2591 & 0.1228 & 0.0654 \\
+ & & 2 & 0.8326 & 0.7721 & 0.7302 & 0.6802 & 0.6054 & 0.3740 & 0.2528 & 0.1227 & 0.0654 \\
+ & & 3 & 0.8140 & 0.7488 & 0.7147 & 0.6907 & 0.6140 & 0.3740 & 0.2543 & 0.1231 & 0.0654 \\
 \midrule
-\multirow{6}{*}{MiniLM} & AdamW-1 & 0.8698 & 0.8000 & 0.7628 & 0.7151 & 0.6450 & 0.3870 & 0.2564 \\
- & AdamW-2 & 0.8698 & 0.8116 & 0.7674 & 0.7186 & 0.6488 & 0.3895 & 0.2580 \\
- & AdamW-3 & 0.8558 & 0.8116 & 0.7705 & 0.7256 & 0.6442 & 0.3891 & 0.2580 \\
- & Lion-1 & 0.8651 & 0.8140 & 0.7783 & 0.7233 & 0.6434 & 0.3847 & 0.2583 \\
- & Lion-2 & 0.8744 & 0.8070 & 0.7643 & 0.7081 & 0.6264 & 0.3821 & 0.2530 \\
- & Lion-3 & 0.8465 & 0.8070 & 0.7442 & 0.6930 & 0.6194 & 0.3809 & 0.2547 \\
+\multirow{6}{*}{MiniLM} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.8698 & 0.8000 & 0.7628 & 0.7151 & 0.6450 & 0.3870 & 0.2564 & 0.1242 & 0.0654 \\
+ & & 2 & 0.8698 & 0.8116 & 0.7674 & 0.7186 & 0.6488 & 0.3895 & 0.2580 & 0.1247 & 0.0654 \\
+ & & 3 & 0.8558 & 0.8116 & 0.7705 & 0.7256 & 0.6442 & 0.3891 & 0.2580 & 0.1245 & 0.0654 \\
+\cmidrule{2-12}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.8651 & 0.8140 & 0.7783 & 0.7233 & 0.6434 & 0.3847 & 0.2583 & 0.1242 & 0.0654 \\
+ & & 2 & 0.8744 & 0.8070 & 0.7643 & 0.7081 & 0.6264 & 0.3821 & 0.2530 & 0.1240 & 0.0654 \\
+ & & 3 & 0.8465 & 0.8070 & 0.7442 & 0.6930 & 0.6194 & 0.3809 & 0.2547 & 0.1240 & 0.0654 \\
 \midrule
-\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.8651 & 0.8163 & 0.7767 & 0.7349 & 0.6612 & 0.3993 & 0.2679 \\
- & AdamW-2 & 0.8605 & 0.7814 & 0.7333 & 0.6953 & 0.6333 & 0.3912 & 0.2658 \\
- & AdamW-3 & 0.8465 & 0.7860 & 0.7411 & 0.6988 & 0.6450 & 0.3958 & 0.2655 \\
- & Lion-1 & 0.8791 & 0.8163 & 0.7783 & 0.7314 & 0.6574 & 0.4033 & 0.2691 \\
- & Lion-2 & 0.8698 & \textbf{0.8209} & 0.7674 & 0.7279 & 0.6574 & \textbf{0.4016} & 0.2678 \\
- & Lion-3 & 0.8651 & 0.8256 & 0.7504 & 0.7093 & 0.6488 & 0.3967 & 0.2651 \\
+\multirow{6}{*}{ModernBERT} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.8651 & 0.8163 & 0.7767 & 0.7349 & 0.6612 & 0.3993 & 0.2679 & 0.1262 & 0.0654\\
+ & & 2 & 0.8605 & 0.7814 & 0.7333 & 0.6953 & 0.6333 & 0.3912 & 0.2658 & 0.1257 & 0.0654 \\
+ & & 3 & 0.8465 & 0.7860 & 0.7411 & 0.6988 & 0.6450 & 0.3958 & 0.2655 & 0.1253 & 0.0654 \\
+\cmidrule{2-12}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.8791 & 0.8163 & 0.7783 & 0.7314 & 0.6574 & 0.4033 & 0.2691 & 0.1265 & 0.0654\\
+ & & 2 & 0.8698 & \textbf{0.8209} & 0.7674 & 0.7279 & 0.6574 & \textbf{0.4016} & 0.2678 & 0.1267 & 0.0654\\
+ & & 3 & 0.8651 & 0.8256 & 0.7504 & 0.7093 & 0.6488 & 0.3967 & 0.2651 & 0.1263 & 0.0654 \\
 \bottomrule
@@
 \end{table*}
+
+% \begin{table*}[t]
+% \centering
+% \caption{Precision metrics at different cutoff levels.}
+% \label{tab:precision_metrics}
+% \small
+% \begin{tabular}{llccccccc}
+% \toprule
+% \textbf{Model} & \textbf{Config} & \textbf{P@5} & \textbf{P@10} & \textbf{P@15} & \textbf{P@20} & \textbf{P@30} & \textbf{P@100} & \textbf{P@200} \\
+% \midrule
+% \multirow{6}{*}{GTE} & AdamW-1 & 0.8930 & 0.8116 & 0.7783 & 0.7372 & 0.6419 & 0.3965 & 0.2664 \\
+%  & AdamW-2 & 0.8698 & 0.8256 & 0.7752 & 0.7326 & 0.6395 & 0.3963 & 0.2641 \\
+%  & AdamW-3 & 0.8558 & 0.8093 & 0.7550 & 0.7128 & 0.6349 & 0.3942 & 0.2634 \\
+%  & Lion-1 & 0.8558 & 0.7884 & 0.7364 & 0.7070 & 0.6318 & 0.3893 & 0.2591 \\
+%  & Lion-2 & 0.8326 & 0.7721 & 0.7302 & 0.6802 & 0.6054 & 0.3740 & 0.2528 \\
+%  & Lion-3 & 0.8140 & 0.7488 & 0.7147 & 0.6907 & 0.6140 & 0.3740 & 0.2543 \\
+% \midrule
+% \multirow{6}{*}{MiniLM} & AdamW-1 & 0.8698 & 0.8000 & 0.7628 & 0.7151 & 0.6450 & 0.3870 & 0.2564 \\
+%  & AdamW-2 & 0.8698 & 0.8116 & 0.7674 & 0.7186 & 0.6488 & 0.3895 & 0.2580 \\
+%  & AdamW-3 & 0.8558 & 0.8116 & 0.7705 & 0.7256 & 0.6442 & 0.3891 & 0.2580 \\
+%  & Lion-1 & 0.8651 & 0.8140 & 0.7783 & 0.7233 & 0.6434 & 0.3847 & 0.2583 \\
+%  & Lion-2 & 0.8744 & 0.8070 & 0.7643 & 0.7081 & 0.6264 & 0.3821 & 0.2530 \\
+%  & Lion-3 & 0.8465 & 0.8070 & 0.7442 & 0.6930 & 0.6194 & 0.3809 & 0.2547 \\
+% \midrule
+% \multirow{6}{*}{ModernBERT} & AdamW-1 & 0.8651 & 0.8163 & 0.7767 & 0.7349 & 0.6612 & 0.3993 & 0.2679 \\
+%  & AdamW-2 & 0.8605 & 0.7814 & 0.7333 & 0.6953 & 0.6333 & 0.3912 & 0.2658 \\
+%  & AdamW-3 & 0.8465 & 0.7860 & 0.7411 & 0.6988 & 0.6450 & 0.3958 & 0.2655 \\
+%  & Lion-1 & 0.8791 & 0.8163 & 0.7783 & 0.7314 & 0.6574 & 0.4033 & 0.2691 \\
+%  & Lion-2 & 0.8698 & \textbf{0.8209} & 0.7674 & 0.7279 & 0.6574 & \textbf{0.4016} & 0.2678 \\
+%  & Lion-3 & 0.8651 & 0.8256 & 0.7504 & 0.7093 & 0.6488 & 0.3967 & 0.2651 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 
@@
 \label{tab:recall_metrics}
-\small
-\begin{tabular}{llcccccccc}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{llcccccccccc}
 \toprule
-\textbf{Model} & \textbf{Config} & \textbf{Recall@5} & \textbf{Recall@10} & \textbf{Recall@15} & \textbf{Recall@20} & \textbf{Recall@30} & \textbf{Recall@100} & \textbf{Recall@200} & \textbf{Recall@500} \\
+\textbf{Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{Recall@5} & \textbf{Recall@10} & \textbf{Recall@15} & \textbf{Recall@20} & \textbf{Recall@30} & \textbf{Recall@100} & \textbf{Recall@200} & \textbf{Recall@500} & \textbf{Recall@1000} \\
 \midrule
-\multirow{6}{*}{GTE} & AdamW-1 & 0.1051 & 0.1689 & 0.2286 & 0.2815 & 0.3376 & 0.5501 & 0.6573 & 0.7161 \\
- & AdamW-2 & 0.1041 & 0.1760 & 0.2295 & 0.2753 & 0.3366 & 0.5538 & 0.6511 & 0.7130 \\
- & AdamW-3 & 0.1051 & 0.1715 & 0.2256 & 0.2692 & 0.3376 & 0.5489 & 0.6501 & 0.7102 \\
- & Lion-1 & 0.1002 & 0.1614 & 0.2158 & 0.2684 & 0.3369 & 0.5401 & 0.6384 & 0.7101 \\
- & Lion-2 & 0.0973 & 0.1610 & 0.2151 & 0.2545 & 0.3153 & 0.5211 & 0.6273 & 0.7105 \\
- & Lion-3 & 0.0929 & 0.1542 & 0.2070 & 0.2574 & 0.3221 & 0.5181 & 0.6227 & 0.7124 \\
+\multirow{6}{*}{GTE} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.1051 & 0.1689 & 0.2286 & 0.2815 & 0.3376 & 0.5501 & 0.6573 & 0.7161 & 0.7389 \\
+ & & 2 & 0.1041 & 0.1760 & 0.2295 & 0.2753 & 0.3366 & 0.5538 & 0.6511 & 0.7130 & 0.7389 \\
+ & & 3 & 0.1051 & 0.1715 & 0.2256 & 0.2692 & 0.3376 & 0.5489 & 0.6501 & 0.7102 & 0.7389 \\
+\cmidrule{2-12}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.1002 & 0.1614 & 0.2158 & 0.2684 & 0.3369 & 0.5401 & 0.6384 & 0.7101 & 0.7389 \\
+ & & 2 & 0.0973 & 0.1610 & 0.2151 & 0.2545 & 0.3153 & 0.5211 & 0.6273 & 0.7105 & 0.7389 \\
+ & & 3 & 0.0929 & 0.1542 & 0.2070 & 0.2574 & 0.3221 & 0.5181 & 0.6227 & 0.7124 & 0.7389 \\
 \midrule
-\multirow{6}{*}{MiniLM} & AdamW-1 & 0.1027 & 0.1652 & 0.2225 & 0.2645 & 0.3362 & 0.5426 & 0.6396 & 0.7187 \\
- & AdamW-2 & 0.1013 & 0.1668 & 0.2239 & 0.2676 & 0.3401 & 0.5446 & 0.6417 & 0.7223 \\
- & AdamW-3 & 0.1007 & 0.1702 & 0.2257 & 0.2719 & 0.3386 & 0.5427 & 0.6396 & 0.7189 \\
- & Lion-1 & 0.1022 & 0.1688 & 0.2257 & 0.2704 & 0.3400 & 0.5338 & 0.6413 & 0.7154 \\
- & Lion-2 & 0.1025 & 0.1641 & 0.2200 & 0.2613 & 0.3289 & 0.5338 & 0.6268 & 0.7145 \\
- & Lion-3 & 0.1001 & 0.1676 & 0.2156 & 0.2542 & 0.3253 & 0.5249 & 0.6341 & 0.7143 \\
+\multirow{6}{*}{MiniLM} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.1027 & 0.1652 & 0.2225 & 0.2645 & 0.3362 & 0.5426 & 0.6396 & 0.7187 & 0.7389 \\
+ & & 2 & 0.1013 & 0.1668 & 0.2239 & 0.2676 & 0.3401 & 0.5446 & 0.6417 & 0.7223 & 0.7389 \\
+ & & 3 & 0.1007 & 0.1702 & 0.2257 & 0.2719 & 0.3386 & 0.5427 & 0.6396 & 0.7189 & 0.7389 \\
+\cmidrule{2-12}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.1022 & 0.1688 & 0.2257 & 0.2704 & 0.3400 & 0.5338 & 0.6413 & 0.7154 & 0.7389 \\
+ & & 2 & 0.1025 & 0.1641 & 0.2200 & 0.2613 & 0.3289 & 0.5338 & 0.6268 & 0.7145 & 0.7389 \\
+ & & 3 & 0.1001 & 0.1676 & 0.2156 & 0.2542 & 0.3253 & 0.5249 & 0.6341 & 0.7143 & 0.7389 \\
 \midrule
-\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.0995 & 0.1678 & 0.2245 & 0.2731 & 0.3475 & 0.5520 & 0.6539 & 0.7224 \\
- & AdamW-2 & 0.0979 & 0.1634 & 0.2148 & 0.2603 & 0.3321 & 0.5443 & 0.6520 & 0.7198 \\
- & AdamW-3 & 0.1019 & 0.1623 & 0.2184 & 0.2627 & 0.3380 & 0.5476 & 0.6461 & 0.7198 \\
- & Lion-1 & 0.1052 & 0.1689 & 0.2276 & 0.2733 & 0.3431 & 0.5542 & 0.6553 & 0.7233 \\
- & Lion-2 & 0.1048 & \textbf{0.1732} & 0.2245 & 0.2699 & 0.3469 & \textbf{0.5608} & \textbf{0.6553} & \textbf{0.7233} \\
- & Lion-3 & 0.1037 & 0.1722 & 0.2185 & 0.2640 & 0.3425 & 0.5570 & 0.6525 & 0.7194 \\
+\multirow{6}{*}{ModernBERT} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.0995 & 0.1678 & 0.2245 & 0.2731 & 0.3475 & 0.5520 & 0.6539 & 0.7224 & 0.7389 \\
+ & & 2 & 0.0979 & 0.1634 & 0.2148 & 0.2603 & 0.3321 & 0.5443 & 0.6520 & 0.7198 & 0.7389 \\
+ & & 3 & 0.1019 & 0.1623 & 0.2184 & 0.2627 & 0.3380 & 0.5476 & 0.6461 & 0.7198 & 0.7389 \\
+\cmidrule{2-12}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.1052 & 0.1689 & 0.2276 & 0.2733 & 0.3431 & 0.5542 & 0.6553 & 0.7233 & 0.7389 \\
+ & & 2 & 0.1048 & \textbf{0.1732} & 0.2245 & 0.2699 & 0.3469 & \textbf{0.5608} & \textbf{0.6553} & \textbf{0.7233} & 0.7389 \\
+ & & 3 & 0.1037 & 0.1722 & 0.2185 & 0.2640 & 0.3425 & 0.5570 & 0.6525 & 0.7194 & 0.7389 \\
 \bottomrule
 \end{tabular}
+}
 \end{table*}
+
+% \begin{table*}[t]
+% \centering
+% \caption{Recall metrics at different cutoff levels.}
+% \label{tab:recall_metrics}
+% \small
+% \begin{tabular}{llcccccccc}
+% \toprule
+% \textbf{Model} & \textbf{Config} & \textbf{Recall@5} & \textbf{Recall@10} & \textbf{Recall@15} & \textbf{Recall@20} & \textbf{Recall@30} & \textbf{Recall@100} & \textbf{Recall@200} & \textbf{Recall@500} \\
+% \midrule
+% \multirow{6}{*}{GTE} & AdamW-1 & 0.1051 & 0.1689 & 0.2286 & 0.2815 & 0.3376 & 0.5501 & 0.6573 & 0.7161 \\
+%  & AdamW-2 & 0.1041 & 0.1760 & 0.2295 & 0.2753 & 0.3366 & 0.5538 & 0.6511 & 0.7130 \\
+%  & AdamW-3 & 0.1051 & 0.1715 & 0.2256 & 0.2692 & 0.3376 & 0.5489 & 0.6501 & 0.7102 \\
+%  & Lion-1 & 0.1002 & 0.1614 & 0.2158 & 0.2684 & 0.3369 & 0.5401 & 0.6384 & 0.7101 \\
+%  & Lion-2 & 0.0973 & 0.1610 & 0.2151 & 0.2545 & 0.3153 & 0.5211 & 0.6273 & 0.7105 \\
+%  & Lion-3 & 0.0929 & 0.1542 & 0.2070 & 0.2574 & 0.3221 & 0.5181 & 0.6227 & 0.7124 \\
+% \midrule
+% \multirow{6}{*}{MiniLM} & AdamW-1 & 0.1027 & 0.1652 & 0.2225 & 0.2645 & 0.3362 & 0.5426 & 0.6396 & 0.7187 \\
+%  & AdamW-2 & 0.1013 & 0.1668 & 0.2239 & 0.2676 & 0.3401 & 0.5446 & 0.6417 & 0.7223 \\
+%  & AdamW-3 & 0.1007 & 0.1702 & 0.2257 & 0.2719 & 0.3386 & 0.5427 & 0.6396 & 0.7189 \\
+%  & Lion-1 & 0.1022 & 0.1688 & 0.2257 & 0.2704 & 0.3400 & 0.5338 & 0.6413 & 0.7154 \\
+%  & Lion-2 & 0.1025 & 0.1641 & 0.2200 & 0.2613 & 0.3289 & 0.5338 & 0.6268 & 0.7145 \\
+%  & Lion-3 & 0.1001 & 0.1676 & 0.2156 & 0.2542 & 0.3253 & 0.5249 & 0.6341 & 0.7143 \\
+% \midrule
+% \multirow{6}{*}{ModernBERT} & AdamW-1 & 0.0995 & 0.1678 & 0.2245 & 0.2731 & 0.3475 & 0.5520 & 0.6539 & 0.7224 \\
+%  & AdamW-2 & 0.0979 & 0.1634 & 0.2148 & 0.2603 & 0.3321 & 0.5443 & 0.6520 & 0.7198 \\
+%  & AdamW-3 & 0.1019 & 0.1623 & 0.2184 & 0.2627 & 0.3380 & 0.5476 & 0.6461 & 0.7198 \\
+%  & Lion-1 & 0.1052 & 0.1689 & 0.2276 & 0.2733 & 0.3431 & 0.5542 & 0.6553 & 0.7233 \\
+%  & Lion-2 & 0.1048 & \textbf{0.1732} & 0.2245 & 0.2699 & 0.3469 & \textbf{0.5608} & \textbf{0.6553} & \textbf{0.7233} \\
+%  & Lion-3 & 0.1037 & 0.1722 & 0.2185 & 0.2640 & 0.3425 & 0.5570 & 0.6525 & 0.7194 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 
@@
 \label{tab:ndcg_metrics}
-\small
-\begin{tabular}{llccccccc}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{lllcccccccc}
 \toprule
-\textbf{Model} & \textbf{Config} & \textbf{NDCG} & \textbf{NDCG@5} & \textbf{NDCG@10} & \textbf{NDCG@20} & \textbf{NDCG@30} & \textbf{NDCG@100} & \textbf{NDCG@500} \\
+\textbf{Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{NDCG} & \textbf{NDCG@5} & \textbf{NDCG@10} & \textbf{NDCG@20} & \textbf{NDCG@30} & \textbf{NDCG@100} & \textbf{NDCG@500} & \textbf{NDCG@1000} \\
 \midrule
-\multirow{6}{*}{GTE} & AdamW-1 & 0.6987 & 0.7567 & 0.7224 & 0.7071 & 0.6802 & 0.6591 & 0.6897 \\
- & AdamW-2 & 0.6949 & 0.7343 & 0.7203 & 0.6967 & 0.6687 & 0.6550 & 0.6845 \\
- & AdamW-3 & 0.6848 & 0.7080 & 0.6902 & 0.6703 & 0.6558 & 0.6426 & 0.6733 \\
- & Lion-1 & 0.6865 & 0.7229 & 0.6970 & 0.6786 & 0.6638 & 0.6433 & 0.6750 \\
- & Lion-2 & 0.6772 & 0.7008 & 0.6792 & 0.6580 & 0.6405 & 0.6248 & 0.6661 \\
- & Lion-3 & 0.6674 & 0.6801 & 0.6571 & 0.6450 & 0.6310 & 0.6124 & 0.6573 \\
+\multirow{6}{*}{GTE} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.6987 & 0.7567 & 0.7224 & 0.7071 & 0.6802 & 0.6591 & 0.6897 & 0.6987 \\
+ & & 2 & 0.6949 & 0.7343 & 0.7203 & 0.6967 & 0.6687 & 0.6550 & 0.6845 & 0.6949 \\
+ & & 3 & 0.6848 & 0.7080 & 0.6902 & 0.6703 & 0.6558 & 0.6426 & 0.6733 & 0.6848 \\
+\cmidrule{2-11}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.6865 & 0.7229 & 0.6970 & 0.6786 & 0.6638 & 0.6433 & 0.6750 & 0.6865 \\
+ & & 2 & 0.6772 & 0.7008 & 0.6792 & 0.6580 & 0.6405 & 0.6248 & 0.6661 & 0.6772 \\
+ & & 3 & 0.6674 & 0.6801 & 0.6571 & 0.6450 & 0.6310 & 0.6124 & 0.6573 & 0.6674 \\
 \midrule
-\multirow{6}{*}{MiniLM} & AdamW-1 & 0.6857 & 0.7207 & 0.7008 & 0.6833 & 0.6698 & 0.6402 & 0.6778 \\
- & AdamW-2 & 0.6910 & 0.7345 & 0.7094 & 0.6908 & 0.6779 & 0.6471 & 0.6843 \\
- & AdamW-3 & 0.6910 & 0.7244 & 0.7127 & 0.6949 & 0.6749 & 0.6474 & 0.6834 \\
- & Lion-1 & 0.6841 & 0.7163 & 0.7031 & 0.6829 & 0.6667 & 0.6369 & 0.6752 \\
- & Lion-2 & 0.6790 & 0.7099 & 0.6916 & 0.6706 & 0.6541 & 0.6295 & 0.6692 \\
- & Lion-3 & 0.6770 & 0.6837 & 0.6808 & 0.6584 & 0.6479 & 0.6268 & 0.6667 \\
+\multirow{6}{*}{MiniLM} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.6857 & 0.7207 & 0.7008 & 0.6833 & 0.6698 & 0.6402 & 0.6778 & 0.6857 \\
+ & & 2 & 0.6910 & 0.7345 & 0.7094 & 0.6908 & 0.6779 & 0.6471 & 0.6843 & 0.6910 \\
+ & & 3 & 0.6910 & 0.7244 & 0.7127 & 0.6949 & 0.6749 & 0.6474 & 0.6834 & 0.6910 \\
+\cmidrule{2-11}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.6841 & 0.7163 & 0.7031 & 0.6829 & 0.6667 & 0.6369 & 0.6752 & 0.6841 \\
+ & & 2 & 0.6790 & 0.7099 & 0.6916 & 0.6706 & 0.6541 & 0.6295 & 0.6692 & 0.6790 \\
+ & & 3 & 0.6770 & 0.6837 & 0.6808 & 0.6584 & 0.6479 & 0.6268 & 0.6667 & 0.6770 \\
 \midrule
-\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.6937 & 0.7171 & 0.7105 & 0.6936 & 0.6831 & 0.6555 & 0.6877 \\
- & AdamW-2 & 0.6869 & 0.7112 & 0.6839 & 0.6701 & 0.6589 & 0.6431 & 0.6795 \\
- & AdamW-3 & 0.6910 & 0.7156 & 0.6959 & 0.6729 & 0.6699 & 0.6497 & 0.6832 \\
- & Lion-1 & 0.6984 & 0.7285 & 0.7142 & 0.6991 & 0.6825 & 0.6610 & 0.6925 \\
- & Lion-2 & \textbf{0.7018} & \textbf{0.7381} & \textbf{0.7225} & \textbf{0.6978} & \textbf{0.6852} & \textbf{0.6638} & \textbf{0.6955} \\
- & Lion-3 & 0.6930 & 0.7165 & 0.7051 & 0.6797 & 0.6722 & 0.6542 & 0.6857 \\
+\multirow{6}{*}{ModernBERT} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.6937 & 0.7171 & 0.7105 & 0.6936 & 0.6831 & 0.6555 & 0.6877 & 0.6937 \\
+ & & 2 & 0.6869 & 0.7112 & 0.6839 & 0.6701 & 0.6589 & 0.6431 & 0.6795 & 0.6869 \\
+ & & 3 & 0.6910 & 0.7156 & 0.6959 & 0.6729 & 0.6699 & 0.6497 & 0.6832 & 0.6910 \\
+\cmidrule{2-11}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.6984 & 0.7285 & 0.7142 & 0.6991 & 0.6825 & 0.6610 & 0.6925 & 0.6984 \\
+ & & 2 & \textbf{0.7018} & \textbf{0.7381} & \textbf{0.7225} & \textbf{0.6978} & \textbf{0.6852} & \textbf{0.6638} & \textbf{0.6955} & 0.7018 \\
+ & & 3 & 0.6930 & 0.7165 & 0.7051 & 0.6797 & 0.6722 & 0.6542 & 0.6857 & 0.6930 \\
 \bottomrule
 \end{tabular}
+}
 \end{table*}
+
+% \begin{table*}[t]
+% \centering
+% \caption{NDCG metrics at different cutoff levels.}
+% \label{tab:ndcg_metrics}
+% \small
+% \begin{tabular}{llccccccc}
+% \toprule
+% \textbf{Model} & \textbf{Config} & \textbf{NDCG} & \textbf{NDCG@5} & \textbf{NDCG@10} & \textbf{NDCG@20} & \textbf{NDCG@30} & \textbf{NDCG@100} & \textbf{NDCG@500} \\
+% \midrule
+% \multirow{6}{*}{GTE} & AdamW-1 & 0.6987 & 0.7567 & 0.7224 & 0.7071 & 0.6802 & 0.6591 & 0.6897 \\
+%  & AdamW-2 & 0.6949 & 0.7343 & 0.7203 & 0.6967 & 0.6687 & 0.6550 & 0.6845 \\
+%  & AdamW-3 & 0.6848 & 0.7080 & 0.6902 & 0.6703 & 0.6558 & 0.6426 & 0.6733 \\
+%  & Lion-1 & 0.6865 & 0.7229 & 0.6970 & 0.6786 & 0.6638 & 0.6433 & 0.6750 \\
+%  & Lion-2 & 0.6772 & 0.7008 & 0.6792 & 0.6580 & 0.6405 & 0.6248 & 0.6661 \\
+%  & Lion-3 & 0.6674 & 0.6801 & 0.6571 & 0.6450 & 0.6310 & 0.6124 & 0.6573 \\
+% \midrule
+% \multirow{6}{*}{MiniLM} & AdamW-1 & 0.6857 & 0.7207 & 0.7008 & 0.6833 & 0.6698 & 0.6402 & 0.6778 \\
+%  & AdamW-2 & 0.6910 & 0.7345 & 0.7094 & 0.6908 & 0.6779 & 0.6471 & 0.6843 \\
+%  & AdamW-3 & 0.6910 & 0.7244 & 0.7127 & 0.6949 & 0.6749 & 0.6474 & 0.6834 \\
+%  & Lion-1 & 0.6841 & 0.7163 & 0.7031 & 0.6829 & 0.6667 & 0.6369 & 0.6752 \\
+%  & Lion-2 & 0.6790 & 0.7099 & 0.6916 & 0.6706 & 0.6541 & 0.6295 & 0.6692 \\
+%  & Lion-3 & 0.6770 & 0.6837 & 0.6808 & 0.6584 & 0.6479 & 0.6268 & 0.6667 \\
+% \midrule
+% \multirow{6}{*}{ModernBERT} & AdamW-1 & 0.6937 & 0.7171 & 0.7105 & 0.6936 & 0.6831 & 0.6555 & 0.6877 \\
+%  & AdamW-2 & 0.6869 & 0.7112 & 0.6839 & 0.6701 & 0.6589 & 0.6431 & 0.6795 \\
+%  & AdamW-3 & 0.6910 & 0.7156 & 0.6959 & 0.6729 & 0.6699 & 0.6497 & 0.6832 \\
+%  & Lion-1 & 0.6984 & 0.7285 & 0.7142 & 0.6991 & 0.6825 & 0.6610 & 0.6925 \\
+%  & Lion-2 & \textbf{0.7018} & \textbf{0.7381} & \textbf{0.7225} & \textbf{0.6978} & \textbf{0.6852} & \textbf{0.6638} & \textbf{0.6955} \\
+%  & Lion-3 & 0.6930 & 0.7165 & 0.7051 & 0.6797 & 0.6722 & 0.6542 & 0.6857 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 
@@
 \small
-\begin{tabular}{llccccccc}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{lllccccccccccc}
 \toprule
-\textbf{Model} & \textbf{Config} & \textbf{MAP} & \textbf{MAP@10} & \textbf{MAP@20} & \textbf{MAP@30} & \textbf{MAP@100} & \textbf{infAP} & \textbf{gm\_map} \\
+\textbf{Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{MAP} & \textbf{MAP@10} & \textbf{MAP@20} & \textbf{MAP@30} & \textbf{MAP@100} & \textbf{MAP@200} & \textbf{MAP@500} & \textbf{MAP@1000} & \textbf{infAP} & \textbf{gm\_map} \\
 \midrule
-\multirow{6}{*}{GTE} & AdamW-1 & 0.5005 & 0.1601 & 0.2499 & 0.2915 & 0.4226 & 0.5005 & 0.4143 \\
- & AdamW-2 & 0.4999 & 0.1632 & 0.2463 & 0.2904 & 0.4247 & 0.4999 & 0.4109 \\
- & AdamW-3 & 0.4899 & 0.1555 & 0.2350 & 0.2827 & 0.4136 & 0.4899 & 0.3995 \\
- & Lion-1 & 0.4794 & 0.1468 & 0.2274 & 0.2771 & 0.4039 & 0.4794 & 0.3878 \\
- & Lion-2 & 0.4577 & 0.1464 & 0.2201 & 0.2631 & 0.3810 & 0.4577 & 0.3623 \\
- & Lion-3 & 0.4521 & 0.1378 & 0.2145 & 0.2589 & 0.3747 & 0.4521 & 0.3525 \\
+\multirow{6}{*}{GTE} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.5005 & 0.1601 & 0.2499 & 0.2915 & 0.4226 & 0.4765 & 0.4964 & 0.5005 & 0.5005 & 0.4143 \\
+ & & 2 & 0.4999 & 0.1632 & 0.2463 & 0.2904 & 0.4247 & 0.4756 & 0.4957 & 0.4999 & 0.4999 & 0.4109 \\
+ & & 3 & 0.4899 & 0.1555 & 0.2350 & 0.2827 & 0.4136 & 0.4654 & 0.4855 & 0.4899 & 0.4899 & 0.3995 \\
+\cmidrule{2-13}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.4794 & 0.1468 & 0.2274 & 0.2771 & 0.4039 & 0.4538 & 0.4755 & 0.4794 & 0.4794 & 0.3878 \\
+ & & 2 & 0.4577 & 0.1464 & 0.2201 & 0.2631 & 0.3810 & 0.4294 & 0.4534 & 0.4577 & 0.4577 & 0.3623 \\
+ & & 3 & 0.4521 & 0.1378 & 0.2145 & 0.2589 & 0.3747 & 0.4240 & 0.4477 & 0.4521 & 0.4521 & 0.3525 \\
 \midrule
-\multirow{6}{*}{MiniLM} & AdamW-1 & 0.4814 & 0.1479 & 0.2272 & 0.2809 & 0.4062 & 0.4814 & 0.3920 \\
- & AdamW-2 & 0.4891 & 0.1531 & 0.2329 & 0.2882 & 0.4148 & 0.4891 & 0.3975 \\
- & AdamW-3 & 0.4908 & 0.1553 & 0.2383 & 0.2895 & 0.4159 & 0.4908 & 0.3972 \\
- & Lion-1 & 0.4858 & 0.1534 & 0.2344 & 0.2845 & 0.4087 & 0.4858 & 0.3919 \\
- & Lion-2 & 0.4755 & 0.1492 & 0.2254 & 0.2735 & 0.4006 & 0.4755 & 0.3823 \\
- & Lion-3 & 0.4706 & 0.1507 & 0.2207 & 0.2690 & 0.3940 & 0.4706 & 0.3736 \\
+\multirow{6}{*}{MiniLM} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.4814 & 0.1479 & 0.2272 & 0.2809 & 0.4062 & 0.4533 & 0.4775 & 0.4814 & 0.4814 & 0.3920 \\
+ & & 2 & 0.4891 & 0.1531 & 0.2329 & 0.2882 & 0.4148 & 0.4618 & 0.4854 & 0.4891 & 0.4891 & 0.3975 \\
+ & & 3 & 0.4908 & 0.1553 & 0.2383 & 0.2895 & 0.4159 & 0.463 & 0.4871 & 0.4908 & 0.4908 & 0.3972 \\
+\cmidrule{2-13}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.4858 & 0.1534 & 0.2344 & 0.2845 & 0.4087 & 0.4584 & 0.4821 & 0.4858 & 0.4858 & 0.3919 \\
+ & & 2 & 0.4755 & 0.1492 & 0.2254 & 0.2735 & 0.4006 & 0.4459 & 0.4717 & 0.4755 & 0.4755 & 0.3823 \\
+ & & 3 & 0.4706 & 0.1507 & 0.2207 & 0.2690 & 0.3940 & 0.4418 & 0.467 & 0.4706 & 0.4706 & 0.3736 \\
 \midrule
-\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.5066 & 0.1535 & 0.2389 & 0.2942 & 0.4274 & 0.5066 & 0.4045 \\
- & AdamW-2 & 0.4893 & 0.1493 & 0.2280 & 0.2803 & 0.4097 & 0.4893 & 0.3875 \\
- & AdamW-3 & 0.4971 & 0.1504 & 0.2304 & 0.2848 & 0.4192 & 0.4971 & 0.3988 \\
- & Lion-1 & 0.5121 & 0.1577 & 0.2448 & 0.2979 & 0.4347 & 0.5121 & 0.4162 \\
- & Lion-2 & \textbf{0.5115} & \textbf{0.1616} & \textbf{0.2417} & \textbf{0.2973} & \textbf{0.4333} & \textbf{0.5115} & \textbf{0.4157} \\
- & Lion-3 & 0.5020 & 0.1583 & 0.2333 & 0.2892 & 0.4239 & 0.5020 & 0.4071 \\
+\multirow{6}{*}{ModernBERT} & \multirow{3}{*}{AdamW} 
+ & 1 & 0.5066 & 0.1535 & 0.2389 & 0.2942 & 0.4274 & 0.482 & 0.5038 & 0.5066 & 0.5066 & 0.4045 \\
+ & & 2 & 0.4893 & 0.1493 & 0.2280 & 0.2803 & 0.4097 & 0.4647 & 0.4865 & 0.4893 & 0.4893 & 0.3875 \\
+ & & 3 & 0.4971 & 0.1504 & 0.2304 & 0.2848 & 0.4192 & 0.4717 & 0.4941 & 0.4971 & 0.4971 & 0.3988 \\
+\cmidrule{2-13}
+ & \multirow{3}{*}{Lion} 
+ & 1 & 0.5121 & 0.1577 & 0.2448 & 0.2979 & 0.4347 & 0.4883 & 0.5094 & 0.5121 & 0.5121 & 0.4162 \\
+ & & 2 & \textbf{0.5115} & \textbf{0.1616} & \textbf{0.2417} & \textbf{0.2973} & \textbf{0.4333} & 0.4866 & 0.509 & 0.5115 & \textbf{0.5115} & \textbf{0.4157} \\
+ & & 3 & 0.5020 & 0.1583 & 0.2333 & 0.2892 & 0.4239 & 0.4763 & 0.4993 & 0.502 & 0.5020 & 0.4071 \\
 \bottomrule
 \end{tabular}
+}
 \end{table*}
+
+% \begin{table*}[t]
+% \centering
+% \caption{MAP metrics at different cutoff levels and related metrics.}
+% \label{tab:map_metrics}
+% \small
+% \begin{tabular}{llccccccc}
+% \toprule
+% \textbf{Model} & \textbf{Config} & \textbf{MAP} & \textbf{MAP@10} & \textbf{MAP@20} & \textbf{MAP@30} & \textbf{MAP@100} & \textbf{infAP} & \textbf{gm\_map} \\
+% \midrule
+% \multirow{6}{*}{GTE} & AdamW-1 & 0.5005 & 0.1601 & 0.2499 & 0.2915 & 0.4226 & 0.5005 & 0.4143 \\
+%  & AdamW-2 & 0.4999 & 0.1632 & 0.2463 & 0.2904 & 0.4247 & 0.4999 & 0.4109 \\
+%  & AdamW-3 & 0.4899 & 0.1555 & 0.2350 & 0.2827 & 0.4136 & 0.4899 & 0.3995 \\
+%  & Lion-1 & 0.4794 & 0.1468 & 0.2274 & 0.2771 & 0.4039 & 0.4794 & 0.3878 \\
+%  & Lion-2 & 0.4577 & 0.1464 & 0.2201 & 0.2631 & 0.3810 & 0.4577 & 0.3623 \\
+%  & Lion-3 & 0.4521 & 0.1378 & 0.2145 & 0.2589 & 0.3747 & 0.4521 & 0.3525 \\
+% \midrule
+% \multirow{6}{*}{MiniLM} & AdamW-1 & 0.4814 & 0.1479 & 0.2272 & 0.2809 & 0.4062 & 0.4814 & 0.3920 \\
+%  & AdamW-2 & 0.4891 & 0.1531 & 0.2329 & 0.2882 & 0.4148 & 0.4891 & 0.3975 \\
+%  & AdamW-3 & 0.4908 & 0.1553 & 0.2383 & 0.2895 & 0.4159 & 0.4908 & 0.3972 \\
+%  & Lion-1 & 0.4858 & 0.1534 & 0.2344 & 0.2845 & 0.4087 & 0.4858 & 0.3919 \\
+%  & Lion-2 & 0.4755 & 0.1492 & 0.2254 & 0.2735 & 0.4006 & 0.4755 & 0.3823 \\
+%  & Lion-3 & 0.4706 & 0.1507 & 0.2207 & 0.2690 & 0.3940 & 0.4706 & 0.3736 \\
+% \midrule
+% \multirow{6}{*}{ModernBERT} & AdamW-1 & 0.5066 & 0.1535 & 0.2389 & 0.2942 & 0.4274 & 0.5066 & 0.4045 \\
+%  & AdamW-2 & 0.4893 & 0.1493 & 0.2280 & 0.2803 & 0.4097 & 0.4893 & 0.3875 \\
+%  & AdamW-3 & 0.4971 & 0.1504 & 0.2304 & 0.2848 & 0.4192 & 0.4971 & 0.3988 \\
+%  & Lion-1 & 0.5121 & 0.1577 & 0.2448 & 0.2979 & 0.4347 & 0.5121 & 0.4162 \\
+%  & Lion-2 & \textbf{0.5115} & \textbf{0.1616} & \textbf{0.2417} & \textbf{0.2973} & \textbf{0.4333} & \textbf{0.5115} & \textbf{0.4157} \\
+%  & Lion-3 & 0.5020 & 0.1583 & 0.2333 & 0.2892 & 0.4239 & 0.5020 & 0.4071 \\
+% \bottomrule
+% \end{tabular}
+% \end{table*}
 

### 2025-04-27 cross_encoder/trec_dl_19_eval_2.py
Enhance TREC evaluation by adding detailed result parsing and output for metrics

@@ -84,14 +84,7 @@ def run_trec_eval(qrels_path, run_path, metrics):
     cmd = ["./trec_eval/trec_eval", "-m","all_trec", qrels_path, run_path]
     
     result = subprocess.run(cmd, capture_output=True, text=True, check=True)
-    print("Result of trec_eval:")
-    print("*************************************************************")
-    print(result.stdout)
-    print("*************************************************************")
-    print("Error of trec_eval:")
-    print("*************************************************************")
-    print(result.stderr)
-    print("*************************************************************")    
+
     # Parse results
     lines = result.stdout.strip().split('\n')
     results = {}
@@ -118,7 +111,45 @@ def run_trec_eval(qrels_path, run_path, metrics):
             # Keep non-numeric values as strings
             results[metric] = value
     
-    return results, None
+    # run again to get the detailed results
+    cmd = ["./trec_eval/trec_eval", "-m","all_trec", "-q", qrels_path, run_path]
+    
+    result_detailed = subprocess.run(cmd, capture_output=True, text=True, check=True)
+    print("Result of trec_eval:")
+    print("*************************************************************")
+    print(result_detailed.stdout)
+    print("*************************************************************")
+    print("Error of trec_eval:")
+    print("*************************************************************")
+    print(result_detailed.stderr)
+    print("*************************************************************")    
+    # Parse results
+    lines = result_detailed.stdout.strip().split('\n')
+    detailed_results = {}
+    
+    for line in lines:
+        line = line.strip()
+        if not line:
+            continue
+            
+        parts = line.split()
+        if len(parts) != 3:
+            continue
+            
+        metric, qid, value = parts
+        
+        # Only keep the 'all' summary values
+        if qid == 'all':
+            continue
+            
+        try:
+            value = float(value)
+            detailed_results[metric] = value
+        except ValueError:
+            # Keep non-numeric values as strings
+            detailed_results[metric] = value
+
+    return results, detailed_results
 
 def main():
     args = parse_args()

### 2025-06-11 cross_encoder/thesis/Appendices/AppendixA.tex
added thesis (new format)

\chapter{Code} % Main appendix title

\label{AppendixA} 
\begin{sloppypar}
\lstset{
  language=Python,
  backgroundcolor=\color{lightgray},
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  commentstyle=\color{green},
  keywordstyle=\color{blue},
  stringstyle=\color{red},
}

\subsection{Image Degradation Code in Python}

\section*{Code for Degradation Simulation}

\begin{lstlisting}[language=Python]
import os
import cv2
import numpy as np
import random
from typing import Tuple
from glob import glob

class DegradationModels:
    def __init__(self, target_size: Tuple[int, int] = (512, 512)):
        self.target_size = target_size
        self.ffhq_path = "inputs/originals"
        self.output_path = "inputs/degraded"
        os.makedirs(self.output_path, exist_ok=True)

    def _resize_image(self, image: np.ndarray) -> np.ndarray:
        return cv2.resize(image, self.target_size)

    def _apply_gaussian_blur(self, image: np.ndarray, kernel_size: int = 0, sigma: float = 0) -> np.ndarray:
        if kernel_size == 0:
            kernel_size = int(2 * np.ceil(2 * sigma) + 1)
        if kernel_size % 2 == 0:
            kernel_size += 1
        return cv2.GaussianBlur(image, (kernel_size, kernel_size), sigma)

    def _add_gaussian_noise(self, image: np.ndarray, std: float) -> np.ndarray:
        noise = np.random.normal(0, std, image.shape).astype(np.float32)
        noisy_image = image.astype(np.float32) + noise
        return np.clip(noisy_image, 0, 255).astype(np.uint8)

    def _jpeg_compression(self, image: np.ndarray, quality: int) -> np.ndarray:
        encode_param = [int(cv2.IMWRITE_JPEG_QUALITY), quality]
        _, encoded_img = cv2.imencode('.jpg', image, encode_param)
        return cv2.imdecode(encoded_img, cv2.IMREAD_COLOR)

    def apply_degradation(self, image: np.ndarray, mode: str = 'BicC', scale_factor: int = 8) -> np.ndarray:
        image = self._resize_image(image)
        h, w = image.shape[:2]

        if mode == 'BicC':
            downsampled = cv2.resize(image, (w//scale_factor, h//scale_factor), interpolation=cv2.INTER_CUBIC)
            upsampled = cv2.resize(downsampled, (w, h), interpolation=cv2.INTER_CUBIC)
            quality = random.randint(90, 95)
            return self._jpeg_compression(upsampled, quality)

        elif mode == 'BicN':
            downsampled = cv2.resize(image, (w//scale_factor, h//scale_factor), interpolation=cv2.INTER_CUBIC)
            upsampled = cv2.resize(downsampled, (w, h), interpolation=cv2.INTER_CUBIC)
            noise_std = random.uniform(20, 25)
            return self._add_gaussian_noise(upsampled, noise_std)

        elif mode == 'BBilN':
            sigma = random.uniform(5, 7)
            blurred = self._apply_gaussian_blur(image, sigma=sigma)
            downsampled = cv2.resize(blurred, (w//scale_factor, h//scale_factor), interpolation=cv2.INTER_LINEAR)
            upsampled = cv2.resize(downsampled, (w, h), interpolation=cv2.INTER_LINEAR)
            noise_std = random.uniform(10, 15)
            return self._add_gaussian_noise(upsampled, noise_std)

        else:
            raise ValueError(f"Unknown degradation mode: {mode}")

    def process_ffhq_images(self):
        modes = ['BicC', 'BicN', 'BBilN']
        image_paths = glob(os.path.join(self.ffhq_path, '*'))

        for mode in modes:
            output_dir = os.path.join(self.output_path, mode)
            os.makedirs(output_dir, exist_ok=True)

            print(f"Processing mode: {mode}")
            for img_path in image_paths:
                img = cv2.imread(img_path)
                if img is None:
                    print(f"Skipping {img_path}, could not read.")
                    continue
                degraded = self.apply_degradation(img, mode=mode)
                filename = os.path.basename(img_path)
                cv2.imwrite(os.path.join(output_dir, filename), degraded)

if __name__ == "__main__":
    degrader = DegradationModels()
    degrader.process_ffhq_images()
\end{lstlisting}

\subsection*{Code for colour degradation}

\begin{lstlisting}[language=Python]
from PIL import Image
import os
from glob import glob

# Settings
input_folder = "inputs/originals"
output_folder = "inputs/degraded_mediancut"
num_colors = 6  # Number of colors for Median Cut

# Create output directory if it doesn't exist
os.makedirs(output_folder, exist_ok=True)

# Get all image paths from ffhq folder
image_paths = glob(os.path.join(input_folder, "*"))

# Process each image
for path in image_paths:
    try:
        # Load image
        original = Image.open(path).convert("RGB")

        # Apply Median Cut quantization
        degraded = original.quantize(colors=num_colors, method=0)

        # Convert back to RGB and save
        degraded_rgb = degraded.convert("RGB")
        filename = os.path.basename(path)
        degraded_rgb.save(os.path.join(output_folder, filename))

        print(f"Processed and saved: {filename}")
    except Exception as e:
        print(f"Failed to process {path}: {e}")
\end{lstlisting}
\begin{lstlisting}[language=Python]
import cv2
import numpy as np
from sklearn.cluster import KMeans
import os
from tqdm import tqdm

def degrade_color_palette(image, num_colors=8):
    """Degrade the color palette of a single image."""
    h, w, _ = image.shape
    img_reshaped = image.reshape((-1, 3))

    # Apply k-means to extract dominant colors (palette)
    kmeans = KMeans(n_clusters=num_colors, random_state=42).fit(img_reshaped)
    palette = np.uint8(kmeans.cluster_centers_)
    labels = kmeans.predict(img_reshaped)

    # Reconstruct degraded image from palette
    degraded = palette[labels].reshape((h, w, 3))
    return degraded

def process_folder(input_folder, output_folder, num_colors=8):
    """
    Process all images in a folder and save degraded versions in an output folder.

    Args:
        input_folder: Path to the folder containing input images
        output_folder: Path to save degraded images
        num_colors: Number of colors to use in the degraded palette
    """
    # Create output folder if it doesn't exist
    os.makedirs(output_folder, exist_ok=True)

    # Get all image files in the input folder
    image_files = [f for f in os.listdir(input_folder) if f.lower().endswith(('.png', '.jpg', '.jpeg'))]

    # Process each image
    for filename in tqdm(image_files, desc="Processing images"):
        # Load image
        img_path = os.path.join(input_folder, filename)
        img = cv2.imread(img_path)
        img = cv2.cvtColor(img, cv2.COLOR_BGR2RGB)

        # Apply color degradation
        degraded = degrade_color_palette(img, num_colors=num_colors)

        # Save degraded image
        output_path = os.path.join(output_folder, filename)
        cv2.imwrite(output_path, cv2.cvtColor(degraded, cv2.COLOR_RGB2BGR))

# Example usage
input_folder = "inputs/originals"  # Folder containing original images
output_folder = "inputs/degraded_kmeans"  # Folder to save degraded images
num_colors = 6  # Number of colors in the degraded palette

process_folder(input_folder, output_folder, num_colors=num_colors)
print(f"Degraded images saved to {output_folder}")

\end{lstlisting}
\end{sloppypar}

### 2025-06-11 cross_encoder/thesis/Chapters/1.introduction.tex
added thesis (new format)

% Chapter 1

\chapter{Introduction} % Main chapter title

\label{Chapter1} % For referencing the chapter elsewhere, use \ref{Chapter1}

%----------------------------------------------------------------------------------------

% Define some commands to keep the formatting separated from the content 
\newcommand{\keyword}[1]{\textbf{#1}}
\newcommand{\tabhead}[1]{\textbf{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\file}[1]{\texttt{\bfseries#1}}
\newcommand{\option}[1]{\texttt{\itshape#1}}

Information Retrieval (IR) systems enable users to find relevant information from vast document collections. Modern IR systems employ a two-stage pipeline: efficient initial retrieval (BM25 \cite{robertson2009probabilistic} or dense vector retrieval \cite{karpukhin2020dense}) followed by sophisticated reranking for precision optimization.

Cross-encoder models represent the state-of-the-art for reranking \cite{nogueira2020passagererankingbert,Nogueira2020Document} due to their ability to model deep query-document interactions. Unlike bi-encoders that process queries and documents independently, cross-encoders simultaneously analyze query-document pairs, enabling rich token-level interactions for accurate relevance estimation.

The effectiveness of cross-encoder models depends heavily on optimization strategies during fine-tuning. While AdamW \cite{loshchilov2019decoupled} has been the standard choice for transformer training, recent developments in optimization algorithms, particularly the Lion optimizer \cite{chen2023symbolic}, have shown promising results across various domains, motivating investigation into their applicability for information retrieval tasks.

%----------------------------------------------------------------------------------------

\section{Cross-Encoder Models for Information Retrieval}

Cross-encoders process query-document pairs by concatenating them with special tokens ([CLS] query [SEP] document [SEP]) and feeding the combined sequence through a transformer architecture \cite{devlin2019bert}. The [CLS] token representation predicts relevance scores through a linear layer with sigmoid activation.

Modern implementations leverage various transformer architectures with distinct characteristics:
\begin{itemize}
\item \textbf{MiniLM-L12-H384}: Distilled BERT variant optimized for efficiency \cite{wang2020minilm} (512 tokens context)
\item \textbf{GTE-multilingual-base}: Extended context model (8192 tokens) with multilingual training \cite{li2023towards}
\item \textbf{ModernBERT-base}: State-of-the-art architecture with RoPE \cite{su2023roformerenhancedtransformerrotary}, Flash Attention \cite{dao2022flashattentionfastmemoryefficientexact}, and GeGLU \cite{shazeer2020gluvariantsimprovetransformer} \cite{modernbert}
\end{itemize}

\begin{figure}[!htp]
    \centering
    \includegraphics[width=0.8\textwidth, height=5cm, keepaspectratio]{Figures/cross_encoder_archi.png}
    \caption{Cross-Encoder Architecture for Query-Document Reranking.}
    \label{fig:Cross-Encoder-Architecture}
\end{figure}

\section{Optimization Strategies}

Cross-encoder effectiveness depends on optimization strategies during fine-tuning. While AdamW \cite{loshchilov2019decoupled} has been the standard choice, the Lion optimizer \cite{chen2023symbolic} presents a novel approach using sign-based momentum: \texttt{update = sign(momentum) * learning\_rate}. This simplification offers potential memory efficiency and different convergence characteristics compared to traditional adaptive methods.

\section{Research Objectives}

This research conducts a comprehensive comparative analysis of Lion and AdamW optimizers for cross-encoder models in information retrieval. Our primary objectives are:

\begin{enumerate}
    \item \textbf{Systematic Evaluation}: Compare Lion and AdamW across three transformer architectures (MiniLM, GTE, ModernBERT)
    \item \textbf{Comprehensive Benchmarking}: Evaluate performance using standard IR metrics on TREC DL 2019 and MS MARCO
    \item \textbf{Training Dynamics Analysis}: Investigate convergence patterns and stability across training epochs
    \item \textbf{Practical Guidelines}: Provide evidence-based recommendations for optimizer selection
\end{enumerate}

\subsection{Expected Contributions}
\begin{itemize}
\item Empirical evidence comparing Lion and AdamW optimizers across multiple architectures
\item Optimization guidelines for cross-encoder training in information retrieval
\item Analysis of training efficiency and convergence characteristics
\end{itemize}

\section{Methodology Overview}

We employ systematic experimentation across three transformer architectures representing different performance-efficiency trade-offs: MiniLM (efficient baseline), GTE (extended context), and ModernBERT (state-of-the-art). Experiments utilize standardized datasets (MS MARCO \cite{DBLP:journals/corr/NguyenRSGTMD16} for training, TREC DL 2019 \cite{craswell2020overview} for evaluation) with cloud infrastructure (Modal platform \cite{modal_labs}, NVIDIA A100-80GB GPUs) ensuring reproducible results. Comprehensive tracking through Weights \& Biases \cite{wandb2020} enables detailed training dynamics analysis.

\section{Thesis Organization}

This thesis systematically presents research methodology, experimental results, and implications:
\textbf{Chapter 2}: Literature review on cross-encoder architectures and optimization algorithms
\textbf{Chapter 3}: Research gaps and proposed comparative evaluation approach  
\textbf{Chapter 4}: Experimental methodology and implementation details
\textbf{Chapter 5}: Comprehensive results with training dynamics and statistical analysis
\textbf{Chapter 6}: Key findings, implications, and future research directions

The thesis contributes empirical insights into optimizer selection for cross-encoder architectures with implications for search engines, question-answering systems, and information retrieval applications where ranking quality is paramount.

### 2025-06-11 cross_encoder/thesis/Chapters/2.literature.tex
added thesis (new format)

\chapter{Literature Review}
\label{Chapter2}

This chapter reviews key literature on cross-encoder models, optimization algorithms, and evaluation methodologies for information retrieval.

\section{Neural Information Retrieval Evolution}

Information retrieval has transformed from traditional term-based matching (TF-IDF, BM25 \cite{robertson2009probabilistic}) to neural approaches that capture semantic relationships. Dense retrieval methods like DPR \cite{karpukhin2020dense} and ANCE \cite{xiong2020approximate} encode queries and documents independently, while cross-encoders enable richer query-document interactions.

\section{Cross-Encoder Architectures}

Cross-encoders leverage transformer architectures for document reranking. Nogueira and Cho \cite{nogueira2020passagererankingbert} demonstrated BERT's effectiveness for reranking by concatenating queries and documents with special tokens. The [CLS] representation predicts relevance scores through fine-tuning on labeled data.

Modern architectures include:
\begin{itemize}
\item \textbf{MiniLM \cite{wang2020minilm}}: Distilled BERT variant for efficiency
\item \textbf{GTE \cite{li2023towards}}: Multilingual model with extended context (8192 tokens)  
\item \textbf{ModernBERT \cite{modernbert}}: Advanced architecture with RoPE, Flash Attention, GeGLU
\end{itemize}

\section{Optimization Algorithms}

\subsection{Traditional Optimizers}
Adam \cite{kingma2017adam} combines momentum with adaptive learning rates using first and second moment estimates. AdamW \cite{loshchilov2019decoupled} improved generalization by decoupling weight decay from gradient updates, becoming the standard for transformer training.

\subsection{Lion Optimizer}
Lion \cite{chen2023symbolic} represents a paradigm shift using sign-based momentum: \texttt{update = sign(momentum) * learning\_rate}. This simplified approach reduces memory requirements while maintaining competitive performance across vision tasks.

\section{Evaluation Benchmarks}

Standard IR evaluation relies on:
\begin{itemize}
\item \textbf{MS MARCO \cite{DBLP:journals/corr/NguyenRSGTMD16}}: Large-scale passage ranking dataset
\item \textbf{TREC DL 2019 \cite{craswell2020overview}}: Deep learning track for passage ranking
\item \textbf{Metrics}: nDCG@10, MRR@10, MAP measure ranking quality
\end{itemize}

\section{Research Gaps}

While cross-encoder effectiveness is well-established, systematic investigation of optimization strategies remains limited. Most studies assume AdamW without exploring alternatives like Lion, particularly for information retrieval tasks where ranking precision is critical.

\subsection{Modern Transformer Architectures}

Recent developments in transformer architectures have introduced models specifically designed for improved efficiency and longer context processing. The General Text Embeddings (GTE) family \cite{li2023towards} demonstrated that models trained on diverse, multilingual corpora could achieve strong performance across various text understanding tasks, including retrieval and reranking.

ModernBERT \cite{modernbert} represents the current state-of-the-art in encoder-only transformers, incorporating several architectural innovations:

\begin{itemize}
    \item \textbf{Rotary Positional Embeddings (RoPE)} \cite{su2023roformerenhancedtransformerrotary}: Enable effective handling of longer sequences by providing more flexible positional encoding.
    \item \textbf{Flash Attention} \cite{dao2022flashattentionfastmemoryefficientexact}: Improves memory efficiency and computational speed for attention operations.
    \item \textbf{GeGLU Activation Functions} \cite{shazeer2020gluvariantsimprovetransformer}: Enhance model expressiveness while maintaining computational efficiency.
\end{itemize}

These architectural improvements enable ModernBERT to process sequences up to 8192 tokens efficiently, making it particularly suitable for long document processing tasks.

%----------------------------------------------------------------------------------------
\section{Optimization Algorithms for Deep Learning}

The success of neural information retrieval models depends heavily on the optimization algorithms used during training. The choice of optimizer affects convergence speed, final performance, and training stability. This section reviews the evolution of optimization algorithms and their specific applications to transformer-based models.

\subsection{Classical Optimization Methods}

Stochastic Gradient Descent (SGD) remains a fundamental optimization algorithm, providing theoretical guarantees and interpretable behavior. However, the challenges of training deep neural networks - including vanishing gradients, saddle points, and varying curvature across parameter space - motivated the development of adaptive optimization methods.

The introduction of momentum to SGD addressed some limitations by accumulating gradients across iterations, helping optimization navigate through areas of high curvature and accelerate convergence in consistent directions. However, SGD with momentum still struggled with the varying scales of different parameters and the need for careful learning rate tuning.

\subsection{Adaptive Learning Rate Methods}

AdaGrad \cite{duchi2011adaptive} introduced the concept of adaptive learning rates, automatically adjusting step sizes based on historical gradient information. By maintaining per-parameter learning rates inversely proportional to the square root of accumulated squared gradients, AdaGrad could handle sparse features effectively and reduce the need for manual learning rate tuning.

RMSprop \cite{tieleman2012lecture} addressed AdaGrad's aggressive learning rate decay by using exponential moving averages instead of accumulating all historical gradients. This modification prevented the learning rate from becoming too small too quickly, enabling continued learning throughout training.

Adam \cite{kingma2017adam} combined the benefits of momentum-based methods with adaptive learning rates, maintaining separate exponentially decaying averages for both gradients (first moment) and squared gradients (second moment). The Adam optimizer became widely adopted due to its robustness across different architectures and tasks, requiring minimal hyperparameter tuning in many scenarios.

\subsection{AdamW and Weight Decay Regularization}

AdamW \cite{loshchilov2019decoupled} represented a significant improvement over Adam by decoupling weight decay regularization from the adaptive learning rate mechanism. The key insight was that traditional L2 regularization in Adam led to suboptimal behavior because the adaptive learning rate scaling affected both gradient and regularization terms.

By separating weight decay from gradient-based updates, AdamW achieved better generalization performance and more stable training dynamics. The decoupled formulation:

\begin{equation}
\theta_{t+1} = \theta_t - \eta \left( \frac{\hat{m}_t}{\sqrt{\hat{v}_t} + \epsilon} + \lambda \theta_t \right)
\end{equation}

where $\lambda$ represents the weight decay coefficient applied directly to parameters, independent of the adaptive learning rate scaling.

AdamW quickly became the standard optimizer for transformer-based models, demonstrating superior performance across various natural language processing tasks and establishing itself as the default choice for most deep learning applications.

\subsection{Lion Optimizer: A Novel Approach}

The Lion (EvoLved Sign mOmeNtum) optimizer \cite{chen2023symbolic} represents a departure from traditional adaptive optimization approaches. Developed through symbolic mathematics and program search, Lion utilizes a simplified update rule that relies primarily on the sign of momentum-based gradients.

The Lion update mechanism follows:

\begin{align}
c_t &= \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
\theta_t &= \theta_{t-1} - \eta \left( \text{sign}(c_t) + \lambda \theta_{t-1} \right) \\
m_t &= \beta_2 m_{t-1} + (1 - \beta_2) g_t
\end{align}

### 2025-06-11 cross_encoder/thesis/Chapters/3.research_gap_and_proposed_solution.tex
added thesis (new format)

% Chapter Template

\chapter{Research Gap \& Methodology} % Main chapter title

\label{Chapter3} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{Chapter3}

\section{Research Gap}

While AdamW has become standard for transformer training, the Lion optimizer's effectiveness for cross-encoder IR tasks remains unexplored. Key gaps include:

\begin{itemize}
    \item \textbf{Novel Optimizer Evaluation}: Lion optimizer's memory efficiency and simplified updates haven't been systematically evaluated for cross-encoder training in IR contexts.
    \item \textbf{Architecture-Optimizer Interaction}: Different transformer architectures (MiniLM, GTE, ModernBERT) may respond differently to optimization strategies.
    \item \textbf{Training Efficiency Trade-offs}: Lion's claimed efficiency advantages need validation for computationally intensive cross-encoder training.
    \item \textbf{Comprehensive IR Benchmarking}: Systematic evaluation across standard IR datasets (MS MARCO, TREC DL) with multiple metrics is needed.
\end{itemize}

\section{Methodology}

This research systematically compares Lion and AdamW optimizers across three cross-encoder architectures for information retrieval tasks.

\subsection{Model Selection}
Three transformer architectures representing different efficiency-effectiveness points:
\begin{itemize}
    \item \textbf{MiniLM-L12-H384}: Compact model (384 hidden dims, 12 layers) for efficiency
    \item \textbf{GTE-multilingual-base}: Multilingual model with 8192 token context support
    \item \textbf{ModernBERT-base}: State-of-the-art encoder with RoPE, Flash Attention, GeGLU
\end{itemize}

\subsection{Experimental Design}
\textbf{Training Setup}: Modal cloud platform with NVIDIA A100-80GB GPUs, batch size 16, 3 epochs, MS MARCO passage dataset.

\textbf{Optimizer Configurations}:
\begin{itemize}
    \item \textbf{AdamW}: Learning rates 2e-5 (MiniLM, GTE), 2e-6 (ModernBERT), weight decay 0.01
    \item \textbf{Lion}: Same learning rates, β₁=0.9, β₂=0.99, sign-based momentum updates
\end{itemize}

\textbf{Evaluation}: TREC 2019 DL Track and MS MARCO dev set using NDCG@10, MAP, MRR@10 metrics.

%----------------------------------------------------------------------------------------

### 2025-06-11 cross_encoder/thesis/Chapters/4.methods.tex
added thesis (new format)

% Chapter Template

\chapter{Cross-Encoder Architecture and Training Framework} % Main chapter title

\label{Chapter4} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{Chapter4}

%----------------------------------------------------------------------------------------

\section{Cross-Encoder Architecture for Information Retrieval}

\subsection{Architectural Overview}

Cross-encoder models represent a significant advancement in neural information retrieval, enabling deep interaction modeling between queries and documents. Unlike bi-encoder approaches that process queries and documents independently, cross-encoders perform joint encoding, allowing for rich token-level interactions that lead to superior relevance estimation.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{Figures/cross_encoder_archi.png}
    \captionsetup{font=footnotesize, width=0.9\linewidth}
    \caption{Cross-Encoder Architecture for Passage Reranking. The model processes concatenated query-passage pairs through transformer layers, producing relevance scores for reranking candidate documents.}
    \label{fig:cross_encoder_architecture}
\end{figure}

The cross-encoder architecture follows a structured approach to relevance modeling:

\begin{enumerate}
    \item \textbf{Input Formatting}: Query and passage text are concatenated with special tokens: [CLS] query [SEP] passage [SEP]
    \item \textbf{Tokenization}: The input sequence is tokenized using subword tokenization (typically BERT-style WordPiece)
    \item \textbf{Transformer Encoding}: The concatenated sequence passes through multiple transformer layers with self-attention mechanisms
    \item \textbf{Relevance Prediction}: The [CLS] token representation is fed to a classification head producing a relevance score
    \item \textbf{Score Normalization}: Sigmoid activation ensures scores fall within [0,1] range for ranking
\end{enumerate}

\subsection{Model Architecture Specifications}

Three distinct transformer architectures were selected to represent different efficiency-effectiveness trade-offs:

\subsubsection{MiniLM-L12-H384-uncased}

MiniLM \cite{wang2020minilm} represents an efficient architecture designed through knowledge distillation:

\textbf{Architectural Parameters}:
\begin{itemize}
    \item Hidden dimensions: 384
    \item Number of layers: 12
    \item Attention heads: 12
    \item Maximum sequence length: 512 tokens
    \item Parameter count: ~33M parameters
    \item Vocabulary size: 30,522 tokens
\end{itemize}

\textbf{Design Philosophy}: MiniLM achieves efficiency through knowledge distillation from larger teacher models while maintaining competitive performance. The reduced hidden dimensions and moderate layer count make it suitable for resource-constrained scenarios while preserving essential language understanding capabilities.

\subsubsection{GTE-multilingual-base}

The General Text Embeddings (GTE) model \cite{li2023towards} extends capabilities to multilingual contexts with enhanced capacity:

\textbf{Architectural Parameters}:
\begin{itemize}
    \item Hidden dimensions: 768
    \item Number of layers: 12
    \item Attention heads: 12
    \item Maximum sequence length: 8192 tokens
    \item Parameter count: ~137M parameters
    \item Multilingual vocabulary support
\end{itemize}

\textbf{Design Philosophy}: GTE models are trained on diverse, multilingual corpora with multi-stage contrastive learning. The extended context length (8192 tokens) enables processing of longer documents, making it particularly suitable for comprehensive passage analysis and cross-lingual retrieval scenarios.

\subsubsection{ModernBERT-base}

ModernBERT \cite{modernbert} incorporates state-of-the-art architectural innovations for improved efficiency and effectiveness:

\textbf{Architectural Parameters}:
\begin{itemize}
    \item Hidden dimensions: 768
    \item Number of layers: 22
    \item Attention heads: 12
    \item Maximum sequence length: 8192 tokens
    \item Parameter count: ~139M parameters
    \item Advanced positional encoding and attention mechanisms
\end{itemize}

\textbf{Architectural Innovations}:
\begin{enumerate}
    \item \textbf{Rotary Position Embedding (RoPE)} \cite{su2023roformerenhancedtransformerrotary}: Enables effective handling of longer sequences through relative position encoding
    \item \textbf{Flash Attention} \cite{dao2022flashattentionfastmemoryefficientexact}: Improves memory efficiency and computational speed for attention operations
    \item \textbf{GeGLU Activation Functions} \cite{shazeer2020gluvariantsimprovetransformer}: Enhances model expressiveness while maintaining computational efficiency
\end{enumerate}

\textbf{Design Philosophy}: ModernBERT represents the current state-of-the-art in encoder-only transformers, combining architectural efficiency improvements with enhanced sequence processing capabilities. The deeper architecture (22 layers) with efficient attention mechanisms enables superior language understanding while maintaining practical computational requirements.

%----------------------------------------------------------------------------------------

\section{Optimizer Analysis and Implementation}

\subsection{AdamW Optimizer}

AdamW \cite{loshchilov2019decoupled} has established itself as the standard optimizer for transformer-based models through improved handling of weight decay regularization.

\subsubsection{Mathematical Formulation}

The AdamW update rule decouples weight decay from gradient-based updates:

\begin{align}
m_t &= \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
v_t &= \beta_2 v_{t-1} + (1 - \beta_2) g_t^2 \\
\hat{m}_t &= \frac{m_t}{1 - \beta_1^t} \\
\hat{v}_t &= \frac{v_t}{1 - \beta_2^t} \\
\theta_t &= \theta_{t-1} - \eta \left( \frac{\hat{m}_t}{\sqrt{\hat{v}_t} + \epsilon} + \lambda \theta_{t-1} \right)
\end{align}

where:
\begin{itemize}
    \item $g_t$: gradient at time step $t$
    \item $m_t$: exponential moving average of gradients (momentum)
    \item $v_t$: exponential moving average of squared gradients
    \item $\eta$: learning rate
    \item $\lambda$: weight decay coefficient
    \item $\beta_1, \beta_2$: exponential decay rates for moment estimates
\end{itemize}

\subsubsection{Key Advantages}

\begin{enumerate}
    \item \textbf{Decoupled Weight Decay}: Separates regularization from adaptive learning rate scaling
    \item \textbf{Adaptive Learning Rates}: Per-parameter learning rate adaptation based on gradient history
    \item \textbf{Momentum Integration}: Incorporates momentum for improved convergence in consistent directions
    \item \textbf{Robustness}: Proven effectiveness across diverse transformer architectures and tasks
\end{enumerate}

\subsection{Lion Optimizer}

The Lion (Evolved Sign Momentum) optimizer \cite{chen2023symbolic} represents a novel approach discovered through symbolic mathematics and program search.

\subsubsection{Mathematical Formulation}

Lion utilizes a simplified update mechanism based on momentum and sign operations:

\begin{align}
c_t &= \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
\theta_t &= \theta_{t-1} - \eta \left( \text{sign}(c_t) + \lambda \theta_{t-1} \right) \\
m_t &= \beta_2 m_{t-1} + (1 - \beta_2) g_t
\end{align}

where:
\begin{itemize}
    \item $c_t$: interpolated momentum for update direction
    \item $\text{sign}(\cdot)$: element-wise sign function
    \item Other parameters follow similar definitions as AdamW
\end{itemize}

\subsubsection{Key Innovations}

\begin{enumerate}
    \item \textbf{Sign-based Updates}: Uses only gradient direction, not magnitude, for parameter updates
    \item \textbf{Memory Efficiency}: Maintains only first-moment estimates, reducing memory requirements by ~50\%
    \item \textbf{Computational Simplicity}: Sign operation is computationally cheaper than square root calculations
    \item \textbf{Robust Performance}: Demonstrated effectiveness across computer vision and vision-language tasks
\end{enumerate}

\subsubsection{Theoretical Advantages}

The sign-based update mechanism provides several theoretical benefits:

\begin{itemize}
    \item \textbf{Noise Resistance}: Sign operation filters out gradient noise while preserving direction information
    \item \textbf{Scale Invariance}: Updates are independent of gradient magnitude, providing consistent step sizes
    \item \textbf{Convergence Properties}: Simplified dynamics may lead to more stable convergence in some scenarios
\end{itemize}

%----------------------------------------------------------------------------------------

\section{Training Framework and Implementation}

\subsection{Data Preprocessing and Input Formatting}

\subsubsection{Dataset Preparation}

The MS MARCO Passage Ranking dataset \cite{DBLP:journals/corr/NguyenRSGTMD16} serves as the primary training corpus:

\begin{itemize}
    \item \textbf{Training Queries}: 502,939 queries with relevance labels
    \item \textbf{Passage Collection}: 8.8M unique passages
    \item \textbf{Relevance Judgments}: Binary labels indicating passage relevance to queries
    \item \textbf{Data Source}: Real user queries from Bing search engine with human-annotated relevance
\end{itemize}

\subsubsection{Input Processing Pipeline}

\textbf{Text Preprocessing}:
\begin{enumerate}
    \item Unicode normalization and cleaning
    \item Whitespace normalization
    \item Special character handling for robustness
\end{enumerate}

\textbf{Sequence Construction}:
\begin{enumerate}
    \item Query-passage concatenation: [CLS] query [SEP] passage [SEP]
    \item Tokenization using model-specific tokenizers
    \item Sequence length truncation based on model capacity
    \item Attention mask creation for proper sequence processing
\end{enumerate}

\textbf{Batch Construction}:
\begin{enumerate}
    \item Dynamic padding to maximum sequence length in batch
    \item Label tensor creation for binary classification
    \item Efficient DataLoader implementation with multi-processing
\end{enumerate}

\subsection{Training Configuration}

\subsubsection{Hyperparameter Settings}

\textbf{Model-Specific Configurations}:

\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Parameter} & \textbf{MiniLM} & \textbf{GTE} & \textbf{ModernBERT} \\
\hline
Learning Rate & 2e-5 & 2e-5 & 2e-6 \\
\hline
Max Sequence Length & 512 & 8192 & 8192 \\
\hline
Batch Size & 16 & 16 & 16 \\
\hline
Weight Decay & 0.01 & 0.01 & 0.01 \\
\hline
Warmup Steps & 1000 & 1000 & 1000 \\
\hline
LR Scheduler & None & None & Cosine Annealing \\
\hline
Training Epochs & 3 & 3 & 3 \\
\hline
\end{tabular}
\caption{Model-specific training configurations for optimal performance}
\label{tab:training_config}
\end{table}

\textbf{Optimizer-Specific Parameters}:

\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|}
\hline
\textbf{Parameter} & \textbf{AdamW} & \textbf{Lion} \\
\hline
$\beta_1$ & 0.9 & 0.9 \\
\hline
$\beta_2$ & 0.999 & 0.99 \\
\hline
$\epsilon$ & 1e-8 & - \\
\hline
Gradient Clipping & 1.0 & 1.0 \\
\hline
\end{tabular}
\caption{Optimizer-specific parameter configurations}
\label{tab:optimizer_config}
\end{table}

\subsubsection{Training Objectives}

\textbf{Loss Function}: Binary Cross-Entropy Loss for relevance prediction:

\begin{equation}
\mathcal{L}_{BCE} = -\frac{1}{N} \sum_{i=1}^{N} [y_i \log(\hat{y}_i) + (1-y_i) \log(1-\hat{y}_i)]
\end{equation}

where $y_i$ represents the true relevance label and $\hat{y}_i$ is the predicted relevance score.

\textbf{Regularization Strategies}:
\begin{itemize}
    \item Weight decay regularization (L2 penalty)
    \item Gradient clipping for training stability
    \item Early stopping based on validation performance
\end{itemize}

\subsection{Infrastructure and Implementation}

\subsubsection{Computational Resources}

\textbf{Hardware Platform}: Modal cloud computing platform \cite{modal_labs}
\begin{itemize}
    \item GPU: NVIDIA A100-80GB
    \item Memory: High-bandwidth memory for large model training
    \item Storage: Fast SSD storage for efficient data loading
    \item Network: High-speed interconnects for distributed training
\end{itemize}

\textbf{Software Stack}:
\begin{itemize}
    \item Framework: PyTorch \cite{paszke2019pytorchimperativestylehighperformance}
    \item Model Library: Sentence Transformers \cite{reimers2019sentence}
    \item Evaluation Tools: TREC Eval \cite{trec_eval_github}, Pyserini \cite{lin2021pyserini}
    \item Experiment Tracking: Weights \& Biases \cite{wandb2020}
\end{itemize}

\subsubsection{Training Pipeline}

\textbf{Model Initialization}:
\begin{enumerate}
    \item Load pre-trained transformer weights
    \item Initialize classification head for relevance prediction
    \item Set up optimizer with appropriate hyperparameters
    \item Configure learning rate scheduler if applicable
\end{enumerate}

\textbf{Training Loop}:
\begin{enumerate}
    \item Forward pass through cross-encoder model
    \item Compute binary cross-entropy loss
    \item Backward propagation with gradient computation
    \item Optimizer step with gradient clipping
    \item Learning rate scheduling update
    \item Validation evaluation at specified intervals
\end{enumerate}

\textbf{Model Checkpointing}:
\begin{itemize}
    \item Save model state at each epoch
    \item Track best model based on validation metrics
    \item Enable training resumption from checkpoints
    \item Model versioning for reproducibility
\end{itemize}

This comprehensive training framework ensures consistent and reproducible comparison between Lion and AdamW optimizers across different model architectures while maintaining state-of-the-art training practices for cross-encoder models in information retrieval.

\section{Implementation Considerations}

\subsection{Computational Resources}
The experimental setup requires significant computational resources due to the intensive nature of cross-encoder training. Modal cloud platform provides the necessary GPU resources with NVIDIA A100 instances, ensuring consistent hardware configuration across all experiments.

Resource allocation considerations include:
\begin{itemize}
    \item GPU memory requirements varying by model size (8GB for MiniLM, 16GB for GTE, 24GB for ModernBERT)
    \item Batch size optimization based on available memory
    \item Distributed training capabilities for larger models
    \item Checkpoint storage and management
\end{itemize}

\subsection{Reproducibility Framework}
Ensuring reproducible results across different runs and environments requires careful attention to:

\textbf{Random Seed Management}:
\begin{itemize}
    \item Fixed seeds for PyTorch, NumPy, and Python random modules
    \item Deterministic CUDA operations where possible
    \item Consistent data shuffling across experiments
\end{itemize}

\textbf{Environment Specification}:
\begin{itemize}
    \item Docker containerization for consistent software environments
    \item Dependency version pinning in requirements files
    \item Hardware specification documentation
    \item Environment variable standardization
\end{itemize}

\subsection{Evaluation Pipeline}
The evaluation pipeline ensures consistent and fair comparison between optimizers:

\textbf{Model Loading and Inference}:
\begin{itemize}
    \item Standardized model loading procedures
    \item Consistent tokenization and preprocessing
    \item Batch size optimization for inference efficiency
    \item GPU memory management during evaluation
\end{itemize}

\textbf{Metric Computation}:
\begin{itemize}
    \item Implementation of standard IR metrics (nDCG, MAP, MRR)
    \item Statistical significance testing procedures
    \item Result aggregation and reporting mechanisms
    \item Cross-validation protocols for robust evaluation
\end{itemize}

\section{Chapter Summary}

This chapter presented a comprehensive methodology for comparing Lion and AdamW optimizers in cross-encoder architectures for information retrieval. The systematic approach encompasses model architecture specifications, optimizer implementations, training frameworks, and evaluation protocols.

Key methodological contributions include:

\begin{itemize}
    \item \textbf{Systematic Architecture Comparison}: Detailed specifications for MiniLM, GTE, and ModernBERT cross-encoder implementations, enabling fair comparison across different model scales and designs.

    \item \textbf{Rigorous Optimizer Implementation}: Mathematical formulations and computational implementations of both Lion and AdamW optimizers, ensuring consistent comparison conditions.

    \item \textbf{Standardized Training Protocol}: Comprehensive training framework with hyperparameter optimization, regularization strategies, and monitoring systems for reproducible results.

    \item \textbf{Robust Evaluation Framework}: Multi-dataset evaluation protocol using TREC 2019 and MS MARCO benchmarks with statistical significance testing.
\end{itemize}

The methodology establishes a foundation for systematic optimizer comparison in neural information retrieval, providing insights into the relationship between optimization algorithms and model architectures. The next chapter presents the comprehensive experimental results obtained through this methodological framework.

### 2025-06-11 cross_encoder/thesis/Chapters/5.results.tex
added thesis (new format)

\chapter{Results and Performance Analysis}
\label{Chapter5}

This chapter presents a comprehensive analysis of our experimental results comparing the Lion and AdamW optimizers for cross-encoder reranking models. We evaluate three distinct transformer architectures (MiniLM, GTE, and ModernBERT) across multiple metrics and examine their training dynamics through detailed visualizations.

\section{Experimental Results}

\subsection{Main Performance Metrics}

Table~\ref{tab:main_results} presents the comprehensive evaluation results for all model configurations on both the TREC 2019 Deep Learning Track and MS MARCO development set. The results are tracked across three training epochs for each model-optimizer combination.

\begin{table}[htbp]
\centering
\caption{Evaluation Results on TREC-DL 2019 and MS-MARCO Dev Passage Ranking}
\label{tab:main_results}
\begin{tabular}{l l c c c c c}
\toprule
\textbf{Model} & \textbf{Optimizer} & \textbf{NDCG@10} & \textbf{MAP} & \textbf{MRR@10} & \textbf{R-Prec} & \textbf{P@10} \\
\midrule
MiniLM & AdamW & 0.7127 & 0.4908 & 0.5826 & 0.4962 & 0.8023 \\
MiniLM & Lion & 0.6808 & 0.4706 & \textbf{0.5988} & 0.4923 & 0.8023 \\
GTE & AdamW & 0.7224 & 0.5005 & 0.5940 & 0.4957 & 0.8140 \\
GTE & Lion & 0.6909 & 0.4921 & 0.5957 & 0.5053 & 0.8140 \\
ModernBERT & AdamW & 0.7105 & 0.5066 & 0.5865 & 0.5161 & 0.8163 \\
ModernBERT & Lion & \textbf{0.7225} & \textbf{0.5115} & 0.5907 & \textbf{0.5183} & \textbf{0.8209} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Training Dynamics Analysis}

To better understand the behavior of each optimizer-model combination, we analyze their training dynamics through various metrics. The following figures present comparative visualizations of training trajectories.

\subsubsection{ModernBERT Training Dynamics}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/modernBERT-Lion-AdamW-eval_loss.png}
    \caption{ModernBERT: Evaluation Loss Comparison between Lion and AdamW}
    \label{fig:modernbert_eval_loss}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/modernBERT-Lion-adamW-train_loss.png}
    \caption{ModernBERT: Training Loss Progression}
    \label{fig:modernbert_train_loss}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/modernBERT-Lion-adamW-lrate.png}
    \caption{ModernBERT: Learning Rate Schedule}
    \label{fig:modernbert_lr}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/modernBERT-Lion-adamW_grad_norm.png}
    \caption{ModernBERT: Gradient Norm Evolution}
    \label{fig:modernbert_grad_norm}
\end{figure}

\subsubsection{GTE Training Dynamics}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/gte_adamW_v_lion_eval_loss.png}
    \caption{GTE: Evaluation Loss Comparison}
    \label{fig:gte_eval_loss}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/gte_adamW_v_lion_train_loss.png}
    \caption{GTE: Training Loss Progression}
    \label{fig:gte_train_loss}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/gte_adamW_v_lion_grad_norm.png}
    \caption{GTE: Gradient Norm Evolution}
    \label{fig:gte_grad_norm}
\end{figure}

\subsubsection{MiniLM Training Dynamics}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/microsoft_MiniLM-L12-H384-uncased_adamW_v_Lion_eval_loss.png}
    \caption{MiniLM: Evaluation Loss Comparison}
    \label{fig:minilm_eval_loss}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/microsoft_MiniLM-L12-H384-uncased_adamW_v_Lion_train_loss.png}
    \caption{MiniLM: Training Loss Progression}
    \label{fig:minilm_train_loss}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/microsoft_MiniLM-L12-H384-uncased_adamW_v_Lion_grad_norm.png}
    \caption{MiniLM: Gradient Norm Evolution}
    \label{fig:minilm_grad_norm}
\end{figure}

\section{Discussion of Results}

\subsection{Optimizer Impact Across Models}

The experimental results reveal distinct patterns in how different models interact with the Lion and AdamW optimizers:

\begin{itemize}
    \item \textbf{ModernBERT Performance:} With Lion optimizer and specialized training configuration (lower learning rate of 2e-6 and Cosine Annealing scheduler), ModernBERT achieved the highest overall performance on TREC DL 2019 metrics (NDCG@10: 0.7225, MAP: 0.5115). The training dynamics (Figures \ref{fig:modernbert_eval_loss}-\ref{fig:modernbert_grad_norm}) show more stable convergence with Lion compared to AdamW.

    \item \textbf{GTE Behavior:} GTE showed stronger performance with AdamW (NDCG@10: 0.7224) using the standard learning rate (2e-5). The training curves (Figures \ref{fig:gte_eval_loss}-\ref{fig:gte_grad_norm}) indicate that AdamW provided more consistent optimization for this model.

    \item \textbf{MiniLM Characteristics:} While MiniLM with AdamW showed better TREC metrics, the Lion optimizer achieved the highest MRR@10 (0.5988) on MS MARCO dev. The training dynamics (Figures \ref{fig:minilm_eval_loss}-\ref{fig:minilm_grad_norm}) suggest that Lion might be particularly effective for certain ranking scenarios.
\end{itemize}

\subsection{Training Dynamics Analysis}

The visualization of training dynamics reveals several key insights:

\begin{itemize}
    \item \textbf{Loss Convergence:} Lion generally shows smoother evaluation loss curves compared to AdamW, particularly evident in the ModernBERT experiments (Figure \ref{fig:modernbert_eval_loss}).

    \item \textbf{Gradient Behavior:} The gradient norm plots (Figures \ref{fig:modernbert_grad_norm}, \ref{fig:gte_grad_norm}, \ref{fig:minilm_grad_norm}) show that Lion maintains more consistent gradient magnitudes throughout training.

    \item \textbf{Learning Rate Impact:} The Cosine Annealing schedule (Figure \ref{fig:modernbert_lr}) proved particularly effective for ModernBERT with Lion, suggesting that adaptive learning rate strategies can significantly influence optimizer performance.
\end{itemize}

\subsection{Model-Specific Considerations}

The results highlight important model-specific characteristics:

\begin{itemize}
    \item \textbf{Context Length Impact:} Models with longer context capabilities (GTE and ModernBERT, supporting 8192 tokens) generally outperformed MiniLM on TREC DL metrics, suggesting the benefit of extended context for reranking.

    \item \textbf{Architecture Influence:} ModernBERT's advanced features (Rotary Positional Embeddings, Flash Attention) appear to synergize well with Lion's optimization approach, particularly with appropriate learning rate scheduling.

    \item \textbf{Model Size Considerations:} Despite being smaller, MiniLM showed competitive performance, especially on MRR@10, indicating that model size alone doesn't determine reranking effectiveness.
\end{itemize}

### 2025-06-11 cross_encoder/thesis/Chapters/6.summary_and_future_scope_of_work.tex
added thesis (new format)

% Chapter Template

\chapter{Summary \& Future Scope of Work}
\label{Chapter6}

This chapter presents a comprehensive summary of our research findings on optimizer effectiveness in cross-encoder reranking and outlines promising directions for future investigation. We reflect on the key insights gained from our experimental analysis and discuss potential avenues for extending this work.

\section{Summary of Research}

Our investigation into the comparative effectiveness of Lion and AdamW optimizers for cross-encoder reranking has yielded several significant findings:

\subsection{Key Findings}

\begin{enumerate}
    \item \textbf{Optimizer-Model Interactions:}
    \begin{itemize}
        \item The effectiveness of optimizers showed strong dependence on model architecture and training configuration
        \item ModernBERT achieved optimal performance with Lion optimizer using specialized learning rate settings
        \item GTE demonstrated superior performance with AdamW under standard training parameters
        \item MiniLM showed varying preferences between optimizers depending on the evaluation metric
    \end{itemize}

    \item \textbf{Performance Achievements:}
    \begin{itemize}
        \item ModernBERT with Lion achieved best-in-class results:
        \begin{itemize}
            \item NDCG@10: 0.7225
            \item MAP: 0.5115
            \item R-Precision: 0.5183
        \end{itemize}
        \item Both MiniLM and ModernBERT with Lion achieved state-of-the-art MRR@10 (0.5988) on MS MARCO dev
    \end{itemize}

    \item \textbf{Training Dynamics:}
    \begin{itemize}
        \item Lion demonstrated more stable evaluation loss curves
        \item Cosine Annealing learning rate schedule proved particularly effective with Lion
        \item Gradient behavior showed distinct patterns between optimizers
    \end{itemize}
\end{enumerate}

\subsection{Technical Insights}

The research revealed several important technical considerations:

\begin{itemize}
    \item \textbf{Learning Rate Sensitivity:} The dramatic impact of learning rate selection on Lion's performance suggests careful tuning is essential
    \item \textbf{Context Length Benefits:} Models supporting longer contexts (8192 tokens) showed generally superior performance
    \item \textbf{Architecture Synergies:} Modern architectural features (RoPE, Flash Attention) appeared to complement Lion's optimization characteristics
\end{itemize}

\section{Future Scope of Work}

Our findings open several promising avenues for future research:

\subsection{Technical Extensions}

\begin{enumerate}
    \item \textbf{Hyperparameter Optimization:}
    \begin{itemize}
        \item Comprehensive grid search for Lion's optimal parameters across different model sizes
        \item Investigation of alternative learning rate schedules
        \item Exploration of momentum parameter impacts
    \end{itemize}

    \item \textbf{Architecture Studies:}
    \begin{itemize}
        \item Evaluation of Lion's effectiveness on emerging transformer variants
        \item Investigation of optimization patterns in multi-query attention mechanisms
        \item Analysis of position embedding schemes' interaction with different optimizers
    \end{itemize}

    \item \textbf{Scaling Studies:}
    \begin{itemize}
        \item Analysis of optimizer behavior with larger model sizes
        \item Investigation of training stability at different batch sizes
        \item Evaluation of memory efficiency at scale
    \end{itemize}
\end{enumerate}

\subsection{Application Extensions}

Several practical applications deserve further investigation:

\begin{itemize}
    \item \textbf{Document-Level Reranking:} Leveraging the 8K context capability for full document reranking
    \item \textbf{Multi-Stage Ranking:} Investigating optimizer impact in cascade ranking architectures
    \item \textbf{Cross-Lingual Applications:} Extending the analysis to multilingual reranking scenarios
    \item \textbf{Domain Adaptation:} Studying optimizer effectiveness in domain transfer settings
\end{itemize}

\subsection{Theoretical Investigations}

Future work should also address theoretical aspects:

\begin{itemize}
    \item Analysis of Lion's convergence properties in ranking optimization
    \item Mathematical characterization of optimizer-architecture interactions
    \item Theoretical bounds on performance with different optimizers
\end{itemize}

\section{Recommendations for Practitioners}

Based on our findings, we recommend:

\begin{itemize}
    \item Consider Lion optimizer for modern architectures with appropriate learning rate scheduling
    \item Maintain AdamW as a robust baseline, especially for established architectures
    \item Carefully tune learning rates when using Lion optimizer
    \item Monitor training dynamics through multiple metrics for optimal checkpoint selection
\end{itemize}

\section{Concluding Remarks}

This research has demonstrated the potential of the Lion optimizer in cross-encoder reranking while highlighting the complexity of optimizer-model interactions. The findings provide a foundation for both practical applications and future research directions in neural information retrieval.

\section*{Acknowledgments}

We gratefully acknowledge Modal Labs (\url{https://modal.com/}) for providing the cloud computing infrastructure and GPU resources (NVIDIA A100-80GB) that made this research possible. Their platform enabled efficient experimentation with large-scale models and datasets, contributing significantly to the comprehensive nature of our analysis.

### 2025-06-11 cross_encoder/thesis/frontmatter.tex
added thesis (new format)

\frontmatter% Use roman page numbering style (i, ii, iii, iv...) for the pre-content pages

\pagestyle{plain} % Default to the plain heading style until the thesis style is called for the body content
%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\begin{titlepage}
\begin{center}
\begin{spacing}{1.2}

{\huge \bfseries\ttitle\par}  % Thesis title

\vspace {5mm}
\textit{A thesis submitted in partial fulfillment of the requirements\\for the award of the degree of}

\vspace{7mm}
\textsc{\huge MASTER OF TECHNOLOGY}

\vspace {3mm}
\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.28]{./IIITA_Logo.png}
\end{figure}
% ************************************
\begin{minipage}[t]{0.5\textwidth}
    \begin{flushleft} \large
        \textit{By:-} \\%[2mm]
            \textsc{SHAHIL KUMAR} % Student Names
    \end{flushleft}
\end{minipage}
\begin{minipage}[t]{0.45\textwidth}
    \begin{flushright} \large
        \textit{Enrollment No.} \\
            \textsc{MML2023008} % Student Enrollment numbers
    \end{flushright}
\end{minipage}\\[1cm]
% *************************************

\textit{Under the Supervision of}\\[2mm]
\textsc{\Large \supname}\\% Supervisor's Name

%\textit{Under the Co-Supervision of}\\[2mm]
%\textsc{\Large Prof. Manish Goswami}\\% Supervisor's Name
\vspace{7mm}
\textit{to the}\\[2mm]
\textsc{\Large \deptname}\\ % Department

\vspace{8mm}
\begin{hindi}
    \textsc{\Large {भारतीय सूचना प्रौद्योगिकी संस्थान, इलाहाबाद}} \\
\end{hindi}
\vspace{3mm}
\textsc{\Large Indian Institute of Information Technology, Allahabad} % university

\vspace{5mm}
{\fontsize{14}{14}\selectfont \text{May , 2025}}

\end{spacing}
\end{center}
\end{titlepage}

%----------------------------------------------------------------------------------------
%	DECLARATION
%----------------------------------------------------------------------------------------
\checktoopen
\begin{figure}[htp]
    % \centering
    \includegraphics[height=3cm,keepaspectratio]{./IIITA_Header.png}
\end{figure}
% \string\setblankpagestyle \space
\thispagestyle{empty}
\vspace{1mm}

\begin{center}
    {\large\bfseries CANDIDATE DECLARATION}
\end{center}

\begin{spacing}{1.5}
\addchaptertocentry{CANDIDATE DECLARATION}
\vspace{10 pt}
I hereby declare that work presented in the report entitled “\textbf{Comparative Analysis of Lion and AdamW Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT}”, submitted towards the fulfillment of MASTER’S THESIS report of M.Tech at Indian Institute of Information Technology Allahabad, is an authenticated original work carried out under supervision of \textbf{Dr. Muneendra Ojha}. Due Acknowledgements have been made in the text to all other material used. the project was done in full compliance with the requirements and constraints of the prescribed curriculum. 
\end{spacing}

\begin{spacing}{2.0}
\begin{flushright}
    \begin{minipage}{0.5\textwidth}
        \flushright \vspace{60 pt}
        \underline{\hspace{6cm}} \\
        \makebox[6cm]{\textbf{Shahil Kumar - MML2023008}} \\[80pt]
    \end{minipage}
\end{flushright}
\end{spacing}
\newpage

%----------------------------------------------------------------------------------------
%	CERTIFICATE
%----------------------------------------------------------------------------------------
\checktoopen
\begin{figure}[htp]
    % \centering
    \includegraphics[height=3cm,keepaspectratio]{./IIITA_Header.png}
\end{figure}
% \string\setblankpagestyle \space
\thispagestyle{empty}
\vspace*{.06\textheight}

\begin{spacing}{1.5}
\addchaptertocentry{CERTIFICATE FROM SUPERVISORS}
\begin{center}
    {\centering\large\bfseries CERTIFICATE FROM SUPERVISORS\par\vspace{10pt}}
\end{center}

\noindent It is certified that the work contained in the thesis titled \enquote{\textbf{Comparative Analysis of Lion and AdamW
Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT}} by \textbf{Shahil kumar} has been carried out under supervision of \textbf{Dr. Muneendra Ojha} and that this work has not been submitted elsewhere for a degree.

\vspace{3.5cm}

\hfill\begin{minipage}{7.5cm}
    \begin{spacing}{1.2}
        \par
        \rule{\textwidth}{0.2pt}\\
        {Dr. Muneendra Ojha} \par
        {\deptname}  \par
        IIIT Allahabad \par
    \end{spacing}
\end{minipage}

\end{spacing}
% \end{certificate}
\cleardoublepage

\begin{figure}[htp]
    % \centering
    \includegraphics[height=3cm,keepaspectratio]{./IIITA_Header.png}
\end{figure}
% \string\setblankpagestyle \space
\thispagestyle{empty}
\vspace*{.06\textheight}

\begin{spacing}{1.5}
\addchaptertocentry{CERIFICATE OF APPROVAL}
\begin{center}
    {\centering\large\bfseries CERTIFICATE OF APPROVAL \par\vspace{10pt}}
\end{center}

 This thesis entitled \textbf{Comparative Analysis of Lion and AdamW Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT}  by \textbf{Shahil Kumar} (MML2023008) is approved for the degree of Master's thesis at IIIT Allahabad  It is understood that by this approval, the undersigned does not necessarily endorse or approve any statement made, opinion expressed, or conclusion drawn therein but approves the thesis only for the purpose for which it is submitted."\\[40 pt]

 Signature and name of the committee members (on final examination and approval of the thesis): \\[10 pt]
 \begin{flushleft}
     \begin{enumerate}
         \item Dr. Muneendra Ojha \\[25pt]
         \item Dr. Kavindra Kandpal\\[25 pt]
         \item Dr. Anand Kumar Tiwari \\[25 pt]

     \end{enumerate}
 \end{flushleft}
 \vspace{20 pt}
 \begin{flushright}
     \begin{minipage}{0.5\textwidth}
        \flushright \vspace{60 pt}
        \underline{\hspace{6cm}} \\
        \makebox[6cm]{\textbf{Dean(A\&R)}} \\[80pt]
    \end{minipage}
 \end{flushright}
\end{spacing}

\newpage
\addchaptertocentry{PLAGIARISM REPORT}
\includepdf[pages=-]{plagrevanth.pdf}
\newpage
\begin{figure}[htp]
    % \centering
    \includegraphics[height=3cm,keepaspectratio]{./IIITA_Header.png}
\end{figure}
% \string\setblankpagestyle \space
\thispagestyle{empty}
\vspace{1mm}

\begin{center}
    {\large\bfseries ACKNOWLEDGEMENT}
\end{center}

\begin{spacing}{1.5}
\addchaptertocentry{ACKNOWLEDGEMENT}
I am thankful to my project supervisor, \textbf{Dr. Muneendra Ojha} for the guidance, support, and invaluable feedback throughout the research process. Their expertise, encouragement, and patience have been instrumental in the completion of this thesis.\\[10pt] 
 I am grateful to Dr. K.P. Singh, Head of the Department of Information Technology, and also my panel members Dr Kavindra Kandpal and Prof. Anand Kumar for their guidance and support.\\[10pt]
 I would also like to thank the Indian Institute of Information Technology, Allahabad for providing the necessary resources and facilities to conduct this research. I am deeply grateful to the participants who generously shared their time and personal information to make this study possible. \\[30 pt]
\end{spacing}

\begin{spacing}{2.0}
\begin{flushright}
    \begin{minipage}{0.5\textwidth}
        \flushright \vspace{60 pt}
        \underline{\hspace{6cm}} \\
        \makebox[6cm]{\textbf{Shahil Kumar - MML2023008}} \\[80pt]
    \end{minipage}
\end{flushright}
\end{spacing}
\newpage

\begin{center}
    {\large\bfseries ABSTRACT}
\end{center}
\begin{spacing}{1.5}
\addchaptertocentry{ABSTRACT} % Add the abstract to the table of contents
     Modern information retrieval systems often employ a two-stage pipeline consisting of an efficient initial retrieval stage followed by a more computationally intensive reranking stage. Cross-encoder models have demonstrated state-of-the-art effectiveness for the reranking task due to their ability to perform deep, contextualized analysis of query-document pairs. The choice of optimizer during the fine-tuning phase can significantly impact the final performance and training efficiency of these models. This paper investigates the impact of using the recently proposed Lion optimizer compared to the widely used AdamW optimizer for fine-tuning cross-encoder rerankers. We fine-tune three distinct transformer models, `microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`, on the MS MARCO passage ranking dataset using both optimizers. Notably, GTE and ModernBERT support longer context lengths (8192 tokens). The effectiveness of the resulting models is evaluated on the TREC 2019 Deep Learning Track passage ranking task and the MS MARCO development set (for MRR@10). Our experiments, facilitated by the Modal cloud computing platform for GPU resource management, show comparative results across three training epochs. ModernBERT trained with Lion achieved the highest NDCG@10 (0.7225) and MAP (0.5121) on TREC DL 2019, while MiniLM trained with Lion tied with ModernBERT with Lion on MRR@10 (0.5988) on MS MARCO dev. We analyze the performance trends based on standard IR metrics, providing insights into the relative effectiveness of Lion versus AdamW for different model architectures and training configurations in the context of passage reranking.
\end{spacing}

\newpage

%----------------------------------------------------------------------------------------
%	LIST OF CONTENTS/FIGURES/TABLES PAGES
%----------------------------------------------------------------------------------------
\addchaptertocentry{Table of Contents}
\tableofcontents % Prints the main table of contents
\listoffigures
\listoftables

%----------------------------------------------------------------------------------------
%	ABBREVIATIONS
%----------------------------------------------------------------------------------------
% \chapter*{List of Abbreviations}
% \addcontentsline{toc}{chapter}{List of Abbreviations}
% \begin{tabular}{@{}ll}
%     Machine Learning Models\\
%     $E_{m}$ & Energy level\\
%     $S_{m}$ & MMSE estimation \\
%     SNR\\
%     Ambient Back scattering Communication \\
%     SVM, RF, DT, XGB \\
%     BER and reflection Coefficient \\
%     Tag decode \\
%     % Add more as needed
% \end{tabular}

%----------------------------------------------------------------------------------------
%	PHYSICAL CONSTANTS/OTHER DEFINITIONS
%----------------------------------------------------------------------------------------

% \begin{constants}{lr@{${}={}$}l} % The list of physical constants is a three column table

% % The \SI{}{} command is provided by the siunitx package, see its documentation for instructions on how to use it

% Speed of Light & $c_{0}$ & \SI{2.99792458e8}{\meter\per\second} (exact)\\
% Constant Name & $Symbol$ & $Constant Value$ with units\\

% \end{constants}

%----------------------------------------------------------------------------------------
%	SYMBOLS
%----------------------------------------------------------------------------------------

% \begin{symbols}{lll} % Include a list of Symbols (a three column table)

% $a$ & distance & \si{\meter} \\
% $P$ & power & \si{\watt} (\si{\joule\per\second}) \\
% %Symbol & Name & Unit \\

% \addlinespace % Gap to separate the Roman symbols from the Greek

% $\omega$ & angular frequency & \si{\radian} \\

% \end{symbols}

%----------------------------------------------------------------------------------------
%	DEDICATION
%----------------------------------------------------------------------------------------

### 2025-06-11 cross_encoder/thesis/main.tex
added thesis (new format)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LaTeX Template
%
% This template was downloaded from:
% http://www.LaTeXTemplates.com
%
% Version 2.x major modifications by:
% Vel (vel@latextemplates.com)
%
% Original Authors:
% Vel (vel@latextemplates.com)
% Johannes Böttcher
% Steve Gunn (http://users.ecs.soton.ac.uk/srg/softwaretools/document/templates/)
% Sunil Patel (http://www.sunilpatel.co.uk/thesis-template/)
%
% This template was modified by Muneendra Ojha to suit the needs of 
% IIIT Allahabad Students. This is a basic template consisting of all 
% requirements as approved by the institute. Please feel free to contact
% if any errors are found or updates needed.
% 
% Do not remove this preamble. Always remember to give credit where it is due.
% 
% Template license:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[
12pt, % The default document font size, options: 10pt, 11pt, 12pt
oneside, % Two side (alternating margins) for binding by default, uncomment to switch to one side
english, % ngerman for German
singlespacing, % options: singlespacing, onehalfspacing or doublespacing
%draft, % Uncomment to enable draft mode (no pictures, no links, overfull hboxes indicated)
% nolistspacing, % If the document is onehalfspacing or doublespacing, uncomment this to set spacing in lists to single
liststotoc, % Uncomment to add the list of figures/tables/etc to the table of contents
% toctotoc, % Uncomment to add the main table of contents to the table of contents
parskip, % Uncomment to add space between paragraphs
%nohyperref, % Uncomment to not load the hyperref package
headsepline, % Uncomment to get a line under the header
% chapterinoneline, % Uncomment to place the chapter title next to the number on one line
% consistentlayout, % Uncomment to change the layout of the declaration, abstract and acknowledgements pages to match the default layout
]{Thesis} % The class file specifying the document structure

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage{mathpazo} % Use the Palatino font by default
\usepackage[backend=biber,
style=numeric,
bibencoding=utf8,
% style=authoryear, % if using style as authoryear, then cite the reference using \parencite{}
% style=reading
]{biblatex}
\DefineBibliographyStrings{english}{
  bibliography = {References},
}
\setlength\bibitemsep{15pt}
\addbibresource{references.bib} % The filename of the bibliography
\usepackage[autostyle=true]{csquotes} % Required to generate language-dependent quotes in the bibliography
\usepackage{array}% http://ctan.org/pkg/array
\usepackage{lipsum}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{pdfpages}
\usepackage{tabularx}
\usepackage{xcolor}  % Optional, for syntax highlighting
\usepackage{float}
\usepackage{hyperref}
\usepackage{etoolbox}
\usepackage{svg}
\usepackage{algorithmic}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{colortbl} % For colored table cells
\usepackage{booktabs} % For better tables
\usepackage{multirow} % For multirow cells in tables
\usepackage{url}      % For URLs
\usepackage{hyperref} % For clickable links (optional)
\usepackage{float}
\usepackage{subcaption}
\usepackage{threeparttable}
\usepackage{tabularx}
\usepackage{caption}
\usepackage[table]{xcolor}

% \usepackage[english]{babel}

%----------------------------------------------------------------------------------------
%	MARGIN SETTINGS
%----------------------------------------------------------------------------------------

\geometry{
	paper=a4paper, % Change to letterpaper for US letter
	inner=20mm, % Inner margin
	outer=20mm, % Outer margin
	bindingoffset=5mm, % Binding offset
	top=30mm, % Top margin
	bottom=20mm, % Bottom margin
	%showframe, % Uncomment to show how the type block is set on the page
}

%----------------------------------------------------------------------------------------
%	THESIS INFORMATION
%----------------------------------------------------------------------------------------

\thesistitle{Comparative Analysis of Lion and AdamW Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT} % Your thesis title, this is used in the title and abstract, print it elsewhere with \ttitle
\supervisor{Dr. Muneendra Ojha} % Your supervisor's name, this is used in the title page, print it elsewhere with \supname
\examiner{} % Your examiner's name, this is not currently used anywhere in the template, print it elsewhere with \examname
\degree{Master of Technology} % Your degree name, this is used in the title page and abstract, print it elsewhere with \degreename
\firstauthor{Shahil Kumar} % Your name, this is used in the title page and abstract, print it elsewhere with \authorname
\enrollmentnumber{MML2023008} % put your university enrollment number here.
\addresses{} % Your address, this is not currently used anywhere in the template, print it elsewhere with \addressname

\subject{Information Technology} % Your subject area, this is not currently used anywhere in the template, print it elsewhere with \subjectname
\keywords{} % Keywords for your thesis, this is not currently used anywhere in the template, print it elsewhere with \keywordnames
\university{Indian Institute of Information Technology, Allahabad} % Your university's name and URL, this is used in the title page and abstract, print it elsewhere with \univname
\department{Department of Information Technology} % Your department's name and URL, this is used in the title page and abstract, print it elsewhere with \deptname
\specialization{Machine Learning and Intelligent Systems} % Your specialization name and URL, this is used in the title page, print it elsewhere with \specializationname
\faculty{Faculty Name} % Your faculty's name and URL, this is used in the title page and abstract, print it elsewhere with \facname

\AtBeginDocument{
\hypersetup{pdftitle=\ttitle} % Set the PDF's title to your title
% \hypersetup{pdfauthor=\authorname} % Set the PDF's author to your name
\hypersetup{pdfkeywords=\keywordnames} % Set the PDF's keywords to your keywords
}
\addto\captionsenglish{
  \renewcommand{\contentsname}{TABLE OF CONTENTS}
  \renewcommand{\listfigurename}{LIST OF FIGURES}
  \renewcommand{\listtablename}{LIST OF TABLES}
}
% Patch TOC title
\patchcmd{\tableofcontents}{\contentsname}{Table Of Contents}{}{}

\begin{document}
\begin{sloppypar}

%----------------------------------------------------------------------------------------
%	FRONT CONTENT
%----------------------------------------------------------------------------------------

\include{frontmatter}

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - CHAPTERS
%----------------------------------------------------------------------------------------

\mainmatter % Begin numeric (1,2,3...) page numbering

\pagestyle{thesis} % Return the page headers back to the "thesis" style
\begin{spacing}{2}

% Include the chapters of the thesis as separate files from the Chapters folder
% Uncomment the lines as you write the chapters

\include{Chapters/1.introduction}
\include{Chapters/2.literature}
\include{Chapters/3.research_gap_and_proposed_solution}
\include{Chapters/4.methods}
\include{Chapters/5.results}
\include{Chapters/6.summary_and_future_scope_of_work} 
\end{spacing}

\addcontentsline{toc}{section}{\textbf{References}}
\nocite{*}
\printbibliography
% %	BIBLIOGRAPHY
% %----------------------------------------------------------------------------------------

% %----------------------------------------------------------------------------------------
\end{sloppypar}
\end{document}

### 2025-06-11 cross_encoder/thesis/research_paper.tex
added thesis (new format)

--- 2025-04-26 cross_encoder/paper/main.tex
@@
 mml2023008@iiita.ac.in}
-\and % Uncomment and repeat for additional authors
-\IEEEauthorblockN{Author 2 Name}
-\IEEEauthorblockA{\textit{Department} \\
-\textit{Institution}\\
-City, Country \\
-email address}
+% \and % Uncomment and repeat for additional authors
+% \IEEEauthorblockN{Author 2 Name}
+% \IEEEauthorblockA{\textit{Department} \\
+% \textit{Institution}\\
+% City, Country \\
+% email address}
 }
@@
 We analyze the performance trends based on standard IR metrics, providing insights into the relative effectiveness of Lion versus AdamW for different model architectures and training configurations in the context of passage reranking.
+The code for training and evaluation is available at \url{https://github.com/skfrost19/Cross-Encoder-Lion-vs-AdamW}, and the trained models are available on Huggingface Model Hub \url{https://huggingface.co/collections/skfrost19/rerenkers-681320776cfb45e44b18f5f1}.
 \end{abstract}

### 2025-06-12 cross_encoder/thesis/Appendices/AppendixA.tex
Refactor results presentation and update appendix in thesis
- Changed table format in Chapter 5 to include epoch-wise results for models.
- Enhanced clarity of evaluation results by adjusting column headers and adding a footnote for metric definitions.
- Updated the table layout to a two-column format for better readability.
- Revised the table of contents to reflect changes in section numbering due to added content.
- Added detailed evaluation metrics in Appendix A, including primary ranking metrics, precision and recall analysis, NDCG analysis, and metric definitions.
- Adjusted references a

@@
-\chapter{Code} % Main appendix title
+\chapter{Detailed Evaluation Metrics} % Main appendix title
 
 \label{AppendixA}
-\begin{sloppypar}
-\lstset{
-  language=Python,
-  backgroundcolor=\color{lightgray},
-  basicstyle=\ttfamily\footnotesize,
-  breaklines=true,
-  commentstyle=\color{green},
-  keywordstyle=\color{blue},
-  stringstyle=\color{red},
-}
 
-\subsection{Image Degradation Code in Python}
+This appendix presents comprehensive evaluation metrics for all model configurations tested in our experiments.
+While the main chapters focus on key performance indicators, these detailed tables provide additional insights into model behavior across various retrieval effectiveness measures.
 
-\section*{Code for Degradation Simulation}
+\section{Primary Ranking Metrics}
+Table~\ref{tab:primary_metrics} presents the main effectiveness metrics including MAP, NDCG@10, MRR@10, and Precision@10 for all tested configurations.
 
-\begin{lstlisting}[language=Python]
-import os
-import cv2
-import numpy as np
-import random
-from typing import Tuple
-from glob import glob
+\begin{table}[htbp]
+\centering
+\caption{Primary ranking metrics comparing cross-encoder models across training epochs.}
+\label{tab:primary_metrics}
+\small
+\begin{tabular}{llccccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{MAP} & \textbf{NDCG@10} & \textbf{MRR@10} & \textbf{P@10} & \textbf{R-Prec} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.5005 & 0.7224 & 0.5940 & 0.8116 & 0.4964 \\
+ & AdamW-2 & 0.4999 & 0.7203 & 0.5942 & 0.8256 & 0.5018 \\
+ & AdamW-3 & 0.4899 & 0.6902 & 0.5972 & 0.8093 & 0.5017 \\
+ & Lion-1 & 0.4794 & 0.6970 & 0.5854 & 0.7884 & 0.4814 \\
+ & Lion-2 & 0.4577 & 0.6792 & 0.5957 & 0.7721 & 0.4642 \\
+ & Lion-3 & 0.4521 & 0.6571 & 0.5931 & 0.7488 & 0.4662 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.4814 & 0.7008 & 0.5828 & 0.8000 & 0.4884 \\
+ & AdamW-2 & 0.4891 & 0.7094 & 0.5818 & 0.8116 & 0.4916 \\
+ & AdamW-3 & 0.4908 & 0.7127 & 0.5826 & 0.8116 & 0.4943 \\
+ & Lion-1 & 0.4858 & 0.7031 & 0.5890 & 0.8140 & 0.4952 \\
+ & Lion-2 & 0.4755 & 0.6916 & 0.5942 & 0.8070 & 0.4803 \\
+ & Lion-3 & 0.4706 & 0.6808 & \textbf{0.5988} & 0.8070 & 0.4809 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.5066 & 0.7105 & 0.5866 & 0.8163 & 0.5161 \\
+ & AdamW-2 & 0.4893 & 0.6839 & 0.5886 & 0.7814 & 0.4946 \\
+ & AdamW-3 & 0.4971 & 0.6959 & 0.5916 & 0.7860 & 0.5116 \\
+ & Lion-1 & \textbf{0.5121} & 0.7142 & 0.5834 & 0.8163 & 0.5148 \\
+ & Lion-2 & 0.5115 & \textbf{0.7225} & 0.5908 & 0.8209 & \textbf{0.5183} \\
+ & Lion-3 & 0.5020 & 0.7051 & \textbf{0.5988} & \textbf{0.8256} & 0.5102 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
-class DegradationModels:
-    def __init__(self, target_size: Tuple[int, int] = (512, 512)):
-        self.target_size = target_size
-        self.ffhq_path = "inputs/originals"
-        self.output_path = "inputs/degraded"
-        os.makedirs(self.output_path, exist_ok=True)
+\section{Precision and Recall Analysis}
+Tables~\ref{tab:precision_metrics} and~\ref{tab:recall_metrics} provide detailed precision and recall values at different cutoff thresholds, revealing the models' behavior at various result list depths.
 
-    def _resize_image(self, image: np.ndarray) -> np.ndarray:
-        return cv2.resize(image, self.target_size)
+\begin{table}[htbp]
+\centering
+\caption{Precision metrics at different cutoff levels.}
+\label{tab:precision_metrics}
+\small
+\begin{tabular}{llcccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{P@5} & \textbf{P@10} & \textbf{P@20} & \textbf{P@100} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.8930 & 0.8116 & 0.7372 & 0.3965 \\
+ & AdamW-2 & 0.8698 & 0.8256 & 0.7326 & 0.3963 \\
+ & AdamW-3 & 0.8558 & 0.8093 & 0.7128 & 0.3942 \\
+ & Lion-1 & 0.8558 & 0.7884 & 0.7070 & 0.3893 \\
+ & Lion-2 & 0.8326 & 0.7721 & 0.6802 & 0.3740 \\
+ & Lion-3 & 0.8140 & 0.7488 & 0.6907 & 0.3740 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.8698 & 0.8000 & 0.7151 & 0.3870 \\
+ & AdamW-2 & 0.8698 & 0.8116 & 0.7186 & 0.3895 \\
+ & AdamW-3 & 0.8558 & 0.8116 & 0.7256 & 0.3891 \\
+ & Lion-1 & 0.8651 & 0.8140 & 0.7233 & 0.3847 \\
+ & Lion-2 & 0.8744 & 0.8070 & 0.7081 & 0.3821 \\
+ & Lion-3 & 0.8465 & 0.8070 & 0.6930 & 0.3809 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.8651 & 0.8163 & 0.7349 & 0.3993 \\
+ & AdamW-2 & 0.8605 & 0.7814 & 0.6953 & 0.3912 \\
+ & AdamW-3 & 0.8465 & 0.7860 & 0.6988 & 0.3958 \\
+ & Lion-1 & 0.8791 & 0.8163 & 0.7314 & 0.4033 \\
+ & Lion-2 & 0.8698 & \textbf{0.8209} & 0.7279 & \textbf{0.4016} \\
+ & Lion-3 & 0.8651 & 0.8256 & 0.7093 & 0.3967 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
-    def _apply_gaussian_blur(self, image: np.ndarray, kernel_size: int = 0, sigma: float = 0) -> np.ndarray:
-        if kernel_size == 0:
-            kernel_size = int(2 * np.ceil(2 * sigma) + 1)
-        if kernel_size % 2 == 0:
-            kernel_size += 1
-        return cv2.GaussianBlur(image, (kernel_size, kernel_size), sigma)
+\begin{table}[htbp]
+\centering
+\caption{Recall metrics at different cutoff levels.}
+\label{tab:recall_metrics}
+\small
+\begin{tabular}{llcccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{Recall@5} & \textbf{Recall@10} & \textbf{Recall@20} & \textbf{Recall@100} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.1051 & 0.1689 & 0.2815 & 0.5501 \\
+ & AdamW-2 & 0.1041 & 0.1760 & 0.2753 & 0.5538 \\
+ & AdamW-3 & 0.1051 & 0.1715 & 0.2692 & 0.5489 \\
+ & Lion-1 & 0.1002 & 0.1614 & 0.2684 & 0.5401 \\
+ & Lion-2 & 0.0973 & 0.1610 & 0.2545 & 0.5211 \\
+ & Lion-3 & 0.0929 & 0.1542 & 0.2574 & 0.5181 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.1027 & 0.1652 & 0.2645 & 0.5426 \\
+ & AdamW-2 & 0.1013 & 0.1668 & 0.2676 & 0.5446 \\
+ & AdamW-3 & 0.1007 & 0.1702 & 0.2719 & 0.5427 \\
+ & Lion-1 & 0.1022 & 0.1688 & 0.2704 & 0.5338 \\
+ & Lion-2 & 0.1025 & 0.1641 & 0.2613 & 0.5338 \\
+ & Lion-3 & 0.1001 & 0.1676 & 0.2542 & 0.5249 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.0995 & 0.1678 & 0.2731 & 0.5520 \\
+ & AdamW-2 & 0.0979 & 0.1634 & 0.2603 & 0.5443 \\
+ & AdamW-3 & 0.1019 & 0.1623 & 0.2627 & 0.5476 \\
+ & Lion-1 & 0.1052 & 0.1689 & 0.2733 & 0.5542 \\
+ & Lion-2 & 0.1048 & \textbf{0.1732} & 0.2699 & \textbf{0.5608} \\
+ & Lion-3 & 0.1037 & 0.1722 & 0.2640 & 0.5570 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
-    def _add_gaussian_noise(self, image: np.ndarray, std: float) -> np.ndarray:
-        noise = np.random.normal(0, std, image.shape).astype(np.float32)
-        noisy_image = image.astype(np.float32) + noise
-        return np.clip(noisy_image, 0, 255).astype(np.uint8)
+\section{NDCG Analysis}
+Table~\ref{tab:ndcg_metrics} provides normalized discounted cumulative gain values at multiple cutoff points.
 
-    def _jpeg_compression(self, image: np.ndarray, quality: int) -> np.ndarray:
-        encode_param = [int(cv2.IMWRITE_JPEG_QUALITY), quality]
-        _, encoded_img = cv2.imencode('.jpg', image, encode_param)
-        return cv2.imdecode(encoded_img, cv2.IMREAD_COLOR)
+\begin{table}[htbp]
+\centering
+\caption{NDCG metrics at different cutoff levels.}
+\label{tab:ndcg_metrics}
+\small
+\begin{tabular}{llcccc}
+\toprule
+\textbf{Model} & \textbf{Config} & \textbf{NDCG@5} & \textbf{NDCG@10} & \textbf{NDCG@20} & \textbf{NDCG@100} \\
+\midrule
+\multirow{6}{*}{GTE} & AdamW-1 & 0.7567 & 0.7224 & 0.7071 & 0.6591 \\
+ & AdamW-2 & 0.7343 & 0.7203 & 0.6967 & 0.6550 \\
+ & AdamW-3 & 0.7080 & 0.6902 & 0.6703 & 0.6426 \\
+ & Lion-1 & 0.7229 & 0.6970 & 0.6786 & 0.6433 \\
+ & Lion-2 & 0.7008 & 0.6792 & 0.6580 & 0.6248 \\
+ & Lion-3 & 0.6801 & 0.6571 & 0.6450 & 0.6124 \\
+\midrule
+\multirow{6}{*}{MiniLM} & AdamW-1 & 0.7207 & 0.7008 & 0.6833 & 0.6402 \\
+ & AdamW-2 & 0.7345 & 0.7094 & 0.6908 & 0.6471 \\
+ & AdamW-3 & 0.7244 & 0.7127 & 0.6949 & 0.6474 \\
+ & Lion-1 & 0.7163 & 0.7031 & 0.6829 & 0.6369 \\
+ & Lion-2 & 0.7099 & 0.6916 & 0.6706 & 0.6295 \\
+ & Lion-3 & 0.6837 & 0.6808 & 0.6584 & 0.6268 \\
+\midrule
+\multirow{6}{*}{ModernBERT} & AdamW-1 & 0.7171 & 0.7105 & 0.6936 & 0.6555 \\
+ & AdamW-2 & 0.7112 & 0.6839 & 0.6701 & 0.6431 \\
+ & AdamW-3 & 0.7156 & 0.6959 & 0.6729 & 0.6497 \\
+ & Lion-1 & 0.7285 & 0.7142 & 0.6991 & 0.6610 \\
+ & Lion-2 & \textbf{0.7381} & \textbf{0.7225} & \textbf{0.6978} & \textbf{0.6638} \\
+ & Lion-3 & 0.7165 & 0.7051 & 0.6797 & 0.6542 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
-    def apply_degradation(self, image: np.ndarray, mode: str = 'BicC', scale_factor: int = 8) -> np.ndarray:
-        image = self._resize_image(image)
-        h, w = image.shape[:2]
+\section{Metric Definitions}
+\label{app:metric_defs}
+Table~\ref{tab:metric_definitions} provides formal definitions of all evaluation metrics used in our experimental analysis.
 
-        if mode == 'BicC':
-            downsampled = cv2.resize(image, (w//scale_factor, h//scale_factor), interpolation=cv2.INTER_CUBIC)
-            upsampled = cv2.resize(downsampled, (w, h), interpolation=cv2.INTER_CUBIC)
-            quality = random.randint(90, 95)
-            return self._jpeg_compression(upsampled, quality)
-
-        elif mode == 'BicN':
-            downsampled = cv2.resize(image, (w//scale_factor, h//scale_factor), interpolation=cv2.INTER_CUBIC)
-            upsampled = cv2.resize(downsampled, (w, h), interpolation=cv2.INTER_CUBIC)
-            noise_std = random.uniform(20, 25)
-            return self._add_gaussian_noise(upsampled, noise_std)
-
-        elif mode == 'BBilN':
-            sigma = random.uniform(5, 7)
-            blurred = self._apply_gaussian_blur(image, sigma=sigma)
-            downsampled = cv2.resize(blurred, (w//scale_factor, h//scale_factor), interpolation=cv2.INTER_LINEAR)
-            upsampled = cv2.resize(downsampled, (w, h), interpolation=cv2.INTER_LINEAR)
-            noise_std = random.uniform(10, 15)
-            return self._add_gaussian_noise(upsampled, noise_std)
-
-        else:
-            raise ValueError(f"Unknown degradation mode: {mode}")
-
-    def process_ffhq_images(self):
-        modes = ['BicC', 'BicN', 'BBilN']
-        image_paths = glob(os.path.join(self.ffhq_path, '*'))
-
-        for mode in modes:
-            output_dir = os.path.join(self.output_path, mode)
-            os.makedirs(output_dir, exist_ok=True)
-
-            print(f"Processing mode: {mode}")
-            for img_path in image_paths:
-                img = cv2.imread(img_path)
-                if img is None:
-                    print(f"Skipping {img_path}, could not read.")
-                    continue
-                degraded = self.apply_degradation(img, mode=mode)
-                filename = os.path.basename(img_path)
-                cv2.imwrite(os.path.join(output_dir, filename), degraded)
-
-if __name__ == "__main__":
-    degrader = DegradationModels()
-    degrader.process_ffhq_images()
-\end{lstlisting}
-
-\subsection*{Code for colour degradation}
-
-\begin{lstlisting}[language=Python]
-from PIL import Image
-import os
-from glob import glob
-
-# Settings
-input_folder = "inputs/originals"
-output_folder = "inputs/degraded_mediancut"
-num_colors = 6  # Number of colors for Median Cut
-
-# Create output directory if it doesn't exist
-os.makedirs(output_folder, exist_ok=True)
-
-# Get all image paths from ffhq folder
-image_paths = glob(os.path.join(input_folder, "*"))
-
-# Process each image
-for path in image_paths:
-    try:
-        # Load image
-        original = Image.open(path).convert("RGB")
-
-        # Apply Median Cut quantization
-        degraded = original.quantize(colors=num_colors, method=0)
-
-        # Convert back to RGB and save
-        degraded_rgb = degraded.convert("RGB")
-        filename = os.path.basename(path)
-        degraded_rgb.save(os.path.join(output_folder, filename))
-
-        print(f"Processed and saved: {filename}")
-    except Exception as e:
-        print(f"Failed to process {path}: {e}")
-\end{lstlisting}
-\begin{lstlisting}[language=Python]
-import cv2
-import numpy as np
-from sklearn.cluster import KMeans
-import os
-from tqdm import tqdm
-
-def degrade_color_palette(image, num_colors=8):
-    """Degrade the color palette of a single image."""
-    h, w, _ = image.shape
-    img_reshaped = image.reshape((-1, 3))
-
-    # Apply k-means to extract dominant colors (palette)
-    kmeans = KMeans(n_clusters=num_colors, random_state=42).fit(img_reshaped)
-    palette = np.uint8(kmeans.cluster_centers_)
-    labels = kmeans.predict(img_reshaped)
-
-    # Reconstruct degraded image from palette
-    degraded = palette[labels].reshape((h, w, 3))
-    return degraded
-
-def process_folder(input_folder, output_folder, num_colors=8):
-    """
-    Process all images in a folder and save degraded versions in an output folder.
-
-    Args:
-        input_folder: Path to the folder containing input images
-        output_folder: Path to save degraded images
-        num_colors: Number of colors to use in the degraded palette
-    """
-    # Create output folder if it doesn't exist
-    os.makedirs(output_folder, exist_ok=True)
-
-    # Get all image files in the input folder
-    image_files = [f for f in os.listdir(input_folder) if f.lower().endswith(('.png', '.jpg', '.jpeg'))]
-
-    # Process each image
-    for filename in tqdm(image_files, desc="Processing images"):
-        # Load image
-        img_path = os.path.join(input_folder, filename)
-        img = cv2.imread(img_path)
-        img = cv2.cvtColor(img, cv2.COLOR_BGR2RGB)
-
-        # Apply color degradation
-        degraded = degrade_color_palette(img, num_colors=num_colors)
-
-        # Save degraded image
-        output_path = os.path.join(output_folder, filename)
-        cv2.imwrite(output_path, cv2.cvtColor(degraded, cv2.COLOR_RGB2BGR))
-
-# Example usage
-input_folder = "inputs/originals"  # Folder containing original images
-output_folder = "inputs/degraded_kmeans"  # Folder to save degraded images
-num_colors = 6  # Number of colors in the degraded palette
-
-process_folder(input_folder, output_folder, num_colors=num_colors)
-print(f"Degraded images saved to {output_folder}")
-
-\end{lstlisting}
-\end{sloppypar}
+\begin{table}[htbp]
+\centering
+\caption{Definitions of key evaluation metrics used in this thesis.}
+\label{tab:metric_definitions}
+\small
+\begin{tabularx}{\textwidth}{lX}
+\toprule
+\textbf{Metric} & \textbf{Definition} \\
+\midrule
+MAP & Mean Average Precision.
+The mean of average precision scores across all queries, where average precision is the average of precision values calculated at every position where a relevant document is retrieved.
+Range: [0,1].
+\\
+NDCG@k & Normalized Discounted Cumulative Gain at rank k. Measures ranking quality with graded relevance and position-based discount. Range: [0,1]. \\
+P@k & Precision at k. The proportion of retrieved documents in the top k results that are relevant. Range: [0,1]. \\
+Recall@k & Recall at k. The proportion of all relevant documents that are retrieved in the top k results. Range: [0,1]. \\
+MRR@10 & Mean Reciprocal Rank at 10. Average of the reciprocal ranks of the first relevant document within the top 10 results. Range: [0,1]. \\
+R-Prec & R-Precision. Precision after R documents retrieved, where R is the number of relevant documents for the query. Range: [0,1]. \\
+bpref & Binary Preference. Measures ranking quality when relevance judgments are incomplete by considering known relevant/non-relevant document pairs. Range: [0,1]. \\
+\bottomrule
+\end{tabularx}
+\end{table}

### 2025-06-12 cross_encoder/thesis/Chapters/5.results.tex
Refactor results presentation and update appendix in thesis
- Changed table format in Chapter 5 to include epoch-wise results for models.
- Enhanced clarity of evaluation results by adjusting column headers and adding a footnote for metric definitions.
- Updated the table layout to a two-column format for better readability.
- Revised the table of contents to reflect changes in section numbering due to added content.
- Added detailed evaluation metrics in Appendix A, including primary ranking metrics, precision and recall analysis, NDCG analysis, and metric definitions.
- Adjusted references a

@@
 
-\begin{table}[htbp]
+\begin{table*}[htbp]
 \centering
@@
 \label{tab:main_results}
-\begin{tabular}{l l c c c c c}
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{l l c c c c c c c}
 \toprule
-\textbf{Model} & \textbf{Optimizer} & \textbf{NDCG@10} & \textbf{MAP} & \textbf{MRR@10} & \textbf{R-Prec} & \textbf{P@10} \\
+\textbf{Base Model} & \textbf{Optimizer} & \textbf{Epoch} & \textbf{NDCG@10} & \textbf{MAP} & \textbf{MRR@10} & \textbf{Recall@10} & \textbf{R-Prec} & \textbf{P@10} \\
 \midrule
-MiniLM & AdamW & 0.7127 & 0.4908 & 0.5826 & 0.4962 & 0.8023 \\
-MiniLM & Lion & 0.6808 & 0.4706 & \textbf{0.5988} & 0.4923 & 0.8023 \\
-GTE & AdamW & 0.7224 & 0.5005 & 0.5940 & 0.4957 & 0.8140 \\
-GTE & Lion & 0.6909 & 0.4921 & 0.5957 & 0.5053 & 0.8140 \\
-ModernBERT & AdamW & 0.7105 & 0.5066 & 0.5865 & 0.5161 & 0.8163 \\
-ModernBERT & Lion & \textbf{0.7225} & \textbf{0.5115} & 0.5907 & \textbf{0.5183} & \textbf{0.8209} \\
+\multirow{3}{*}{MiniLM-L12-H384} & \multirow{3}{*}{AdamW}
+    & 1 & 0.7008 & 0.4814 & 0.5828 & 0.1712 & 0.4899 & 0.8047 \\
+    & & 2 & 0.7094 & 0.4891 & 0.5818 & 0.1715 & 0.5017 & 0.8093 \\
+    & & 3 & 0.7127 & 0.4908 & 0.5826 & 0.1706 & 0.4962 & 0.8023 \\
+\midrule
+\multirow{3}{*}{MiniLM-L12-H384} & \multirow{3}{*}{Lion}
+    & 1 & 0.7031 & 0.4858 & 0.5890 & 0.1698 & 0.4904 & 0.8070 \\
+    & & 2 & 0.6916 & 0.4755 & 0.5942 & 0.1724 & 0.5041 & 0.8116 \\
+    & & 3 & 0.6808 & 0.4706 & \cellcolor{yellow!50}\textbf{0.5988} & 0.1701 & 0.4923 & 0.8023 \\
+\midrule
+\multirow{3}{*}{GTE-multilingual-base} & \multirow{3}{*}{AdamW}
+    & 1 & 0.7224 & 0.5005 & 0.5940 & 0.1733 & 0.4957 & 0.8140 \\
+    & & 2 & 0.7203 & 0.4999 & 0.5942 & \cellcolor{yellow!50}\textbf{0.1733} & 0.5067 & 0.8163 \\
+    & & 3 & 0.6902 & 0.4899 & 0.5972 & 0.1730 & 0.5069 & 0.8140 \\
+\midrule
+\multirow{3}{*}{GTE-multilingual-base} & \multirow{3}{*}{Lion}
+    & 1 & 0.6785 & 0.4754 & 0.5854 & 0.1684 & 0.4849 & 0.7953 \\
+    & & 2 & 0.6909 & 0.4921 & 0.5957 & 0.1721 & 0.5053 & 0.8140 \\
+    & & 3 & 0.6904 & 0.4912 & 0.5931 & 0.1719 & 0.5041 & 0.8093 \\
+\midrule
+\multirow{3}{*}{Modern-BERT-base} & \multirow{3}{*}{AdamW}
+    & 1 & 0.7105 & 0.5066 & 0.5865 & 0.1678 & 0.5161 & 0.8163 \\
+    & & 2 & 0.6839 & 0.4893 & 0.5885 & 0.1634 & 0.4946 & 0.7814 \\
+    & & 3 & 0.6959 & 0.4971 & 0.5916 & 0.1623 & 0.5116 & 0.7860 \\ % Corrected P@10 based on context
+\midrule
+\multirow{3}{*}{Modern-BERT-base} & \multirow{3}{*}{Lion}
+    & 1 & 0.7142 & \cellcolor{yellow!50}\textbf{0.5121} & 0.5834 & 0.1689 & 0.5148 & 0.8163 \\
+    & & 2 & \cellcolor{yellow!50}\textbf{0.7225} & 0.5115 & 0.5907 & 0.1732 & \cellcolor{yellow!50}\textbf{0.5183} & 0.8209 \\
+    & & 3 & 0.7051 & 0.5020 & \cellcolor{yellow!50}\textbf{0.5988}* & 0.1722 & 0.5102 & \cellcolor{yellow!50}\textbf{0.8256} \\
 \bottomrule
-\end{tabular}
-\end{table}
+\end{tabular}%
+}
+\vspace{1em}\\
+\footnotesize{* MRR@10 is calculated on the MS MARCO v1.1 passage dataset development split.
+All other metrics are calculated on TREC DL 2019.
+Best result for each metric highlighted.
+(*Note: MiniLM+Lion also achieved 0.5988 MRR@10 at epoch 3, see Appendix~\ref{AppendixA} for complete metrics).}
+\end{table*}
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[H]
     \centering

### 2025-06-13 cross_encoder/thesis/Chapters/2.literature.tex
Implement code changes to enhance functionality and improve performance

@@
 
-\section{Research Gaps}
+\section{Summary}
 
-While cross-encoder effectiveness is well-established, systematic investigation of optimization strategies remains limited.
-Most studies assume AdamW without exploring alternatives like Lion, particularly for information retrieval tasks where ranking precision is critical.
-
-\subsection{Modern Transformer Architectures}
-
-Recent developments in transformer architectures have introduced models specifically designed for improved efficiency and longer context processing.
-The General Text Embeddings (GTE) family \cite{li2023towards} demonstrated that models trained on diverse, multilingual corpora could achieve strong performance across various text understanding tasks, including retrieval and reranking.
-
-ModernBERT \cite{modernbert} represents the current state-of-the-art in encoder-only transformers, incorporating several architectural innovations:
-
-\begin{itemize}
-    \item \textbf{Rotary Positional Embeddings (RoPE)} \cite{su2023roformerenhancedtransformerrotary}: Enable effective handling of longer sequences by providing more flexible positional encoding.
-    \item \textbf{Flash Attention} \cite{dao2022flashattentionfastmemoryefficientexact}: Improves memory efficiency and computational speed for attention operations.
-    \item \textbf{GeGLU Activation Functions} \cite{shazeer2020gluvariantsimprovetransformer}: Enhance model expressiveness while maintaining computational efficiency.
-\end{itemize}
-
-These architectural improvements enable ModernBERT to process sequences up to 8192 tokens efficiently, making it particularly suitable for long document processing tasks.
+This literature review establishes the foundation for understanding cross-encoder architectures, optimization algorithms, and evaluation methodologies in information retrieval.
+The reviewed work demonstrates the effectiveness of transformer-based cross-encoders and the dominance of AdamW as the standard optimizer.
+However, the systematic exploration of alternative optimization strategies, particularly the recently proposed Lion optimizer, remains an open area for investigation in the context of information retrieval tasks.
 
 %----------------------------------------------------------------------------------------
-\section{Optimization Algorithms for Deep Learning}
-
-The success of neural information retrieval models depends heavily on the optimization algorithms used during training.
-The choice of optimizer affects convergence speed, final performance, and training stability.
-This section reviews the evolution of optimization algorithms and their specific applications to transformer-based models.
-
-\subsection{Classical Optimization Methods}
-
-Stochastic Gradient Descent (SGD) remains a fundamental optimization algorithm, providing theoretical guarantees and interpretable behavior.
-However, the challenges of training deep neural networks - including vanishing gradients, saddle points, and varying curvature across parameter space - motivated the development of adaptive optimization methods.
-
-The introduction of momentum to SGD addressed some limitations by accumulating gradients across iterations, helping optimization navigate through areas of high curvature and accelerate convergence in consistent directions.
-However, SGD with momentum still struggled with the varying scales of different parameters and the need for careful learning rate tuning.
-
-\subsection{Adaptive Learning Rate Methods}
-
-AdaGrad \cite{duchi2011adaptive} introduced the concept of adaptive learning rates, automatically adjusting step sizes based on historical gradient information.
-By maintaining per-parameter learning rates inversely proportional to the square root of accumulated squared gradients, AdaGrad could handle sparse features effectively and reduce the need for manual learning rate tuning.
-
-RMSprop \cite{tieleman2012lecture} addressed AdaGrad's aggressive learning rate decay by using exponential moving averages instead of accumulating all historical gradients.
-This modification prevented the learning rate from becoming too small too quickly, enabling continued learning throughout training.
-
-Adam \cite{kingma2017adam} combined the benefits of momentum-based methods with adaptive learning rates, maintaining separate exponentially decaying averages for both gradients (first moment) and squared gradients (second moment).
-The Adam optimizer became widely adopted due to its robustness across different architectures and tasks, requiring minimal hyperparameter tuning in many scenarios.
-
-\subsection{AdamW and Weight Decay Regularization}
-
-AdamW \cite{loshchilov2019decoupled} represented a significant improvement over Adam by decoupling weight decay regularization from the adaptive learning rate mechanism.
-The key insight was that traditional L2 regularization in Adam led to suboptimal behavior because the adaptive learning rate scaling affected both gradient and regularization terms.
-
-By separating weight decay from gradient-based updates, AdamW achieved better generalization performance and more stable training dynamics. The decoupled formulation:
-
-\begin{equation}
-\theta_{t+1} = \theta_t - \eta \left( \frac{\hat{m}_t}{\sqrt{\hat{v}_t} + \epsilon} + \lambda \theta_t \right)
-\end{equation}
-
-where $\lambda$ represents the weight decay coefficient applied directly to parameters, independent of the adaptive learning rate scaling.
-
-AdamW quickly became the standard optimizer for transformer-based models, demonstrating superior performance across various natural language processing tasks and establishing itself as the default choice for most deep learning applications.
-
-\subsection{Lion Optimizer: A Novel Approach}
-
-The Lion (EvoLved Sign mOmeNtum) optimizer \cite{chen2023symbolic} represents a departure from traditional adaptive optimization approaches.
-Developed through symbolic mathematics and program search, Lion utilizes a simplified update rule that relies primarily on the sign of momentum-based gradients.
-
-The Lion update mechanism follows:
-
-\begin{align}
-c_t &= \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
-\theta_t &= \theta_{t-1} - \eta \left( \text{sign}(c_t) + \lambda \theta_{t-1} \right) \\
-m_t &= \beta_2 m_{t-1} + (1 - \beta_2) g_t
-\end{align}

### 2025-06-13 cross_encoder/thesis/Chapters/3.research_gap_and_proposed_solution.tex
Implement code changes to enhance functionality and improve performance

@@
 
-While AdamW has become standard for transformer training, the Lion optimizer's effectiveness for cross-encoder IR tasks remains unexplored. Key gaps include:
+Despite the established effectiveness of cross-encoder models for information retrieval, several critical gaps exist in current research that motivate this investigation:
+
+\subsection{Limited Optimization Strategy Exploration}
+
+While AdamW has become the de facto standard for transformer training, the systematic investigation of optimization strategies for cross-encoder models remains limited.
+Most studies assume AdamW without exploring alternatives, particularly for information retrieval tasks where ranking precision is critical.
+The recently proposed Lion optimizer \cite{chen2023symbolic}, with its novel sign-based momentum approach, presents an unexplored opportunity for cross-encoder optimization.
+
+\subsection{Architecture-Optimizer Interaction Analysis}
+
+Different transformer architectures may respond differently to optimization strategies, yet this interaction has not been systematically studied. Modern architectures like:
 
 \begin{itemize}
-    \item \textbf{Novel Optimizer Evaluation}: Lion optimizer's memory efficiency and simplified updates haven't been systematically evaluated for cross-encoder training in IR contexts.
-    \item \textbf{Architecture-Optimizer Interaction}: Different transformer architectures (MiniLM, GTE, ModernBERT) may respond differently to optimization strategies.
-    \item \textbf{Training Efficiency Trade-offs}: Lion's claimed efficiency advantages need validation for computationally intensive cross-encoder training.
-    \item \textbf{Comprehensive IR Benchmarking}: Systematic evaluation across standard IR datasets (MS MARCO, TREC DL) with multiple metrics is needed.
+    \item \textbf{ModernBERT} \cite{modernbert}: Incorporating Rotary Positional Embeddings (RoPE), Flash Attention, and GeGLU activations with 8192 token context support
+    \item \textbf{GTE models} \cite{li2023towards}: Multilingual architectures trained on diverse corpora with extended context capabilities
+    \item \textbf{MiniLM} \cite{wang2020minilm}: Efficient distilled variants designed for computational efficiency
 \end{itemize}
+
+Each may exhibit different optimization dynamics with various optimizers, requiring systematic evaluation.
+
+\subsection{Information Retrieval Context Specificity}
+
+Most optimizer comparisons focus on language modeling or general NLP tasks. Information retrieval presents unique challenges:
+\begin{itemize}
+    \item \textbf{Ranking Quality Focus}: IR metrics prioritize ranking precision over classification accuracy
+    \item \textbf{Query-Document Interaction}: Cross-encoder architectures require different optimization considerations than standard sequence models
+    \item \textbf{Training Efficiency}: Large-scale IR datasets demand memory-efficient training approaches
+\end{itemize}
+
+\subsection{Modern Transformer Architecture Gaps}
+
+Recent architectural innovations have not been thoroughly evaluated with alternative optimizers:
+
+\subsubsection{ModernBERT Architectural Advances}
+ModernBERT \cite{modernbert} represents the current state-of-the-art in encoder-only transformers, incorporating several key innovations:
+
+\begin{itemize}
+    \item \textbf{Rotary Positional Embeddings (RoPE)} \cite{su2023roformerenhancedtransformerrotary}: Enable effective handling of longer sequences up to 8192 tokens by providing more flexible positional encoding compared to traditional absolute positional embeddings.
+    \item \textbf{Flash Attention} \cite{dao2022flashattentionfastmemoryefficientexact}: Improves memory efficiency and computational speed for attention operations, particularly beneficial for longer sequences.
+    \item \textbf{GeGLU Activation Functions} \cite{shazeer2020gluvariantsimprovetransformer}: Enhance model expressiveness while maintaining computational efficiency compared to traditional ReLU activations.
+\end{itemize}
+
+These architectural improvements enable efficient processing of long documents but their interaction with different optimizers, particularly Lion's memory-efficient approach, remains unexplored.
+
+\subsubsection{GTE Model Family}
+The General Text Embeddings (GTE) family \cite{li2023towards} demonstrated that models trained on diverse, multilingual corpora could achieve strong performance across various text understanding tasks.
+The GTE-multilingual-base model supports 75 languages and 8192 token context length, but optimization strategy comparisons for this architecture in IR contexts are lacking.
+
+\subsection{Comprehensive IR Benchmarking Gap}
+
+Existing optimizer evaluations typically focus on limited metrics or datasets.
+A comprehensive evaluation across standard IR benchmarks (MS MARCO, TREC DL) with multiple effectiveness metrics (NDCG@10, MAP, MRR@10, Recall@10, R-Prec, P@10) is needed to establish optimizer performance in IR contexts.
+
+\subsection{Training Efficiency and Memory Usage Analysis}
+
+Lion optimizer claims memory efficiency advantages due to its simplified momentum tracking, but these benefits have not been validated for computationally intensive cross-encoder training scenarios typical in information retrieval applications.
 
@@
 
-This research systematically compares Lion and AdamW optimizers across three cross-encoder architectures for information retrieval tasks.
+This research addresses the identified gaps through a systematic comparative study of Lion and AdamW optimizers across three distinct cross-encoder architectures for information retrieval tasks.
 
-\subsection{Model Selection}
-Three transformer architectures representing different efficiency-effectiveness points:
+\subsection{Research Objectives}
+
 \begin{itemize}
-    \item \textbf{MiniLM-L12-H384}: Compact model (384 hidden dims, 12 layers) for efficiency
-    \item \textbf{GTE-multilingual-base}: Multilingual model with 8192 token context support
-    \item \textbf{ModernBERT-base}: State-of-the-art encoder with RoPE, Flash Attention, GeGLU
+    \item \textbf{Primary Objective}: Compare Lion and AdamW optimizer effectiveness for cross-encoder reranking across multiple transformer architectures
+    \item \textbf{Secondary Objectives}: 
+    \begin{itemize}
+        \item Analyze architecture-optimizer interaction patterns
+        \item Evaluate training dynamics and convergence behavior
+        \item Assess memory efficiency and computational requirements
+        \item Provide comprehensive IR benchmark results
+    \end{itemize}
+\end{itemize}
+
+\subsection{Model Selection Rationale}
+
+Three transformer architectures representing different efficiency-effectiveness trade-offs:
+
+\subsubsection{MiniLM-L12-H384-uncased}
+\begin{itemize}
+    \item \textbf{Architecture}: Compact model (384 hidden dimensions, 12 layers)
+    \item \textbf{Design Focus}: Efficiency through knowledge distillation from larger BERT models
+    \item \textbf{Context Length}: Standard 512 tokens
+    \item \textbf{Use Case}: Resource-constrained environments requiring fast inference
+\end{itemize}
+
+\subsubsection{GTE-multilingual-base}
+\begin{itemize}
+    \item \textbf{Architecture}: Multilingual encoder with transformer++ backbone (BERT + RoPE + GLU)
+    \item \textbf{Context Support}: 8192 tokens for long document processing
+    \item \textbf{Language Coverage}: 75+ languages with strong cross-lingual capabilities
+    \item \textbf{Training Data}: Diverse multilingual corpora for robust text understanding
+\end{itemize}
+
+\subsubsection{ModernBERT-base}
+\begin{itemize}
+    \item \textbf{Architecture}: State-of-the-art encoder with advanced optimizations
+    \item \textbf{Key Features}: RoPE, Flash Attention, GeGLU activations
+    \item \textbf{Context Length}: 8192 tokens with efficient long-context processing
+    \item \textbf{Training Scale}: 2 trillion tokens for comprehensive language understanding
 \end{itemize}
@@
 \subsection{Experimental Design}
-\textbf{Training Setup}: Modal cloud platform with NVIDIA A100-80GB GPUs, batch size 16, 3 epochs, MS MARCO passage dataset.
 
-\textbf{Optimizer Configurations}:
+\subsubsection{Training Configuration}
+
+\textbf{Infrastructure}: Modal cloud platform with NVIDIA A100-80GB GPUs providing consistent computational environment
+
+\textbf{Training Parameters}:
 \begin{itemize}
-    \item \textbf{AdamW}: Learning rates 2e-5 (MiniLM, GTE), 2e-6 (ModernBERT), weight decay 0.01
-    \item \textbf{Lion}: Same learning rates, β₁=0.9, β₂=0.99, sign-based momentum updates
+    \item \textbf{Batch Size}: 16 per GPU for memory efficiency
+    \item \textbf{Training Duration}: 3 epochs with checkpoint evaluation after each epoch
+    \item \textbf{Dataset}: MS MARCO passage ranking triplets (~2M query-passage pairs)
+    \item \textbf{Precision}: BF16 for improved training efficiency
+    \item \textbf{Random Seed}: 12 for reproducible results
 \end{itemize}
 
-\textbf{Evaluation}: TREC 2019 DL Track and MS MARCO dev set using NDCG@10, MAP, MRR@10 metrics.
+\subsubsection{Optimizer Configuration Details}
+
+\textbf{AdamW Configuration}:
+\begin{itemize}
+    \item \textbf{Learning Rates}: 
+    \begin{itemize}
+        \item 2e-5 for MiniLM and GTE (standard rate)
+        \item 2e-5 for ModernBERT (baseline comparison)
+    \end{itemize}
+    \item \textbf{Weight Decay}: 0.01 for regularization
+    \item \textbf{Beta Parameters}: β₁=0.9, β₂=0.999 (default values)
+    \item \textbf{Scheduler}: Linear warmup for 10\% of training steps
+\end{itemize}
+
+\textbf{Lion Configuration}:
+\begin{itemize}
+    \item \textbf{Learning Rates}: 
+    \begin{itemize}
+        \item 2e-5 for MiniLM and GTE (consistency with AdamW)
+        \item 2e-6 for ModernBERT (optimized based on architecture characteristics)
+    \end{itemize}
+    \item \textbf{Beta Parameters}: β₁=0.9, β₂=0.99 (recommended values)
+    \item \textbf{Weight Decay}: 0.01 for fair comparison
+    \item \textbf{Scheduler}: Cosine Annealing for ModernBERT, linear for others
+\end{itemize}
+
+\subsection{Evaluation Protocol}
+
+\subsubsection{Datasets and Metrics}
+
+\textbf{Primary Evaluation}: TREC 2019 Deep Learning Track passage ranking
+\begin{itemize}
+    \item \textbf{Query Set}: 43 queries with graded relevance judgments
+    \item \textbf{Candidate Retrieval}: Top-1000 BM25 candidates per query
+    \item \textbf{Metrics}: NDCG@10, MAP, Recall@10, R-Precision, P@10
+\end{itemize}
+
+\textbf{Secondary Evaluation}: MS MARCO passage development set
+\begin{itemize}
+    \item \textbf{Focus Metric}: MRR@10 (Mean Reciprocal Rank at cutoff 10)
+    \item \textbf{Purpose}: Complementary perspective on larger query set
+\end{itemize}
+
+\subsubsection{Evaluation Framework}
+
+\begin{itemize}
+    \item \textbf{Reranking Process}: Cross-encoder scoring of BM25 candidate sets
+    \item \textbf{Evaluation Tools}: Standard trec\_eval for consistent metric calculation
+    \item \textbf{Checkpoint Strategy}: Evaluation after each training epoch
+    \item \textbf{Performance Tracking}: Weights \& Biases for experiment monitoring
+\end{itemize}
+
+\subsection{Controlled Variables and Experimental Validity}
+
+\begin{itemize}
+    \item \textbf{Hardware Consistency}: All experiments on identical A100-80GB GPUs
+    \item \textbf{Software Environment}: Standardized Modal containers with fixed library versions
+    \item \textbf{Data Consistency}: Identical training and evaluation data across all configurations
+    \item \textbf{Random Seed Control}: Fixed seed for reproducible initialization and data shuffling
+    \item \textbf{Hyperparameter Documentation}: Complete parameter logging for replication
+\end{itemize}
 

### 2025-06-13 cross_encoder/thesis/Chapters/6.summary_and_future_scope_of_work.tex
Implement code changes to enhance functionality and improve performance

@@
 
-This chapter presents a comprehensive summary of our research findings on optimizer effectiveness in cross-encoder reranking and outlines promising directions for future investigation.
-We reflect on the key insights gained from our experimental analysis and discuss potential avenues for extending this work.
+This chapter summarizes the key contributions of our research and outlines promising directions for future investigation in optimizer effectiveness for cross-encoder reranking.
 
-\section{Summary of Research}
+\section{Research Summary}
 
-Our investigation into the comparative effectiveness of Lion and AdamW optimizers for cross-encoder reranking has yielded several significant findings:
+This thesis investigated the comparative effectiveness of Lion and AdamW optimizers for cross-encoder reranking across three transformer architectures: MiniLM, GTE, and ModernBERT.
 
-\subsection{Key Findings}
+\subsection{Primary Contributions}
 
 \begin{enumerate}
-    \item \textbf{Optimizer-Model Interactions:}
-    \begin{itemize}
-        \item The effectiveness of optimizers showed strong dependence on model architecture and training configuration
-        \item ModernBERT achieved optimal performance with Lion optimizer using specialized learning rate settings
-        \item GTE demonstrated superior performance with AdamW under standard training parameters
-        \item MiniLM showed varying preferences between optimizers depending on the evaluation metric
-    \end{itemize}
+    \item \textbf{First Systematic Evaluation}: Conducted the first comprehensive comparison of Lion optimizer for cross-encoder information retrieval tasks across multiple architectures.
 
-    \item \textbf{Performance Achievements:}
-    \begin{itemize}
-        \item ModernBERT with Lion achieved best-in-class results:
-        \begin{itemize}
-            \item NDCG@10: 0.7225
-            \item MAP: 0.5115
-            \item R-Precision: 0.5183
-        \end{itemize}
-        \item Both MiniLM and ModernBERT with Lion achieved state-of-the-art MRR@10 (0.5988) on MS MARCO dev
-    \end{itemize}
+    \item \textbf{Architecture-Optimizer Interaction Analysis}: Demonstrated that optimizer effectiveness varies significantly based on model architecture and training configuration.
 
-    \item \textbf{Training Dynamics:}
-    \begin{itemize}
-        \item Lion demonstrated more stable evaluation loss curves
-        \item Cosine Annealing learning rate schedule proved particularly effective with Lion
-        \item Gradient behavior showed distinct patterns between optimizers
-    \end{itemize}
+    \item \textbf{State-of-the-art Results}: Achieved best-in-class performance with ModernBERT + Lion (NDCG@10: 0.7225) and tied best MRR@10 (0.5988) on standard IR benchmarks.
+
+    \item \textbf{Training Dynamics Insights}: Provided detailed analysis of convergence patterns and gradient behavior differences between optimizers.
 \end{enumerate}
 
-\subsection{Technical Insights}
+\subsection{Key Findings Summary}
 
-The research revealed several important technical considerations:
-
+Our experimental results establish that:
 \begin{itemize}
-    \item \textbf{Learning Rate Sensitivity:} The dramatic impact of learning rate selection on Lion's performance suggests careful tuning is essential
-    \item \textbf{Context Length Benefits:} Models supporting longer contexts (8192 tokens) showed generally superior performance
-    \item \textbf{Architecture Synergies:} Modern architectural features (RoPE, Flash Attention) appeared to complement Lion's optimization characteristics
+    \item \textbf{No Universal Optimizer}: Neither Lion nor AdamW consistently outperforms across all architectures
+    \item \textbf{Configuration Sensitivity}: Optimizer effectiveness depends heavily on learning rate and scheduling choices
+    \item \textbf{Modern Architecture Benefits}: Advanced transformer features synergize well with Lion's optimization approach
+    \item \textbf{Metric-Dependent Performance}: Different optimizers excel on different evaluation metrics
 \end{itemize}
@@
 
-Our findings open several promising avenues for future research:
+This research opens several promising avenues for future investigation:
 
-\subsection{Technical Extensions}
+\subsection{Immediate Extensions}
 
 \begin{enumerate}
-    \item \textbf{Hyperparameter Optimization:}
+    \item \textbf{Comprehensive Hyperparameter Studies}:
     \begin{itemize}
-        \item Comprehensive grid search for Lion's optimal parameters across different model sizes
-        \item Investigation of alternative learning rate schedules
-        \item Exploration of momentum parameter impacts
+        \item Systematic grid search for Lion's optimal parameters across model architectures
+        \item Investigation of alternative learning rate schedules beyond Cosine Annealing
+        \item Analysis of momentum parameter sensitivity
     \end{itemize}
 
-    \item \textbf{Architecture Studies:}
+    \item \textbf{Extended Architecture Coverage}:
     \begin{itemize}
-        \item Evaluation of Lion's effectiveness on emerging transformer variants
-        \item Investigation of optimization patterns in multi-query attention mechanisms
-        \item Analysis of position embedding schemes' interaction with different optimizers
+        \item Evaluation on larger model sizes (large and extra-large variants)
+        \item Testing on emerging transformer architectures (e.g., Mamba, RetNet)
+        \item Investigation of decoder-only models for reranking tasks
     \end{itemize}
 
-    \item \textbf{Scaling Studies:}
+    \item \textbf{Enhanced Evaluation Scope}:
     \begin{itemize}
-        \item Analysis of optimizer behavior with larger model sizes
-        \item Investigation of training stability at different batch sizes
-        \item Evaluation of memory efficiency at scale
+        \item Extended evaluation on additional IR benchmarks (BEIR, LoTTE)
+        \item Cross-domain evaluation for generalization assessment
+        \item Multi-language reranking performance analysis
     \end{itemize}
@@
 
-\subsection{Application Extensions}
-
-Several practical applications deserve further investigation:
+\subsection{Advanced Research Directions}
 
 \begin{itemize}
-    \item \textbf{Document-Level Reranking:} Leveraging the 8K context capability for full document reranking
-    \item \textbf{Multi-Stage Ranking:} Investigating optimizer impact in cascade ranking architectures
-    \item \textbf{Cross-Lingual Applications:} Extending the analysis to multilingual reranking scenarios
-    \item \textbf{Domain Adaptation:} Studying optimizer effectiveness in domain transfer settings
+    \item \textbf{Theoretical Analysis}: Mathematical characterization of Lion's convergence properties in ranking optimization contexts
+
+    \item \textbf{Memory and Efficiency Studies}: Detailed analysis of training time, memory usage, and computational efficiency at scale
+
+    \item \textbf{Hybrid Optimization Strategies}: Investigation of combining multiple optimizers during different training phases
+
+    \item \textbf{Auto-ML Integration}: Development of automated optimizer selection based on architecture and dataset characteristics
 \end{itemize}
 
-\subsection{Theoretical Investigations}
+\subsection{Practical Applications}
 
-Future work should also address theoretical aspects:
+Future work should explore:
+\begin{itemize}
+    \item \textbf{Production Deployment}: Large-scale evaluation in real-world search systems
+    \item \textbf{Specialized Domains}: Medical, legal, and scientific document reranking
+    \item \textbf{Multi-Modal Extensions}: Optimization strategies for vision-language retrieval tasks
+\end{itemize}
+
+\section{Practical Recommendations}
+
+For practitioners working with cross-encoder reranking:
 
 \begin{itemize}
-    \item Analysis of Lion's convergence properties in ranking optimization
-    \item Mathematical characterization of optimizer-architecture interactions
-    \item Theoretical bounds on performance with different optimizers
+    \item \textbf{Start with AdamW}: Use AdamW as a robust baseline for initial experiments
+    \item \textbf{Consider Lion for Modern Architectures}: Explore Lion optimizer with appropriate learning rate tuning for recent transformer variants
+    \item \textbf{Monitor Training Dynamics}: Track multiple metrics and loss curves for optimal checkpoint selection
+    \item \textbf{Architecture-Specific Tuning}: Adjust optimizer parameters based on model architecture characteristics
 \end{itemize}
 
-\section{Recommendations for Practitioners}
+\section{Conclusion}
 
-Based on our findings, we recommend:
-
-\begin{itemize}
-    \item Consider Lion optimizer for modern architectures with appropriate learning rate scheduling
-    \item Maintain AdamW as a robust baseline, especially for established architectures
-    \item Carefully tune learning rates when using Lion optimizer
-    \item Monitor training dynamics through multiple metrics for optimal checkpoint selection
-\end{itemize}
-
-\section{Concluding Remarks}
-
-This research has demonstrated the potential of the Lion optimizer in cross-encoder reranking while highlighting the complexity of optimizer-model interactions.
-The findings provide a foundation for both practical applications and future research directions in neural information retrieval.
+This thesis demonstrates that optimizer choice significantly impacts cross-encoder reranking performance, with the effectiveness varying by architecture and configuration.
+The Lion optimizer shows particular promise for modern transformer architectures when properly tuned, while AdamW remains a reliable choice across different settings.
+These findings provide both practical guidance for practitioners and a foundation for future research in neural information retrieval optimization.
 
@@
 
-We gratefully acknowledge Modal Labs (\url{https://modal.com/}) for providing the cloud computing infrastructure and GPU resources (NVIDIA A100-80GB) that made this research possible.
-Their platform enabled efficient experimentation with large-scale models and datasets, contributing significantly to the comprehensive nature of our analysis.
+We gratefully acknowledge Modal Labs for providing the cloud computing infrastructure and GPU resources (NVIDIA A100-80GB) that enabled this comprehensive experimental study.

### 2025-06-16 cross_encoder/thesis/generate_graph.py
Add GPU usage analysis and optimizer comparison for AdamW and Lion
- Introduced JSON files for GPU usage statistics of GTE, miniLM, and modernBERT models with AdamW and Lion optimizers.
- Added corresponding PNG visualizations for GPU usage comparisons.
- Updated main thesis document to include a new section on GPU usage and resource efficiency analysis, detailing the performance of both optimizers across different models.
- Enhanced discussion on optimizer impact, training dynamics, and resource implications, highlighting the efficiency gains achieved with the Lion optimizer.

@@ -1,0 +1,203 @@
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+import numpy as np
+import os
+import json
+import glob
+from pathlib import Path
+
+# Set seaborn style
+sns.set_style("whitegrid")
+sns.set_palette("husl")
+
+def create_output_folder():
+    """Create gpu_usage folder if it doesn't exist"""
+    output_folder = "gpu_usage"
+    if not os.path.exists(output_folder):
+        os.makedirs(output_folder)
+    return output_folder
+
+def extract_gpu_columns(df, optimizer):
+    """Extract GPU memory columns for a specific optimizer"""
+    gpu_cols = []
+    for col in df.columns:
+        # Case-insensitive matching for optimizer names
+        col_lower = col.lower()
+        optimizer_lower = optimizer.lower()
+        if optimizer_lower in col_lower and 'memoryallocated' in col_lower and '__min' not in col_lower and '__max' not in col_lower:
+            gpu_cols.append(col)
+    return gpu_cols
+
+def process_csv_file(csv_file, output_folder):
+    """Process a single CSV file and generate plot + stats"""
+    print(f"Processing {csv_file}...")
+    
+    # Read the CSV file
+    df = pd.read_csv(csv_file)
+    
+    # Get the base filename without extension
+    base_name = Path(csv_file).stem
+    
+    # Clean the data - replace empty strings with NaN and convert to float
+    for col in df.columns:
+        if col != 'Relative Time (Process)':
+            df[col] = pd.to_numeric(df[col], errors='coerce')
+    
+    # Drop rows where time is NaN
+    df = df.dropna(subset=['Relative Time (Process)'])
+    
+    # Identify optimizers in the data
+    optimizers = []
+    for col in df.columns:
+        col_lower = col.lower()
+        if 'adamw' in col_lower:
+            if 'AdamW' not in optimizers:
+                optimizers.append('AdamW')
+        elif 'lion' in col_lower:
+            if 'Lion' not in optimizers:
+                optimizers.append('Lion')
+    
+    if not optimizers:
+        print(f"No optimizer data found in {csv_file}")
+        return
+    
+    # Initialize statistics dictionary
+    stats = {
+        'file_name': base_name,
+        'optimizers': {}
+    }
+    
+    # Create the plot with seaborn styling
+    plt.figure(figsize=(12, 8))
+    
[…]
